---
title:  |
  Probability Proportional \
  to Size Sampling
subtitle: "PUBHBIO 7225 Lecture 14"
# author: "Generative AI acknowledgment: MS Copilot was used to generate alt text for some images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
editor: 
  markdown: 
    wrap: 72
---

## Outline

::: hidden
\newcommand{\thetahat}{\hat{\theta}}

<!-- \newcommand{\that}{\hat{t}} -->

<!-- \newcommand{\tbar}{\bar{t}} -->

<!-- \newcommand{\thatclus}{\hat{t}_{clus}} -->

<!-- \newcommand{\ybar}{\bar{y}} -->

<!-- \newcommand{\ybarclus}{\hat{\bar{y}}_{clus}} -->

<!-- \newcommand{\ybarr}{\hat{\bar{y}}_r} -->

<!-- \newcommand{\yij}{y_{ij}} -->

<!-- \newcommand{\Mbarhat}{\hat{\bar{M}}} -->

<!-- \newcommand{\fpcpar}{\left(1-\frac{n}{N}\right)} -->

<!-- \newcommand{\Cov}{\text{Cov}} -->
:::

::::: columns
::: {.column width="60%"}
**Topics**

-   Sampling with unequal probabilities:\
    Probability Proportional to Size (PPS)

    -   PPS with replacement

    -   PPS without replacement
:::

::: {.column width="40%"}
**Activities**

-   14.1 PPS Sampling
:::
:::::

**Readings**

-   *Probability Proportional to Size (PPS) Sampling*, Encyclopedia of
    Survey Research Methods, 2008 (editor: P Lavrakas) (PDF on Carmen)

**Assignments**

-   Problem Set 4 due Thursday 10/23/2025 11:59pm via Carmen\
    (this is after fall break)

## So Far: SRS at Each Stage

-   **One-stage designs**

    -   Sample $n$ of $N$ PSUs (sample = $\mathcal{S}$)

    -   $\pi_i = \frac{n}{N}$ for PSU $i$     →     $w_i = \frac{N}{n}$

    -   Unbiased estimate of total $t$:
        $\hat{t}= \sum_{i \in \mathcal{S}} \frac{t_i}{\pi_i} = \sum_{i \in \mathcal{S}} w_i t_i = \frac{N}{n} \sum_{i \in \mathcal{S}} t_i$

-   **Two-stage designs**

    -   Stage 1: Sample $n$ of $N$ PSUs (sample = $\mathcal{S}$)

    -   Stage 2: Sample $m_i$ of $M_i$ SSUs for sampled PSU $i$ (sample
        = $\mathcal{S}_i$)

    -   $\pi_{ij} = \frac{n}{N}\frac{m_i}{M_i}$ for SSU $j$ in PSU $i$  
          →     $w_{ij} =  \frac{N}{n}\frac{M_i}{m_i}$

    -   Unbiased estimate of total $t$:
        $\hat{t}=  \sum_{i \in \mathcal{S}} \sum_{j \in \mathcal{S}_i} \frac{y_{ij}}{\pi_{ij}}= \sum_{i \in \mathcal{S}} \sum_{j \in \mathcal{S}_i} w_{ij} y_{ij}= \frac{N}{n} \sum_{i \in \mathcal{S}} M_i \bar{y}_i$

These designs do not take into account the fact that totals $(t_i)$ are
often correlated with the size of PSUs $(M_i)$

We saw that [ratio estimation]{.alert2} can improve estimation of
$\bar{y}_U$ -- but [can we improve on estimation of $t$?]{.alert}

## Time Machine: Back to Lecture 3 (Sampling Clinics)

```{r}
id <- 1:5
docs <- c(2,3,4,5,10)
patients <- c(10,20,50,60,150)
# SRS
a <- c(1,1,1,1,2,2,2,3,3,4)
b <- c(2,3,4,5,3,4,5,4,5,5)
# (patients[a] + patients[b])/2
# Clinic 5 with certainty
a <- c(1,2,3,4)
b <- c(5,5,5,5)
# (4*patients[a] + 1*patients[b])/5
```

-   There are $N=5$ clinics in a town and you want to estimate the
    average number of patients seen in a day (in the town)

-   The population of clinics is:

::: center
|           |                   |                                        |
|:---------:|:-----------------:|:--------------------------------------:|
| Clinic ID | Number of Doctors | Number of Patients Seen in a Day ($y$) |
|     1     |         2         |                   10                   |
|     2     |         3         |                   20                   |
|     3     |         4         |                   50                   |
|     4     |         5         |                   60                   |
|     5     |        10         |                  150                   |
:::

-   True mean number of patients: $\bar{y}_U = 58$

-   The number of patients seen is correlated with the number of doctors
    (makes sense!).

-   We used this **auxiliary data** to create a more efficient sampling
    scheme than an SRS

-   That is the idea behind **PPS** sampling!

## Sampling Clinics (con't)

:::: smaller-text
**Design 1:** SRS of $n=2$ clinics (for each sampled clinic:
$\pi_i = n/N = 2/5$)

|                        |      |      |      |      |      |      |      |      |      |      |
|------:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|
|       Clinics Sampled: | 1, 2 | 1, 3 | 1, 4 | 1, 5 | 2, 3 | 2, 4 | 2, 5 | 3, 4 | 3, 5 | 4, 5 |
| Selection Probability: | 1/10 | 1/10 | 1/10 | 1/10 | 1/10 | 1/10 | 1/10 | 1/10 | 1/10 | 1/10 |
|         Weighted Mean: |  15  |  30  |  35  |  80  |  35  |  40  |  85  |  55  | 100  | 105  |

-   $E(\bar{y})=58$ (unbiased)

-   $V(\bar{y})=921$

**Design 2:** Select largest clinic with **certainty** $(\pi_i = 1)$,
one of the others with equal probability $(\pi_i = 1/4)$

::: center
|                        |      |      |      |      |
|-----------------------:|:----:|:----:|:----:|:----:|
|       Clinics Sampled: | 1, 5 | 2, 5 | 3, 5 | 4, 5 |
| Selection Probability: | 1/4  | 1/4  | 1/4  | 1/4  |
|         Weighted Mean: |  38  |  46  |  70  |  78  |
:::

-   $E(\hat{\theta})=58$ (unbiased)

-   $V(\hat{\theta})=272$ $\rightarrow$ a whole lot smaller than SRS!
    (less than 1/3 of SRS variance!)
::::

## Example: Sampling Schools

-   Consider Activity 13.1, where you sampled students within schools
    using a 2-stage cluster sampling scheme

-   **Population:** $N=30$ schools with a total of
    $M_0 = \sum_{i = 1}^N M_i = 14{,}989$ students

-   Assume it is inexpensive to measure all students in a school, so use
    **1-stage cluster sampling**

::::: columns
::: {.column width="60%"}
-   $N=30$ schools

-   $M_i$ = number of students per school = size variable

-   $y_{ij}=$ indicator for if $j$th student in $i$th school is female

-   $t_i= \sum_{j=1}^{M_i} y_{ij}$ = [total number of females]{.alert}
    in school $i$
:::

::: {.column width="40%"}
|        |                  |                 |
|--------|:----------------:|:---------------:|
|        |    Number of     |    Number of    |
| School | Students $(M_i)$ | Females $(t_i)$ |
| 1      |       643        |       324       |
| 2      |       492        |       249       |
| 3      |       574        |       284       |
| 4      |       324        |       163       |
| ⋮      |        ⋮         |        ⋮        |
:::
:::::

-   Let's consider different ways we can sample $n=2$ schools from this
    population to estimate $t$, the [total number of females in the
    population]{.alert}

## Tangent: Why Sample Only $n=2$? {.pause}

-   Many large-scale surveys have complex designs, with stratification
    and multi-stage clustering

    -   Use lots of strata to increase precision

    -   Often, so highly stratified that each stratum contains a small
        number of PSUs

. . .

-   But, we need to have at least 2 PSUs per stratum to estimate
    variances

    -   Suppose you took an SRS in each stratum (as we have done before)

    -   If you only sample $n_h=1$ you can't estimate variance!

. . .

-   Note that despite this, sometimes only 1 PSU is sampled in highly
    stratified designs

    -   In these cases, the sampled PSUs are divided up into
        **pseudo-PSUs** for the purposes of variance estimation

    -   You use these pseudo-PSUs just like "real" PSUs

    -   You often see this when using publicly available survey data
        ("masked variance PSUs" or "masked variance units")

. . .

**Main point:** Sampling only 2 PSUs is actually pretty common!

## Sampling Design A: SRS

**Design A:** SRS (without replacement) of $n=2$ from the $N=30$ schools

-   I took an SRS and obtained this sample:

|        |                            |                           |
|--------|:--------------------------:|:-------------------------:|
| School | Number of Students $(M_i)$ | Number of Females $(t_i)$ |
| 9      |            473             |            237            |
| 22     |            415             |            204            |

-   To estimate the total number of females I have:

    -   $\pi_i = n/N = 2/30 = 1/15$

    -   $w_i = 1/\pi_i = N/n = 15$

    -   $\hat{t}= \sum_{i \in \mathcal{S}} w_i t_i = (15 \times 237) + (15 \times 204) = 6615$

<p>

-   *Note:* SRS sample is EPSEM, so could alternatively calculate as:
    $\hat{t}= N \bar{t}= 30 \times (237 + 204)/2 = 30 \times 220.5 = 6615$

## SRS (con't)

-   By properties of SRS estimators (think way back!):

|  |  |
|:-----------------------------------|:-----------------------------------|
| True (theoretical) variance: | $\displaystyle V(\hat{t})=N^2 \left(1-\frac{n}{N}\right)\frac{S^2}{n}$ |
| Estimate from a sample: | $\displaystyle \widehat{V}(\hat{t})=N^2 \left(1-\frac{n}{N}\right)\frac{s^2}{n}$ |

-   In my sample,

$s^2 = \frac{1}{n-1} \sum_{i \in \mathcal{S}} (t_i - \bar{t})^2 = \frac{1}{2-1} \left[(237 - 220.5)^2 + (204 - 220.5)^2 \right] = 544.5$

-   Thus my estimated variance is

$\displaystyle \widehat{V}(\hat{t})=N^2 \left(1-\frac{n}{N}\right)\frac{s^2}{n} = 30^2 \left(1 - \frac{2}{30} \right) \frac{544.5}{22} = 228690$

$\displaystyle \widehat{SE}(\hat{t}) = \sqrt{228690} = 478.2$

## Activity 14.1 (Part 1)

PPS Sampling (Part 1)

## Probability Proportional to Size (PPS) Sampling

-   Think about the efficiency gain in the clinics example -- *always
    sampling the much larger clinic resulted in a much smaller variance*
    -- generalizing this idea produces PPS sampling

> [**Probability Proportional to Size (PPS) sampling**]{.alert2} =
> sample selection methods in which the probability of selection for a
> sampling unit is directly proportional to a size measure, $X_i$, which
> is known for all units before sampling

-   We will use throughout this lecture $X_i = M_i =$ number of SSUs in
    the PSU

    -   I.e., $X_i$ = the "size" of the PSU = $M_i$

-   But, other $X_i$ are possible

    -   Example: Sampling businesses using $X_i$ = last year's revenue
        as reported to IRS

-   Note that we want to choose an $X_i$ that is:

    -   Proportional to the $y_i$ that we are trying to estimate

    -   Strictly positive $(>0)$

## Sampling Design B: PPS With Replacement

> [**PPSWR**]{.alert} = Simplest PPS scheme, where inclusion
> probabilities are proportional to $X_i$ and **a PSU can be sampled
> more than once (with replacement)**

-   Conduct $n$ independent draws of a single PSU from the population

-   The probability of selecting PSU $i$ on each draw is proportional to
    $X_i$

-   A PSU **can** be sampled more than once (with replacement)

-   Probability PSU $i$ selected on any one draw = ("psi") =
    $\displaystyle \psi_i  = \frac{X_i}{\sum_{i=1}^N X_i}$

**Note the difference:**

-   $\psi_i$ = probability PSU $i$ is [selected on a single
    draw]{.alert2}

-   $\pi_i$ = probability PSU $i$ is [in the sample]{.alert} =
    $2 \psi_i - \psi_i^2$   *(via addition rule for probabilities)*

-   $\psi_i \ne \pi_i$

## PPSWR (con't)

-   For the population of schools:

| School | $M_i$ | $\psi_i = M_i/(\sum_{i=1}^N M_i)$ |
|:------:|:-----:|:---------------------------------:|
|   1    |  643  |      643/14989 = 0.042898125      |
|   2    |  492  |      492/14989 = 0.032824071      |
|   3    |  574  |      574/14989 = 0.038294749      |
|   4    |  324  |      324/14989 = 0.021615852      |
|   5    |  493  |      493/14989 = 0.032890787      |
|   ⋮    |   ⋮   |                 ⋮                 |
|  Sum:  | 14989 |                 1                 |

<br>

-   The probability of selection (on a single draw), $\psi_i$, is larger
    for the larger schools

## PPSWR Algorithms

::: smaller-text
-   To obtain a PPS With Replacement, many possible sampling algorithms:

    -   Cumulative Size Method, Cumulative $\psi_i$ Method, Lahiri's
        Rejective Method, others...

-   Easiest, conceptually, is the **Cumulative Size Method**:

    1.  Create a list of the PSUs with their sizes, $M_i$

    2.  Calculate the cumulative sums of $M_i$ for each PSU

        -   PSU 1: cumulative sum = $M_1$

        -   PSU 2: cumulative sum = $M_1+M_2$

        -   PSU 3: cumulative sum = $M_1+M_2+M_3$

        -   etc.

    3.  Randomly select $n$ integers between 1 and
        $M_0 = \sum_{i=1}^N M_i$

    4.  Sample the PSUs whose cumulative sum contains these $n$ integers

        -   Select PSU 1 if integer between $1$ and $M_1$

        -   Select PSU 2 if integer between $M_1+1$ and $M_1+M_2$

        -   Select PSU 3 if integer between $M_1+M_2+1$ and
            $M_1+M_2+M_3$

        -   etc.
:::

## Example of Cumulative Size Method

::: smaller-text
-   For the population of schools:

| School | $M_i$ | Cumulative Size   | Cumulative Size (Values)     |        Range |
|:-------------:|:-------------:|---------------|:--------------|--------------:|
|   1    |  643  | $M_1$             | 643                          |     1 to 643 |
|   2    |  492  | $M_1+M_2$         | 643 + 492 = 1153             |  644 to 1153 |
|   3    |  574  | $M_1+M_2+M_3$     | 643 + 492 + 574 = 1709       | 1154 to 1709 |
|   4    |  324  | $M_1+M_2+M_3+M_4$ | 643 + 492 + 574 + 324 = 2033 | 1710 to 2033 |
|   ⋮    |   ⋮   | ⋮                 | ⋮                            |            ⋮ |
|  Sum:  | 14989 |                   |                              |              |

-   Randomly select $n=2$ integers from 1 to 14989 (One easy way in
    Excel: `RANDBETWEEN()` function)

    -   My integers: 3572, 11002

    -   These correspond to:

        -   PSU 8: range 3376 to 3875

        -   PSU 23: range 10935 to 11624

    -   So my sample is PSUs 8 & 23

-   *Note: a PSU could be selected twice!*
:::

## PPSWR: Estimating the Population Total

**Steps**

1.  Estimate the total based on each draw -- i.e., from each sample of
    size 1

    -   Estimated total based on one PSU =
        $\frac{\text{total for that PSU}}{\text{probability of selection on 1 draw}} = \frac{t_i}{\psi_i}$

2.  Average these $n$ estimated totals to produce [unbiased estimator of
    the total =
    $\hat{t}_{wr} = \frac{1}{n}\sum_{i \in \mathcal{S}} \frac{t_i}{\psi_i}$]{.alert}

**Example**

-   My PPSWR sample of $n=2$ schools:

| School | $M_i$ | $\psi_i = M_i/(\sum_{i=1}^N M_i)$ | $t_i$ | $\frac{t_i}{\psi_i}$ |
|:-------------:|:-------------:|:--------------|:-------------:|:-------------:|
| 8 | 500 | $\frac{500}{14989} = 0.0333578$ | 251 | $\frac{251}{0.0333578} = 7524.477$ |
| 23 | 690 | $\frac{690}{14989} = 0.04603376$ | 343 | $\frac{343}{0.04603376} = 7451.0533$ |

-   Estimated total =
    $\hat{t}_{wr} = \frac{1}{n}\sum_{i \in \mathcal{S}} \frac{t_i}{\psi_i} = (7524.5 + 7451.1)/2 = 7487.8$

*Note: If a PSU is selected into the sample more than once, you'll
include its estimate multiple times*

## PPSWR: Sampling Weights

-   Can rewrite the formula for the estimated total to see the sampling
    weights:

$$\hat{t}_{wr} = \frac{1}{n}\sum_{i \in \mathcal{S}} \frac{t_i}{\psi_i} = \sum_{i \in \mathcal{S}} \frac{t_i}{n \psi_i} = \sum_{i \in \mathcal{S}} \textcolor{red}{\frac{1}{n \psi_i}} t_i = \sum_{i \in \mathcal{S}} \textcolor{red}{w_i} t_i$$

-   Weight for PSU $i$:
    $w_i= \frac{1}{n \psi_i} = \frac{1}{\text{expected \# times PSU $i$ appears in sample}}$

-   *Note: sampling weights are **not** the inverse of inclusion
    probabilities for this design*

For my sample,

| School | $M_i$ | $t_i$ | $\psi_i$ | $w_i = \frac{1}{n\psi_i}$ |
|:-------------:|:-------------:|:-------------:|:--------------|:--------------|
| 8 | 500 | 251 | 0.0333578 | $\frac{1}{2*0.0333578} = \frac{1}{0.0667156} = 14.988998$ |
| 23 | 690 | 343 | 0.04603376 | $\frac{1}{2*0.04603376} =\frac{1}{0.09206752} = 10.861594$ |

-   So I can also calculate the estimated total as:
    $\hat{t}_{wr} = \sum_{i \in \mathcal{S}} w_i t_i = (14.988998\times 251) + (10.861594 \times 343) = 7487.8$
    *(Same estimate, of course!)*

## PPSWR: Estimating the Variance of the Population Total {#slide-ppswr-var}

::: smaller-text
> **Estimated variance** of $\hat{t}_{wr}$:
> $$\widehat{V}(\hat{t}_{wr}) = \frac{1}{n} \frac{1}{n-1} \sum_{i \in \mathcal{S}} \left( \frac{t_i}{\psi_i} - \hat{t}_{wr} \right)^2 = \frac{s^2}{n}$$
> where $s^2$ = variance of the $n$ individual estimates of the total
> (the $\frac{t_i}{\psi_i}$)

-   My PPSWR sample of $n=2$ schools:

|  School  | $\psi_i$   | $t_i$ | $\frac{t_i}{\psi_i}$ |
|:--------:|:-----------|:-----:|:--------------------:|
|    8     | 0.0333578  |  251  |       7524.477       |
|    23    | 0.04603376 |  343  |      7451.0533       |
| average: |            |       |        7487.8        |

$\widehat{V}(\hat{t}_{wr}) = \frac{1}{n} \frac{1}{n-1} \sum_{i \in \mathcal{S}} \left( \frac{t_i}{\psi_i} - \hat{t}_{wr} \right)^2
= \frac{1}{2} \frac{1}{(2-1)} \left[(7524.477-7487.8)^2 + (7451.0533-7487.8)^2 \right] = \frac{2695.5}{2} = 1347.6$

$\widehat{SE}(\hat{t}_{wr}) = \sqrt{1347.6} = 36.7$
:::

## Activity 14.1 (Part 2)

PPS Sampling (Part 2)

## Sampling Design C: PPS Without Replacement {.pause}

-   We know that for SRS, variance is reduced when sampling **without**
    replacement (compared to with replacement) -- same holds for PPS!

-   But, now it is trickier to take a sample because the probability of
    selecting PSU $i$ on the second draw **is not the same** as the
    probability of selecting PSU $i$ on the first draw

-   Simple example illustrates this:

    -   $P($select PSU 1 on 1st draw$) =\psi_1$\
        but

    -   $P($select PSU 1 on 2nd draw $|$ **selected** PSU 1 on 1st
        draw$)=0$

    -   $P($select PSU 1 on 2nd draw $|$ **didn't select** PSU 1 on 1st
        draw$)>0$

. . .

-   Many (complex!) algorithms for selecting PPS samples, all require
    that:

    1.  You specify the inclusion probabilities, $\pi_i = n \psi_i$

    2.  Which imply the weights:
        $w_i = \frac{1}{\pi_i} = \frac{1}{n \psi_i}$ (same weights as
        PPSWR!)

. . .

-   Note the implied restriction: $n \psi_i < 1$ since inclusion
    probabilities $(\pi_i)$ cannot be $>$ 1\
    Therefore no one PSU can be so big its $M_i$ takes up more than
    $1/n$ of the total of the $M_i$

## PPSWOR: Estimating the Population Total

> **Horvitz-Thompson Estimator** for the total under PPSWOR:
> $$\hat{t}_{HT} = \sum_{i \in \mathcal{S}} \frac{t_i}{\pi_i} = \sum_{i \in \mathcal{S}} \frac{t_i}{n \psi_i} = \sum_{i \in \mathcal{S}} w_i t_i$$

-   For the population of schools, I used Stata to take a PPSWOR sample
    of $n=2$:

| School | $M_i$ | $t_i$ | $\psi_i$ | $\pi_i = n \psi_i$ | $w_i = \frac{1}{\pi_i}$ |
|:----------:|:----------:|:----------:|:-----------|:-----------|:-----------|
| 14 | 531 | 267 | $\frac{531}{14989} = 0.035426$ | $2*0.035426 = 0.070852$ | $\frac{1}{0.070852} = 14.11$ |
| 24 | 335 | 166 | $\frac{335}{14989} = 0.022350$ | $2*0.022350 = 0.044699$ | $\frac{1}{0.044699} = 22.37$ |

-   My estimate of the total number of females:
    $$\hat{t}_{HT} = \sum_{i \in \mathcal{S}} w_i t_i = 267 \times 14.11 + 166 \times 22.37 = 7482  $$

## PPSWOR: Estimating the Variance of the Population Total

-   The variance of the **PPSWOR** estimator depends on two sets of
    probabilities:

    1.  Selection (Inclusion) probabilities: $\pi_i = P($PSU $i$ in
        sample$)$

    2.  **Pairwise** Selection (Inclusion) probabilities:
        $\pi_{ij}=P($PSU $i$ **and** PSU $j$ in sample$)$

-   The $\pi_{ij}$ are **not** just products of the $\pi_i$!

-   For $n=2$:

    -   $P(i$ first, $j$ second$) = P(i$ first$)\times P(j$ second $| i$
        first$) = \psi_i \times \frac{\psi_j}{1-\psi_i}$\

    -   $P(j$ first, $i$ second$) = P(j$ first$)\times P(i$ second $| j$
        first$) = \psi_j \times \frac{\psi_i}{1-\psi_j}$\

    -   $\pi_{ij} = P(i$ and $j$ in sample$) = P(i$ first, $j$
        second$) + P(j$ first, $i$ second$)$\
        $\pi_{ij} = \psi_i \times \frac{\psi_j}{1-\psi_i}+ \psi_j \times \frac{\psi_i}{1-\psi_j}$

-   Must have the $\pi_{ij}$ to get an unbiased estimate of
    $V(\hat{t}_{HT})$

Complex!

## PPSWOR: Estimating the Variance (con't)

::: smaller-text
-   **Theoretical** variance of $\hat{t}_{HT}$ is:

$$V(\hat{t}_{HT}) = \sum_{i=1}^N \frac{1- \pi_i}{\pi_i} t_i^2 + \sum_{i=1}^N \sum_{k=1, k\ne i}^N \frac{\pi_{ik}-\pi_i \pi_k}{\pi_i \pi_k}t_i t_k = \frac{1}{2} \sum_{i=1}^N \sum_{k=1, k\ne i}^N (\pi_i \pi_k - \pi_{ik}) \left(\frac{t_i}{\pi_i}-\frac{t_k}{\pi_k}\right)^2 $$

> **Horvitz-Thompson** variance estimator (based on 1st expression)
> $$\widehat{V}_{HT}(\hat{t}_{HT}) = \sum_{i \in \mathcal{S}} (1-\pi_i)\frac{t_i^2}{\pi_i^2} + \sum_{i \in \mathcal{S}} \sum_{k \in \mathcal{S}, k \ne i} \frac{\pi_{ik}-\pi_i \pi_k}{\pi_{ik}} \frac{t_i}{\pi_i} \frac{t_k}{\pi_k}$$

> **Sen-Yates-Grundy** variance estimator (based on 2nd expression)
> $$\widehat{V}_{SYG}(\hat{t}_{HT}) = \frac{1}{2}\sum_{i \in \mathcal{S}} \sum_{k \in \mathcal{S}, k \ne i} \frac{\pi_{ik}-\pi_i \pi_k}{\pi_{ik}} \left( \frac{t_i}{\pi_i} - \frac{t_k}{\pi_k} \right)^2$$

-   Though it's the same theoretical variance, $\widehat{V}_{HT}$ and
    $\widehat{V}_{SYG}$ **will not be the same**
:::

## Problems With These Variance Estimators {.pause}

-   Several problems with these variance estimators,
    $\widehat{V}_{HT}(\hat{t}_{HT})$ and
    $\widehat{V}_{SYG}(\hat{t}_{HT})$

    1.  Publicly available datasets *usually don’t have the joint
        inclusion probabilities included*

        -   Not practical! For a sample of size $n$ there are $n(n-1)$
            pairwise inclusion probabilities!

    2.  For some designs the $\pi_{ij}$ may be hard to calculate

    3.  $\widehat{V}_{HT}(\hat{t}_{HT})$ and
        $\widehat{V}_{SYG}(\hat{t}_{HT})$ can actually produce
        **negative** variance estimates for some designs!

. . .

-   For these reasons, we often use the **with replacement variance
    estimate** (the one on [this slide](#slide-ppswr-var))
    $$\widehat{V}_{WR}(\hat{t}_{HT}) = \frac{1}{n} \frac{1}{n-1} \sum_{i \in \mathcal{S}} \left( \frac{t_i}{\psi_i} - \hat{t}_{HT} \right)^2 = \frac{n}{n-1}\sum_{i \in \mathcal{S}} \left( \frac{t_i}{\pi_i} - \frac{\hat{t}_{HT}}{n} \right)^2$$

    -   **Consequence:** result is (usually slightly) conservative --
        i.e., variance too big

    -   **Benefit:** don't need pairwise inclusion probabilities, and
        never get a negative variance!

    -   **Convenience:** $\widehat{V}_{WR}(\hat{t}_{HT})$ is what is
        calculated in most software (including Stata and R)

## Problems With These Variance Estimators (con't)

```{r}
library(tidyverse)
library(haven)
library(survey)
theme_set(theme_bw())

pop <- read_dta("./datasets/smoke_schools.dta")
samp <- pop %>%
  filter(school %in% c(14,24)) %>%
  mutate(wt = 1/(2*Mi/14989))
```

-   For my PPSWOR sample that contains schools 14 and 24:

    -   Using **with-replacement** variance estimation
    -   Notice there is no FPC for PPS sampling

```{r}
#| echo: true
#| classes: .cell-output

des.ppswor <- svydesign(id = ~1, weight = ~wt, data=samp)
svytotal(~totfemale, des.ppswor)
```

-   $\hat{t}_{HT} = 7482$ (same as the hand-calculation)

-   $\widehat{SE}_{WR}(\hat{t}_{HT}) = 54.7$

-   Compare to **without-replacement** variance estimation:

    -   $\widehat{SE}_{HT}(\hat{t}_{HT}) = \text{N/A}$ (variance
        estimate is negative!)

    -   $\widehat{SE}_{SYG}(\hat{t}_{HT}) = 53.1$ (slightly smaller than
        with-replacement estimate)

## Activity 14.1 (Part 3)

PPS Sampling (Part 3)

## Comparing the Designs

-   Three designs compared (for sampling schools): (A) SRS, (B)
    PPSWR, (C) PPSWOR

-   All provide **unbiased estimates** of the true total

-   In terms of theoretical variances, **if size variable is related
    to** $t$, $V_{SRS} > V_{PPSWR} > V_{PPSWOR}$

    -   In a single set of samples you might not see the gain (remember:
        variance is over repeated samples)

-   For the schools data, relationship between size and totals:

::::: columns
::: {.column width="50%"}
```{r}
#| fig-cap: ""
#| fig-alt: "Scatterplot with number of students on x-axis and number of females on y-axis, showing a very linear relationship along an approximately 45 degree line."
pop %>%
  ggplot(aes(x = Mi, y = totfemale)) + 
  geom_point(size = 3) +
  labs(x = "Number of Students (in the school) - Size Variable",
       y = "Number of Females (in the school)") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16))
```
:::

::: {.column width="50%"}
```{r}
#| fig-cap: ""
#| fig-alt: "Scatterplot with number of students on x-axis and number of smokers on y-axis, showing no clear pattern, a random scattering of points."
pop %>%
  ggplot(aes(x = Mi, y = totsmoke)) + 
  geom_point(size = 3) +
  labs(x = "Number of Students (in the school) - Size Variable",
       y = "Number of Smokers (in the school)") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16))
```
:::
:::::

<!-- ![image](./figures/PPS_totfemale_Mi){width="30%"} -->

<!-- ![image](./figures/PPS_totsmoke_Mi){width="30%"} -->

-   Let's look at our estimates for total number of females and total
    number of smokers. What do we see?
