---
title: "Stratified Sampling"
subtitle: "PUBHBIO 7225 Lecture 6"
# author: "Generative AI acknowledgment: MS Copilot was used to generate alt text for images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
---

## Outline

::::: columns
::: {.column width="50%"}
**Topics**

-   Stratified Sampling
-   Design-based Theory for Stratified Samples
-   Sampling Weights
:::

::: {.column width="50%"}
**Activities**

-   6.1 Construct a Stratified Sample Estimator
:::
:::::

<br>

**Assignments**

-   Problem Set 2 due Thursday 9/18/2025 11:59pm via Carmen

## What is Stratified Sampling?

The process:

**STEP 1:** Divide the population into $H$ subpopulations, called [**strata**]{.alert}

-   Strata do not overlap (disjoint/mutually exclusive)
-   Each sampling unit belongs to exactly one stratum (mutually exhaustive)
-   Strata are identified **before** sampling (i.e., available on the frame)

**STEP 2:** Draw an independent probability sample from each stratum (could be an SRS, or other design), then pool the information to obtain overall population estimates

-   Observations [within]{.alert2} strata tend to be [more homogeneous]{.alert2} than observations in the population as a whole

-   [Reduced variance within strata]{.alert2} often leads to a reduced variance for the population-level estimate

-   Thus we want the stratification variable(s) to be related to the $y$ variable(s) of interest -- want to get different means of $y$ across strata

Throughout this lecture we will **assume an SRS is taken in each stratum**

## Why Stratify?

Reasons to stratify include:

-   Protect against (small) chance of getting a really bad sample

    -   Example: sampling $n=200$ OSU undergraduates
    -   If taking an SRS, you might (by chance) get 0 students from outside Ohio
    -   Stratifying by in-state/out-of-state guarantees this doesn't happen

-   Want to ensure specific precision for certain subgroups

    -   Example: want to make sure precision is similar for in-state and out-of-state students, even though number in the population differs

-   May be a cost benefit (easier to administer survey in some strata)

-   Lower variance for overall estimates (e.g, estimates for $\bar{y}_U$)

    -   Variance of $y$ within each stratum is often lower than the variance of $y$ in the whole population
    -   Can lead to cost savings because you can take a smaller sample

<!-- If $y$ has different mean values in different subpopulations, taking a **[stratified sample]{.alert}** can increase precision for estimating population quantities (like $\bar{y}_U$) -->

## Activity 6.1 (Part 1)

Construct a Stratified Sample Estimator (Part 1)

## Constructing an Estimator from a Stratified Sample

-   You took a sample of size $n=15$ from a population of size $N=35$

-   But each of the 15 sampled units **doesn't "represent" the same number of population units**

| **Stratum** | [**Units in Population**]{.alert} | [**Units in Sample**]{.alert2} | **Each Sampled Unit "Represents":** |
|------------------|:----------------:|:----------------:|:----------------:|
| Stratum A | 15 | 5 | 15/5 = 3 |
| Stratum B | 11 | 5 | 11/5 = 2.2 |
| Stratum C | 9 | 5 | 9/5 = 1.8 |

-   This sample **is not EPSEM** -- simple unweighted average of the sampled $y$ will be **biased**

-   Larger strata are "under-represented" in the sample

    -   43% of the [population]{.alert} comes from Stratum A (15/35), but

    -   33% of the [sample]{.alert2} comes from Stratum A (5/15)!

-   We need to "upweight" the units that "represent" more units

## Notation for Stratified Sampling

-   $U$ = the finite population
-   $N$ = size of the population
-   $H$ = number of strata
-   $N_h$ = number of units in stratum $h$ in the population, with $\displaystyle \sum_{h=1}^H N_h = N$
-   $\mathcal{S}_h$ = a particular sample in stratum $h$, with $\displaystyle \bigcup_{h=1}^H \mathcal{S}_h = \mathcal{S}$
-   $n_h$ = size of the sample in stratum $h$, with $\displaystyle \sum_{h=1}^H n_h = n$
-   $y_{hj}$ = characteristic of interest for the $j$th unit in the $h$th stratum

## Example (from Activity)

-   $U$ = population (collection of all the units)
-   $N=35$ (total number of units in the population)
-   $H=3$ strata (A, B, C)

| Stratum | [Population]{.alert} Size | [Sample]{.alert2} Size |
|:--------|:-------------------------:|:----------------------:|
| A       |        $N_1 = 15$         |        $n_1=5$         |
| B       |        $N_2 = 11$         |        $n_2=5$         |
| C       |         $N_3 = 9$         |        $n_3=5$         |
| Total   |    $N=N_1+N_2+N_3=35$     |   $n=n_1+n_2+n_3=15$   |

-   Examples of $y_{hj}$:
    -   value of $y$ for unit $j=2$ in stratum $h=1$ (stratum A) → $y_{12} = 9$
    -   value of $y$ for unit $j=6$ in stratum $h=2$ (stratum B) → $y_{26} = 6$
    -   value of $y$ for unit $j=1$ in stratum $h=3$ (stratum C) → $y_{31} = 11$

## Estimates (Step 1): Stratum-Specific Estimates

| **Quantity** | **Estimand (Truth)** | **Estimate** |
|:-----------------------|:-----------------------|:-----------------------|
| Stratum Mean | $\displaystyle \bar{y}_{hU}=\frac{1}{N_h} \sum_{j=1}^{N_h} y_{hj}$ | $\displaystyle \bar{y}_h = \frac{1}{n_h} \sum_{j\in \mathcal{S}_h} y_{hj}$ |
| Stratum Total | $\displaystyle t_h = \sum_{j=1}^{N_h} y_{hj}= N_h \bar{y}_{hU}$ | $\displaystyle \hat{t}_h = N_h \bar{y}_h = \frac{N_h}{n_h} \sum_{j\in \mathcal{S}_h} y_{hj}$ |
| Stratum Variance (of $y$) | $\displaystyle S_h^2=\frac{1}{N_h-1} \sum_{j=1}^{N_h} (y_{hj}- \bar{y}_{hU})^2$ | $\displaystyle s_h^2=\frac{1}{n_h-1} \sum_{j\in \mathcal{S}_h} (y_{hj}- \bar{y}_h)^2$ |
| Stratum Proportion | $\displaystyle p_{hU}=\bar{y}_{hU}=\frac{1}{N_h} \sum_{j=1}^{N_h} y_{hj}$ | $\displaystyle \hat{p}_h = \frac{1}{n_h} \sum_{j\in \mathcal{S}_h} y_{hj}$ |
| Stratum Variance (of binary $y$) | $\displaystyle S_h^2 = \frac{N_h}{N_h-1} p_{hU}(1-p_{hU})$ | $\displaystyle s_h^2 = \frac{n_h}{n_h-1} \hat{p}_{h}(1-\hat{p}_{h})$ |

-   [These are just the SRS formulae]{.alert}, with extra $h$ subscript (and $j$ instead of $i$ for unit)\]

## Example (from Activity)

::::: columns
::: {.column width="35%"}
Suppose my activity sample was:

![](./figures/activity6_1sample){width="80%" fig-alt="A table with data from three strata (A, B, and C). Rows with yellow highlighting indicate a stratified random sample. Stratum A has a sample of 5 rows, Stratum B has a sample of 5, and Stratum C has a sample of 5."}
:::

::: {.column width="65%"}
We can then calculate/estimate:

[**Stratum A** $(h=1)$:]{.alert} $N_1 = 15, n_1=5$\
Stratum A [**population**]{.alert} mean: $$\begin{aligned}
%\bar{y}_{hU} & =\frac{1}{N_h} \sum_{j=1}^{N_h} y_{hj}\\
\bar{y}_{1U} &=\frac{1}{N_1} \sum_{j=1}^{N_1} y_{1j} = \frac{1}{15} \left(6+9+3+9+\cdots+5\right) = \textcolor{red}{\bf 4.4}
\end{aligned}$$

Stratum A [**sample**]{.alert2} mean: $$\begin{aligned}
%\bar{y}_h &= \frac{1}{n_h} \sum_{j\in \mathcal{S}_h} y_{hj}\\
\bar{y}_1 &= \frac{1}{n_1} \sum_{j\in \mathcal{S}_1} y_{1j} = \frac{1}{5} \left(9+10+1+3+5\right) = \textcolor{blue}{\bf 5.6}
\end{aligned}$$
:::
:::::

My [estimate]{.alert2} of the mean from my sample is [5.6]{.alert2}; close to the [true]{.alert} mean of [4.4]{.alert}

Not new formula -- just the SRS formulae

## Example (from Activity)

<!-- We can also calculate/estimate the variance of $y$ in a stratum: -->

::: {.column width="50%"}
Stratum A [**population**]{.alert} variance of $y$:

$$\begin{aligned}
S_h^2&=\frac{1}{N_h-1} \sum_{j=1}^{N_h} (y_{hj}- \bar{y}_{hU})^2 \\
S_1^2&=\frac{1}{N_1-1} \sum_{j=1}^{N_1} (y_{1j} - \bar{y}_{1U})^2 \\
&= \frac{1}{15-1} \sum_{j=1}^{15} (y_{1j} - 4.4)^2 \\
& = \frac{1}{15-1} \left[ (6-4.4)^2 + \cdots + (5-4.4)^2 \right] \\
&= \textcolor{red}{\bf 10.83}
\end{aligned}$$
:::

::: {.column width="50%"}
Stratum A [**sample**]{.alert2} variance of $y$:

$$\begin{aligned}
s_h^2&=\frac{1}{n_h-1} \sum_{j\in \mathcal{S}_h} (y_{hj}- \bar{y}_h)^2 \\
s_1^2&=\frac{1}{n_1-1} \sum_{j\in \mathcal{S}_h} (y_{1j} - \bar{y}_1)^2 \\
&= \frac{1}{5-1} \sum_{j\in \mathcal{S}_h} (y_{1j} - 5.6)^2 \\
&=  \frac{1}{5-1}\left[ (9-5.6)^2 + \cdots + (5-5.6)^2 \right] \\
&= \textcolor{blue}{\bf 14.8}
\end{aligned}$$
:::

My [estimate]{.alert2} of $V(y)$ from my sample is [14.8]{.alert2}; close-ish to the [true]{.alert} variance of [10.8]{.alert}

Again, not new formula -- just the SRS formulae

## Example (from Activity)

Similar calculations for the other 2 strata yield:

**Means:**

| Stratum            | [Population]{.alert}   | [My Sample]{.alert2} |
|-----------------------|:-----------------------|:---------------------|
| **Stratum A** $(h=1)$ | $\bar{y}_{1U} = 4.4$   | $\bar{y}_1 =  5.6$   |
| **Stratum B** $(h=2)$ | $\bar{y}_{2U} = 8.36$  | $\bar{y}_2 = 7.8$    |
| **Stratum C** $(h=3)$ | $\bar{y}_{3U} = 11.22$ | $\bar{y}_3 = 11.8$   |

**Variances:** *(variance of* $y$, not of the mean of $y$)

| Stratum               | [Population]{.alert} | [My Sample]{.alert2} |
|-----------------------|:---------------------|:---------------------|
| **Stratum A** $(h=1)$ | $S_1^2 = 10.83$      | $s_1^2 = 14.8$       |
| **Stratum B** $(h=2)$ | $S_2^2 = 6.45$       | $s_2^2 = 6.7$        |
| **Stratum C** $(h=3)$ | $S_3^2 = 7.44$       | $s_3^2 = 6.7$        |

Next step will be to combine the stratum estimates to get overall estimate (of the mean)

## Activity 6.1 (Part 2)

Construct a Stratified Sample Estimator (Part 2)

## Estimates (Step 2): Combining Stratum Estimates

Then combine **stratum-specific** estimates to get **overall estimates**:

| **Quantity** | **Estimand (Truth)** | **Estimate** |
|:-----------------------|:-----------------------|:-----------------------|
| Overall Total | $\displaystyle t = \sum_{h=1}^H \sum_{j=1}^{N_h} y_{hj}=  \sum_{h=1}^H t_h = \sum_{h=1}^H N_h \bar{y}_{hU}$ | $\displaystyle \hat{t}_{str}= \sum_{h=1}^H \hat{t}_h= \sum_{h=1}^H N_h \bar{y}_h$ |
| Overall Mean | $\displaystyle \bar{y}_U = \frac{t}{N} = \frac{1}{N} \sum_{h=1}^H N_h \bar{y}_{hU} = \sum_{h=1}^H \frac{N_h}{N} \bar{y}_{hU}$ | $\displaystyle \bar{y}_{str}= \sum_{h=1}^H \frac{N_h}{N} \bar{y}_h$ |
| Overall Proportion | $\displaystyle p = \bar{y}_U = \sum_{h=1}^H \frac{N_h}{N} p_{hU}$ | $\displaystyle \hat{p}_{str}= \sum_{h=1}^H \frac{N_h}{N} \hat{p}_h$ |

**Intuition:**

-   [Total:]{.alert} sum up the stratum totals
-   [Means/Proportions:]{.alert} weighted average of stratum quantities, weighting by the proportion of the population in each stratum

<!-- The [estimates]{.alert} (right-most column) are [**random variables**]{.alert} due to random sampling -->

## Example (from Activity)

|   |   |   | [Population]{.alert} |   | [My Sample]{.alert2} |   |
|-----------|-----------|-----------|:----------|-----------|:----------|-----------|
|  |  |  | Mean | Variance | Mean | Variance |
| Stratum A | $N_1=15$ | $n_1=5$ | $\bar{y}_{1U} = 4.4$ | $S_1^2 = 10.83$ | $\bar{y}_1 = 5.6$ | $s_1^2 = 14.8$ |
| Stratum B | $N_2=11$ | $n_2=5$ | $\bar{y}_{2U} = 8.36$ | $S_2^2 = 6.45$ | $\bar{y}_2 = 7.8$ | $s_2^2 = 6.7$ |
| Stratum C | $N_3=9$ | $n_3=5$ | $\bar{y}_{3U} = 11.22$ | $S_3^2 = 7.44$ | $\bar{y}_3 = 11.8$ | $s_3^2 = 6.7$ |

Overall [**population**]{.alert} mean: $$\bar{y}_U = \sum_{h=1}^H \frac{N_h}{N} \bar{y}_{hU} = \frac{N_1}{N} \bar{y}_{1U} +  \frac{N_2}{N} \bar{y}_{2U} +  \frac{N_3}{N} \bar{y}_{3U} = \frac{15}{35} (4.4) +  \frac{11}{35} (8.36) + \frac{9}{35} (11.22) = \textcolor{red}{\bf 7.4}$$

<!-- *Note: if you simply averaged the 35 population values (*$y_i$), you'd get the same answer (7.4) -->

My sample [**estimate**]{.alert2} of overall mean: $$\begin{aligned}
\bar{y}_{str}= \sum_{h=1}^H \frac{N_h}{N} \bar{y}_h &= \frac{N_1}{N} \bar{y}_1 +  \frac{N_2}{N} \bar{y}_2 +  \frac{N_3}{N} \bar{y}_3 =  \frac{15}{35} (5.6) +  \frac{11}{35} (7.8) + \frac{9}{35} (11.8) = \textcolor{blue}{\bf 7.89} \\
\end{aligned}$$

<!-- *Note: simply averaging the 15 sample values will NOT give you the same answer!* -->

## Expectation and Variance of the Estimated Mean

-   Since we are taking an SRS within each stratum, [stratum-specific estimates are unbiased]{.alert}: $$E(\bar{y}_h) = \bar{y}_{hU} \quad\quad\quad E(\hat{t}_h) = t_{hU} \quad\quad\quad E(\hat{p}_h) = p_{hU}$$

-   This implies that: $$\begin{flalign}
    E(\bar{y}_{str}) &= E\left(\sum_{h=1}^H \frac{N_h}{N} \bar{y}_h \right) = &% \sum_{h=1}^H \frac{N_h}{N} E(\ybar_h) = \sum_{h=1}^H \frac{N_h}{N} \ybar_{hU}  = \ybar_U
    \end{flalign}$$

-   Samples are taken **independently** in each stratum.

-   Remember that if $X$ and $Y$ are **independent**, $V(X+Y) = V(X)+V(Y)$. Thus: $$\begin{flalign}
    V(\bar{y}_{str}) &= V\left(\sum_{h=1}^H \frac{N_h}{N} \bar{y}_h \right) =&% = \sum_{h=1}^H V \left( \frac{N_h}{N} \ybar_h \right) = \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2 V(\ybar_h) = \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2 \fpchpar \frac{S_h^2}{n_h}
    \end{flalign}$$

## Estimators are Random Variables

::: smaller-text
| **Estimate** | **Expected Value** | **Variance** | **Formula** |
|------------------|:-----------------|------------------|------------------|
| Mean: $\bar{y}_{str}$ | $E(\bar{y}_{str})= \bar{y}_U$ | Truth: | $\displaystyle V(\bar{y}_{str})= \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2  \left(1-\frac{n_h}{N_h}\right)\frac{S_h^2}{n_h}$ |
|  |  | Estimate: | $\displaystyle \widehat{V}(\bar{y}_{str})= \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2  \left(1-\frac{n_h}{N_h}\right)\frac{s_h^2}{n_h}$ |
| Total: $\hat{t}_{str}$ | $E(\hat{t}_{str})=t$ | Truth: | $\displaystyle V(\hat{t}_{str})=\sum_{h=1}^H N_h^2 \left(1-\frac{n_h}{N_h}\right)\frac{S_h^2}{n_h}$ |
|  |  | Estimate: | $\displaystyle \widehat{V}(\hat{t}_{str})=\sum_{h=1}^H N_h^2 \left(1-\frac{n_h}{N_h}\right)\frac{s_h^2}{n_h}$ |
| Proportion: $\hat{p}_{str}$ | $E(\hat{p}_{str})=p$ | Truth: | $\displaystyle V(\hat{p}_{str})= \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2 \left(\frac{N_h-n_h}{N_h-1}\right) \frac{p_{hU}(1-p_{hU})}{n_h}$ |
|  |  | Estimate: | $\displaystyle \widehat{V}(\hat{p}_{str}) = \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2 \left(1-\frac{n_h}{N_h}\right)\frac{\hat{p}_h(1-\hat{p}_h)}{n_h-1}$ |
:::
<!-- These expectations and variances come from the $Z_i$; $y_i$ is considered fixed -->

## Example (from Activity)

|   |   |   | [Population]{.alert} |   | [My Sample]{.alert2} |   |
|-----------|-----------|-----------|:----------|-----------|:----------|-----------|
|  |  |  | Mean | Variance | Mean | Variance |
| Stratum A | $N_1=15$ | $n_1=5$ | $\bar{y}_{1U} = 4.4$ | $S_1^2 = 10.83$ | $\bar{y}_1 = 5.6$ | $s_1^2 = 14.8$ |
| Stratum B | $N_2=11$ | $n_2=5$ | $\bar{y}_{2U} = 8.36$ | $S_2^2 = 6.45$ | $\bar{y}_2 = 7.8$ | $s_2^2 = 6.7$ |
| Stratum C | $N_3=9$ | $n_3=5$ | $\bar{y}_{3U} = 11.22$ | $S_3^2 = 7.44$ | $\bar{y}_3 = 11.8$ | $s_3^2 = 6.7$ |

[**True**]{.alert} **sampling variance** of $\bar{y}_{str}$: $$\begin{aligned}
V(\bar{y}_{str}) &= \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2  \left(1-\frac{n_h}{N_h}\right)\frac{S_h^2}{n_h} \\
 & = \left(\frac{15}{35}\right)^2 \left(1 - \frac{5}{15} \right) \frac{10.83}{5} + \left(\frac{11}{35}\right)^2 \left(1 - \frac{5}{11} \right) \frac{6.45}{5} + \left(\frac{9}{35}\right)^2 \left(1 - \frac{5}{9} \right) \frac{7.44}{5} = \textcolor{red}{0.378}
\end{aligned}$$

## Example (from Activity)

|   |   |   | [Population]{.alert} |   | [My Sample]{.alert2} |   |
|-----------|-----------|-----------|:----------|-----------|:----------|-----------|
|  |  |  | Mean | Variance | Mean | Variance |
| Stratum A | $N_1=15$ | $n_1=5$ | $\bar{y}_{1U} = 4.4$ | $S_1^2 = 10.83$ | $\bar{y}_1 = 5.6$ | $s_1^2 = 14.8$ |
| Stratum B | $N_2=11$ | $n_2=5$ | $\bar{y}_{2U} = 8.36$ | $S_2^2 = 6.45$ | $\bar{y}_2 = 7.8$ | $s_2^2 = 6.7$ |
| Stratum C | $N_3=9$ | $n_3=5$ | $\bar{y}_{3U} = 11.22$ | $S_3^2 = 7.44$ | $\bar{y}_3 = 11.8$ | $s_3^2 = 6.7$ |

[**Estimate**]{.alert2} of **sampling variance** of $\bar{y}_{str}$: $$\begin{aligned}
\widehat{V}(\bar{y}_{str}) &= \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2  \left(1-\frac{n_h}{N_h}\right)\frac{s_h^2}{n_h} \\
 & = \left(\frac{15}{35}\right)^2 \left(1 - \frac{5}{15} \right) \frac{14.8}{5} + \left(\frac{11}{35}\right)^2 \left(1 - \frac{5}{11} \right) \frac{6.7}{5} + \left(\frac{9}{35}\right)^2 \left(1 - \frac{5}{9} \right) \frac{6.7}{5} = \textcolor{blue}{0.474}
\end{aligned}$$

## Activity 6.1 (Part 3)

Construct a Stratified Sample Estimator (Part 3)

## Simulated Data from Activity

-   I repeated the activity 10,000 times (using R!), taking two types of samples:

-   

    1.  SRS of size $n=15$

    -   $\bar{y}=$ unweighted average of the 15 sampled units (SRS estimator)

-   

    2.  Stratified sample of $n_h=5$ in each stratum (total of 15 units sampled)

    -   $\bar{y}_{str}=$ weighted average of the stratum means (stratified sample estimator)

-   Resulting estimates of $\bar{y}_U$ (averaged over replicates):

```{r}
#| echo: false

# Simulate taking SRS and stratified samples so as to see the gain
yA <- c(6,9,3,9,10,2,6,1,3,7,3,1,1,0,5)
yB <- c(13,7,12,8,11,6,7,8,8,5,7)
yC <- c(11,12,12,13,8,14,9,15,7)

n_h <- 5
NREPS <- 10000

srs_means <- strat_means <- vector(length=NREPS)
set.seed(525)
for (rep in 1:NREPS)
{
  # SRS
  srs_means[rep] <- mean(sample(c(yA,yB,yC), 3*n_h, replace=FALSE))
  # stratified sample
  yA_s <- sample(yA, n_h, replace=FALSE)
  yB_s <- sample(yB, n_h, replace=FALSE)
  yC_s <- sample(yC, n_h, replace=FALSE)
  strat_means[rep] <- mean(yA_s)*(15/35) + mean(yB_s)*(11/35) + mean(yC_s)*(9/35)
}
```

::::: columns
::: {.column width="50%"}
```{r, out.width='80%'}
#| echo: false
#| fig-cap: ""
#| fig-alt: "Two histograms, one on top of the other, showing the normal-shaped distributions of the mean from an SRS (top) and the mean from a stratified sample (bottom)"

library(tidyverse)

tibble(rep=1:NREPS, srs=srs_means, strat=strat_means) %>%
  pivot_longer(srs:strat, values_to = "ybar", names_to = "method") %>%
  mutate(method_f = factor(method, levels = c("srs","strat"), labels=c("SRS", "Stratified"))) %>%
  ggplot(aes(x=ybar, fill=method_f)) + 
  geom_histogram(bins=45) + 
  facet_grid(method_f ~ .) + 
  theme_bw() + 
  theme(legend.position="none", strip.text.y = element_text(size=18, face="bold")) + 
  labs(x="Mean Estimate", y = "Frequency")
```
:::

::: {.column width="50%"}
| **Method** | **Mean** | **Variance of Mean** |
|-----------------------------|:----------------------:|:------------------:|
| SRS $(\bar{y})$ | `r round(mean(srs_means), 4)` | `r round(var(srs_means), 4)` |
| Stratified $(\bar{y}_{str})$ | `r round(mean(strat_means), 4)` | `r round(var(strat_means), 4)` |
| Truth $(\bar{y}_U)$ | `r round(mean(c(yA, yB, yC)), 4)` |  |

Stratified estimator: **unbiased** and **more precise**!
:::
:::::

<!-- Stratification produced an **unbiased** estimator that is **more precise** than SRS! -->

<!-- Also -- note the average estimated variance of the mean under stratification was [0.3795]{.alert2} -- really close to the true (theoretical) value we calculated of [0.378]{.alert}! -->

## Confidence Intervals for Stratified Samples

-   If the size of the strata ($n_h$) are all large, or the number of strata ($H$) is large, then an approximately $\alpha$-level confidence interval for the true mean $\bar{y}_U$ is given by: $$\left(
    \bar{y}_{str}- z_{\alpha/2} \sqrt{\widehat{V}(\bar{y}_{str})}, \quad \bar{y}_{str}+ z_{\alpha/2} \sqrt{\widehat{V}(\bar{y}_{str})}
    \right)$$ where $z_{\alpha/2}$ is the $(1-\alpha/2)$th percentile of the standard normal distribution

-   In practice, often use ${t}$ distribution with [DF = $n-H$]{.alert2} instead of $z_{\alpha/2}$.

    -   $\textcolor{blue}{DF = n-H}$ should remind you of ANOVA

        -   **One-way ANOVA:** [denominator DF = $N-k$]{.alert2} = \# observations $-$ \# groups

        -   **Stratified Sampling:** [DF = $n-H$]{style="color: purple"} = \# observations $-$ \# strata

    -   We lose a DF for each group/stratum mean we estimate

    -   You may see this DF "rule" written as" **DF = \# of PSUs -- \# of Strata**

        -   PSU = Primary Sampling Unit (which so far has been the observation unit)

## Selection Probabilities and Sampling Weights

When an SRS is taken in each stratum:

[**Selection (Inclusion) Probabilities**]{.alert}

-   Stratum $h$ has $N_h$ units

-   Take an SRS of $n_h$ units

-   Thus, $P($unit $j$ in stratum $h$ selected$) = \textcolor{red}{\pi_{hj} = \frac{\text{Sample size in the stratum}}{\text{Population size in the stratum}} = \frac{n_h}{N_h}}$

[**Sampling Weights**]{.alert2}

-   Sample weight is the inverse of the selection (inclusion) probability

-   Thus, sample weight for unit $j$ in stratum $h$ = $\textcolor{blue}{w_{hj} = \frac{1}{\pi_{hj}}=\frac{N_h}{n_h}}$

Note also that [the sum of the weights for sampled units = population size, $N$]{.alert}: $$\sum_{i \in \mathcal{S}} w_i = \sum_{h=1}^H \sum_{j\in \mathcal{S}_h} w_{hj} = \sum_{h=1}^H \sum_{j\in \mathcal{S}_h} \frac{N_h}{n_h} = \sum_{h=1}^H n_h \frac{N_h}{n_h} = \sum_{h=1}^H N_h = N$$

## Horvitz-Thompson Estimators

-   We can re-write the estimate of the (overall) total using the weights: $$\begin{aligned}
    \hat{t}_{str}&= \sum_{h=1}^H N_h \textcolor{myGreen}{\bar{y}_h} = \sum_{h=1}^H N_h \textcolor{myGreen}{\frac{1}{n_h} \sum_{j\in \mathcal{S}_h} y_{hj}} = \sum_{h=1}^H \sum_{j\in \mathcal{S}_h} \textcolor{red}{\frac{N_h}{n_h}} y_{hj}= \sum_{h=1}^H \sum_{j\in \mathcal{S}_h} \textcolor{red}{w_{hj}} y_{hj}\\
    \text{(re-index) } &= \sum_{i \in \mathcal{S}} \textcolor{red}{w_i} y_i
    \end{aligned}$$

-   This is the [Horvitz-Thompson estimator (of the total)!]{.alert}

-   Similarly, the estimate of the (overall) mean written using the weights: $$\bar{y}_{str}= \sum_{h=1}^H \frac{N_h}{N} \bar{y}_h = \frac{\hat{t}_{str}}{N} = \frac{\sum_{i \in \mathcal{S}} w_i y_i}{\sum_{i \in \mathcal{S}} w_i}$$

-   This is the [Horvitz-Thompson estimator of the mean]{.alert}

::: smaller-text
*(not something new, just showing that the stratified sampling formulas to estimate the mean and total are the Horvitz-Thompson estimators)*
:::

## Activity 6.1 (Part 4)

Construct a Stratified Sample Estimator (Part 4)

## Can a Stratified Sample be EPSEM?

> (Reminder) **EPSEM** = inclusion probability (and thus sample weights) same for all population units

-   Consider this scenario:

|           | $N_h$ | $n_h$ |
|-----------|:------|:------|
| Stratum A | 1,000 | 10    |
| Stratum B | 4,000 | ?     |
| Stratum C | 500   | ?     |

-   Can you come up with sample sizes for stratum B and stratum C that is EPSEM?

## An EPSEM Stratified Sample

-   Stratified sample is EPSEM [**if sampling fraction is the same in all strata**]{.alert2}

    -   I.e., if $\frac{n_h}{N_h}$ is the same for all $h$ (for all strata)

-   In an EPSEM stratified sample, sample weights are identical but this is [**not** an SRS!]{.alert}

-   Variance calculations [must take into account the design (stratification)]{.alert}

-   What if the population and your sample look like this:

```{r, out.width='60%'}
#| echo: false
#| fig-cap: ""
#| fig-alt: "Two histograms, one on top of the other, showing tri-modal distributions from the full population (top) and a single stratified sample (bottom)"

set.seed(1977)
pop <- tibble(y = c(rnorm(1000), 5+rnorm(4000), 10+rnorm(500)),
             stratum = c(rep(1,1000), rep(2,4000), rep(3,500))) %>%
  add_column(set = "Population")
samp <- pop %>%
  group_by(stratum) %>%
  slice_sample(prop = 0.01) %>%
  ungroup() %>%
  mutate(set = "Sample")
bind_rows(pop, samp) %>%
  ggplot(aes(x = y, fill = factor(stratum))) +
  geom_histogram(bins=35, alpha = 0.5) + 
  theme_bw() + 
  labs(x="y", y = "Frequency") + 
  facet_grid(set ~ ., scales="free_y") + 
  theme(legend.position="none", strip.text.y = element_text(size=18, face="bold"))
```

-   Your variance estimator for the sample mean would be way too large if you treated this like an SRS!
