---
title:  |
  Linear Regression \ 
  with Survey Data
subtitle: "PUBHBIO 7225 Lecture 20"
author: "Generative AI acknowledgment: Google Gemini was used to assist in generating alt text for most images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
editor: 
  markdown: 
    wrap: 72
---

## Outline

<!-- ::: hidden -->

<!-- \newcommand{\ybar}{\bar{y}} -->

<!-- ::: -->

::::: columns
::: {.column width="60%"}
**Topics**

-   Linear regression with survey data (from a probability sample)
:::

::: {.column width="40%"}
**Activities**

-   20.1 Linear Regression with NHANES
:::
:::::

**Readings**

-   See last slide for optional readings (PDFs on Carmen)

**Assignments**

-   Problem Set 5 (last one!) due Thursday 11/13/2025 11:59pm via
    Carmen\
    (this will be graded by Dr Andridge, not peer reviewed)
-   Quiz 5 (last one!) due Thursday 11/20/2025 11:59pm via Carmen

## Regression With Survey Data {.pause}

-   Until now, we've estimated **totals**, **means**, and
    **proportions**, for a population or subpopulation

-   For these types of estimates, it is (hopefully by now) obvious why
    we [need to use weights]{.alert2} to account for the survey design

. . .

-   For analyses that evaluate **relationships** among variables (e.g.,
    regression) the issue is less straightforward

. . .

-   **Broad generality:** weighting "matters" more for estimating
    [marginal]{.alert} quantities (means/proportions/totals) than it
    does for estimating [relationships]{.alert2} between/among variables
    (like via regression)

-   **Thought experiment:**

    -   I want to learn about student loan balances among OSU students

    -   I sample students via a design that *oversamples* students from
        out of state (who pay higher tuition)

    -   If I ignore the design and estimate the following quantities,
        under what situations might they be biased?

        -   Mean loan balance, population of *all* OSU students

        -   Mean loan balance, population of OSU students *from outside
            Ohio* (out-of-state)

        -   *Difference* in mean loan balance between in-state and
            out-of-state students

## Does Weighting Matter for Regression? {.pause}

-   [**Unweighted**]{.alert} regression estimator is unbiased for true
    linear regression parameters if:

    -   Mean model is [**correctly specified**]{.alert}, (correctly
        specified $X$s), and

    -   Sampling is [**ignorable**]{.alert} -- meaning that
        $Z_i \perp\!\!\!\!\perp y | x$

. . .

-   [**Weighted**]{.alert2} regression estimator protects us if one of
    these conditions is not met

    -   If model is [**not correctly specified**]{.alert2}, weighted
        estimates give us an unbiased estimate of the mis-specified
        relationship (whereas unweighted estimates would be biased)

        -   Ex: True relationship quadratic, but you model as linear --
            weights give you the best estimate of that incorrectly
            specified relationship (best linear approximation to the
            quadratic relationship)

    -   If sampling is [**non-ignorable**]{.alert2}, unweighted
        estimates will generally be biased, but weighted estimates will
        be unbiased

        -   Need the weights to properly account for sampling design
            since $Z$ is related to $y$

. . .

-   [Should we always use weighted regression?]{.alert3}

    -   Penalty for doing weighted regression = loss of efficiency
        (larger variances) when model is correctly specified and
        sampling is ignorable

## Example: 1988 NMIHS

-   Classic example: Gestational age (GA) and birthweight (BW) in the
    1988 National Maternal and Infant Health Survey (NMIHS)

    -   NMIHS = Probability sample of babies whose mothers were aged 15
        and older in the U.S.

    -   Oversampled African-American and low-birthweight babies

    -   Sample size roughly 10,000 babies

-   Comparing weighted and unweighted analysis using BW (g) to predict
    GA (weeks):

| Parameter       | Weighted Estimate (SE) | Unweighted Estimate (SE) |
|:----------------|:----------------------:|:------------------------:|
| Intercept       |      32.08 (.17)       |       26.36 (.14)        |
| Birthweight (g) |    .00215 (.00005)     |     .00383 (.00005)      |

-   **Weighted**: 100 gram reduction in BW associated with a 0.215 week
    (1.5 day) decrease in GA

-   **Unweighted**: 100 gram reduction in BW associated with a 0.383
    week (2.7 day) decrease in GA

    -   Quite different results! Why?

## Example (con't)

-   Difference is due to the **non-linearity** in the
    birthweight-gestational age association and the **oversampling of
    low birthweight babies**

:::::: columns
::: {.column width="25%"}
**Unweighted**

![](./figures/NMIHS_unweighted){fig-alt="A scatter plot showing the relationship between mean birthweight in grams (on the x-axis) and mean gestational age in weeks (on the y-axis). There are 19 data points plotted as circles, and they form an increasing quadratic relationship. A straight line is drawn through the data, showing a positive linear correlation between the two variables. This line indicates that as mean birthweight increases, so does mean gestational age. The data points appear to follow this trend but are not perfectly aligned on the line since the true relationship is quadratic."
fig-align="center"}
:::

::: {.column width="25%"}
**Weighted**

![](./figures/NMIHS_weighted){fig-alt="A scatter plot also showing the relationship between weighted mean birthweight in grams (on the x-axis) and weighted mean gestational age in weeks (on the y-axis). The plot contains 19 data points, represented as circles of varying sizes. The size of each circle is proportional to its weight, with larger circles appearing at higher birthweights and gestational ages. The points form an increasing quadratic relationship. A straight line is drawn through the data, showing a positive linear correlation. Compared to the unweighted plot, this graph's data points are less scattered and more tightly clustered around the trend line, especially at the higher end of the scale, suggesting that weighting has produced a better approximation of the linear trend."
fig-align="center"}
:::

::: {.column width="50%"}
-   True relationship appears to be *quadratic* -- but model
    misspecified as *linear*

-   Weighted model estimates give the best estimate of the linear
    approximation

-   Alternative: include a quadratic term in the model (thus weighted
    $\approx$ unweighted)
:::
::::::

## Conditioning on Design Variables

-   One approach: include **all design variables as predictors**, along
    with any variables involved in creating the weights, and use
    **unweighted regression**

    -   This by definition makes the sampling [**ignorable**]{.alert3}

        -   $Z \perp\!\!\!\!\perp y|x$ if $x$ determines $Z$!

    -   This would potentially mean including in the model:

        -   Stratum indicators

        -   Cluster indicators (via random effects, perhaps)

        -   Variables involved in nonresponse adjustments,
            poststratification/raking, etc.

    -   A daunting task, especially for surveys with complex designs and
        post-survey weight adjustments

-   If it can be done, regression coefficient estimates and SEs will be
    **unbiased**

## Conditioning on Design Variables (con't)

-   Sometimes (often), however, **true** design variables are **not
    provided to end users**, or weights are otherwise manipulated

    -   "Pseudo-strata" instead of the true strata

    -   "Masked variance units (MVUs)" = "Pseudo-PSUs" instead of of the
        true PSUs

-   Additional manipulations to weights also common

    -   Weight trimming methods (truncating very large weights) can't
        easily be incorporated

-   So this approach isn't actually possible most of the time with
    publicly available data

## Design Variable as the Outcome

-   Sometimes we might be interested in predictors of a variable that
    was used in the sampling design

-   Example: Birthweight (BW) and smoking in the 1988 NMIHS

    -   Comparing weighted and unweighted analysis using maternal
        smoking status to predict BW (g) using ANCOVA:

        | Mother Smoked? | Weighted Mean (SE) | Unweighted Mean (SE) |
        |:---------------|:------------------:|:--------------------:|
        | No             |     3409 (8.0)     |      2923 (12)       |
        | Yes            |     3232 (12)      |      2664 (18)       |
        | Difference     |      177 (15)      |       260 (21)       |

    -   Difference between weighted and unweighted is due to the
        **oversampling of low birthweight babies**

    -   Can't "adjust this away" with predictors -- because the outcome
        (birthweight) is the design variable!

## How Do I Know If Weighting Matters for MY Regression? {.pause}

-   General consensus is you should use weights in the regression when:

    -   The weight is a function of the dependent (outcome) variable

    -   You don't know enough about how the weights were constructed\
        (may be the case for some publicly available datasets)

    -   Large sample size -- even if weights are not strictly necessary,
        the loss of power from using weights won't be as much of a
        concern

. . .

-   Possible ways to check to see if weighted regression is necessary:

    -   Perform [both weighted and unweighted regressions]{.alert}, and
        compare coefficients estimates

        -   If not significantly different, weights not necessary

    -   Add the [weight and an interaction of the weight with each
        predictor]{.alert} to the model

        -   If coefficients of weights and interactions are not
            significant, weights not necessary

. . .

-   Remember: If weights "don't matter" for regression estimates,
    penalty for using weighted regression anyway is a loss of efficiency
    (variance too large) -- might not be a big deal with a large sample
    size (like many surveys)

## Details: What IS Weighted Regression?

-   From a finite population perspective, linear regression is a fitting
    of a line to a finite population of $N$ units

    -   Describing the relationship between outcome $y$ and predictor
        $x$ in a finite set of units

::::: columns
::: {.column width="60%"}
-   Thinking specifically about a simple linear regression model, we
    have the best fitting **least squares** regression line:
    $$y = B_0 + B_1 x$$

    -   $B_0$ and $B_1$ are the coefficients that minimize the quantity:
        $$\sum_{i =1}^N (y_i - B_0 - B_1 x_i)^2$$

    <!-- -   Note the summation over the **entire finite population** -->
:::

::: {.column width="40%"}
```{r}
#| fig-alt: "Scatterplot of a finite population of 1000 points, with x-axis labeled X and y-axis labeled y. The relationship is linear and shows a positive association. There is a red line showing the least squares regression line, and the values of the slope and intercept are shown in the top left as B0 = 9.92 and B1 = 2.06."
#| fig-asp: 0.7
library(tidyverse)
theme_set(theme_bw())
set.seed(9919)
N <- 1000
x <- runif(N)
y <- 10 + 2*x + rnorm(N)
pop <- tibble(x = x, y = y)
fit <- lm(y ~ x, data=pop)
B0 <- as.numeric(coef(fit)[1])
B1 <- as.numeric(coef(fit)[2])
lab_B0 <- paste0("B[0] ==", B0)
lab_B1 <- paste0("B[1] ==", B1)
pop %>%
  ggplot(aes(x = x, y = y)) + 
  geom_point(shape=1) +
  geom_smooth(method = "lm", se = FALSE, lwd = 2, col = "red") +
  theme(axis.title = element_text(size = 20), 
        axis.text = element_text(size = 18)) +
  annotate("text", x = 0, y = 14, label = lab_B0, parse = TRUE, hjust = 0, size = 8) + 
  annotate("text", x = 0, y = 13.5, label = lab_B1, parse = TRUE, hjust = 0, size = 8)
```
:::
:::::

-   There is [**no assumption of normality**]{.alert} and [**no
    assumption about (equal) variance**]{.alert}

    -   A line describes the relationship between $x$ and $y$ in the
        finite population, and we take a sample from the population to
        use to estimate this line

## Details (con't)

-   Closed-form expressions for $B_0$ and $B_1$ are:

::: smaller-text
$$\begin{aligned}
    \text{Slope: }B_1 &= \frac{\sum_{i=1}^N x_i y_i - \frac{1}{N} \left(\sum_{i=1}^N x_i \right) \left(\sum_{i=1}^N y_i\right)}{\sum_{i=1}^N x_i^2 - \frac{1}{N} \left(\sum_{i=1}^N x_i \right)^2} &= \frac{t_{xy} - \frac{t_x t_y}{N}}{t_{x^2} - \frac{(t_x)^2}{N}}\\
    \text{Intercept: }B_0 &= \frac{1}{N} \sum_{i=1}^N y_i - B_1 \frac{1}{N} \sum_{i=1}^N x_i &= \frac{t_y-B_1 t_x}{N}
    \end{aligned}$$
:::

-   These are functions of **totals** -- and we know how to estimate
    totals from any complex sampling design!

::: smaller-text
$$\begin{aligned}
    \hat{B}_1 &= \frac{\hat{t}_{xy} - \frac{\hat{t}_x \hat{t}_y}{\hat{N}}}{\hat{t}_{x^2} - \frac{(\hat{t}_x)^2}{\hat{N}}} = \frac{\sum_{i \in \mathcal{S}} w_i x_i y_i - \frac{\left(\sum_{i \in \mathcal{S}} w_i x_i \right) \left(\sum_{i \in \mathcal{S}} w_i y_i\right)}{\sum_{i \in \mathcal{S}} w_i} }{\sum_{i \in \mathcal{S}} w_i x_i^2 - \frac{\left(\sum_{i \in \mathcal{S}} w_i x_i \right)^2}{\sum_{i \in \mathcal{S}} w_i} } \\
    \hat{B}_0 &= \frac{\hat{t}_y-\hat{B}_1 \hat{t}_x}{\hat{N}} = \frac{\sum_{i \in \mathcal{S}} w_i y_i - \hat{B}_1\sum_{i \in \mathcal{S}} w_i x_i}{\sum_{i \in \mathcal{S}} w_i} 
    \end{aligned}$$
:::

## Details (con't) {.pause}

-   Clearly in practice you wouldn't hand-calculate these
    -   In fact, software uses better algorithms that aren't subject to
        as much potential rounding error
-   Also -- we need standard errors!

. . .

-   Once again we can use **linearization** to estimate standard errors

    -   $\hat{B}_0, \hat{B}_1$ are [non-linear]{.alert2} functions of
        totals

    -   Use linearization to approximate these quantities as linear
        functions of totals

    -   Then, estimate the variance of this linear quantity -- since we
        know how to estimate the variance for a total from any complex
        sample design

. . .

-   Use SEs to compute CIs and perform hypothesis tests (e.g., Wald
    tests for regression coefficients)

## Details (con't) {.pause}

-   **Note #1: Subtle difference in CI interpretation from infinite
    population inference**

    -   With a design-based CI, the confidence level (e.g., 95%) for a
        coefficient (say, $B_1$) is the [sum of the probabilities of all
        possible samples $\mathcal{S}$]{.alert} that could be taken with
        the specified design [for which the CI constructed from that
        sample will contain the true $B_1$]{.alert}

<br>

. . .

-   **Note #2: Survey-Weighted Regression is not Weighted Least Squares
    (WLS)**

    -   The point estimates you get from survey weighted linear
        regression will be the **same as** point estimates from WLS --
        but the standard errors **will not be the same**.
    -   We need the linearization estimator to properly account for the
        sample design

## Degrees of Freedom

-   In the [**infinite population**]{.alert} setting:

    -   Estimating a mean: DF = $n-1$

    -   Estimating a regression coefficient: DF = $n-1-p$   ($p$ = \#
        coefficients excluding the intercept)

-   For [**finite population**]{.alert2} sampling:

    -   Estimating a mean: DF = \# PSUs $-$ \# Strata

    -   Estimating a regression coefficient:

        -   **Option 1:** DF = \# PSUs $-$ \# Strata $-$ $p$

        -   **Option 2:** DF = \# PSUs $-$ \# Strata     (no extra
            subtraction of $p$)

-   **Option 1** is:

    -   *Correct* if all covariates are *PSU-level*
    -   *Very conservative* if covariates are *SSU-level*

-   Different software does different things by default...

-   Sometimes (for designs with small DF) these differences in DF can
    result in observably different p-values

## Example: Sampling Schools

```{r}
library(haven)
library(survey)
# pop <- read_dta("./datasets/apipop.dta") %>%
#   mutate(stype = factor(stype, levels = 1:3, labels = c("E","M","H")))
samp <- read_dta("./datasets/apipop_clustersample_2stage.dta") %>%
  mutate(stype = factor(stype, levels = 1:3, labels = c("E","M","H")))

# fit_pop <- lm(api00 ~ ell + stype, data = pop)
# summary(fit_pop)

# pop %>%
#   ggplot(aes(x = ell, y = api00, color = stype, shape = stype)) + geom_point(size = 1) +
#   geom_abline(intercept = coef(fit_pop)[1], slope = coef(fit_pop)[2], color = "red", lwd = 1.5) + 
#   geom_abline(intercept = coef(fit_pop)[1] + coef(fit_pop)[3], slope = coef(fit_pop)[2], color = "green", lwd = 1.5) + 
#   geom_abline(intercept = coef(fit_pop)[1] + coef(fit_pop)[4], slope = coef(fit_pop)[2], color = "blue", lwd = 1.5) + 
#   lims(x = c(0,100), y = c(300, 1000))

# samp %>%
#   ggplot(aes(x = ell, y = api00, color = stype, shape = stype)) + geom_point(size = 2) +
#   geom_abline(intercept = coef(fit_samp)[1], slope = coef(fit_samp)[2], color = "red", lwd = 1.5) + 
#   geom_abline(intercept = coef(fit_samp)[1] + coef(fit_samp)[3], slope = coef(fit_samp)[2], color = "green", lwd = 1.5) + 
#   geom_abline(intercept = coef(fit_samp)[1] + coef(fit_samp)[4], slope = coef(fit_samp)[2], color = "blue", lwd = 1.5) + 
#   lims(x = c(0,100), y = c(300, 1000))

des <- svydesign(id = ~dnum + snum, weight = ~wt, fpc = ~N + Mi, data=samp)
fit_samp <- svyglm(api00 ~ ell + stype, design = des)
```

-   Data from Lecture 13

-   Two-stage cluster sample (no stratification)

    -   Sampled 38 PSUs (districts)

    -   Sampled 1-2 SSUs (schools) per PSU (district)

    -   Total of 65 SSUs (schools) sampled

-   Regression of interest is predicting API (performance index) from
    percent English language learners and school type
    (Elementary/Middle/High School)

    -   **Option 1:** DF = \# PSUs $-$ \# Strata = $38 - 1 = 37$

    -   **Option 2:** DF = \# PSUs $-$ \# Strata $-$ \# predictors
        excluding intercept = $38 - 1 - 3 = 34$

## Example (con't)

```{css, echo=FALSE}
.cell .cell-output-stdout pre code {
  font-size: 0.5em !important;
  max-height: 650px; // Adjust this value as needed
  overflow-y: auto; // Add this to handle overflow
}

.reveal code {
  font-size: 0.55em !important;
}
```

::::: {columns}
::: {.column width="50%"}
**DF** = \# PSUs $-$ \# Strata = 37

```{r}
#| echo: true
summary(fit_samp, df.resid=degf(des))
```
:::

::: {.column width="50%"}
**DF** = \# PSUs $-$ \# Strata $-$ $p$ = 34

```{r}
#| echo: true
summary(fit_samp)
```
:::
:::::

-   The $t$-distributions with 34 and 37 are pretty similar, doesn't
    make much difference which DF you use

-   But in a design with fewer PSUs, it might matter more

## References for Futher Reading

-   Interested in more perspectives on whether or not to use weights in
    analyzing survey data?

    -   Bollen KA, Biemer PP, Karr AF, Tueller S, and Berzofsky
        ME. (2016) Are survey weights needed? A review of diagnostic
        tests in regression analysis. *Annual Review of Statistics and
        Its Application*, 3:375-392.

    -   Solon G, Haider SJ, and Wooldridge JM (2015). What are we
        weighting for? *J. Human Resources*, 50:301-316

    -   Gelman A (2007). Struggles with survey weighting and regression
        modeling. *Statistical Science*, 22(2):153-164

    -   Winship C and Radbill L (1994). Sampling weights and regression
        analysis. *Sociological Methods and Research*, 23:230-257

(PDFs on Carmen)

-   Final thought: **USE SURVEY-WEIGHTED REGRESSION**

## Activity 20.1

Linear Regression with NHANES
