---
title: "Simple Random Sampling: Software (Stata, R)"
subtitle: "PUBHBIO 7225 Lecture 5"
# author: "Generative AI acknowledgment: MS Copilot was used to generate alt text for images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
    eval: false
    echo: true
    # code-line-numbers: false
    highlight-style: arrow #default
---

## Outline

::::: columns
::: {.column width="50%"}
**Topics**

-   Stata for SRS
-   R for SRS
:::

::: {.column width="50%"}
**Activities**

-   5.1 Stata/R for SRS
:::
:::::

<br>

**Assignments**

-   Quiz 1 due Thursday 9/11/2025 11:59pm via Carmen
-   Problem Set 2 due Thursday 9/18/2025 11:59pm via Carmen

## Stata Commands for SRS

Select an SRS of size 100:

```{r}
set seed 72947
sample 100, count
```

Select an SRS that is 15% of the population:

```{r}
set seed 72947
sample 15
```

Create a new variable called `N` that contains the value 10,000 for all records:

```{r}
generate N = 10000
# can abbreviate with:
gen N = 10000
```

Tell Stata about the sample design for an SRS:

```{r}
svyset [pweight=WEIGHTVAR], fpc(N)
# WEIGHTVAR = variable containing sampling weights
# N = variable that contains finite population size
```

## (Some) Stata Commands for Survey Estimation

Must first run `svyset` command. Variables are `y`, `x1`, `x2`.

Estimate a mean:

```{r}
svy: mean y
```

Estimate a proportion:

```{r}
svy: proportion x1
# can abbreviate with:
svy: prop x1
```

Estimate a total:

```{r}
svy: total y
```

Linear and logistic regression:

```{r}
# linear regression (y on x1 and x2)
svy: regress y x1 x2
# logistic regression (x1 on x2)
svy: logistic x1 x2
```

## R Commands for SRS

Select an SRS of size 100 (from a dataset called `POPDATA`):\

```{r}
set.seed(72947)
DF <- POPDATA %>% slice_sample(n = 100)
# DF = resulting dataset with 100 observations
```

Select an SRS that is 15% of the population (from a dataset called `POPDATA`):

```{r}
set.seed(72947)
DF <- POPDATA %>% slice_sample(prop = 0.15)
```

Create a new variable called `N` that contains the value 10,000 for all records:

```{r}
DF$N <- 10000
# or, tidyverse-style:
DF <- DF %>% mutate(N = 10000)
```

Tell R about the sample design for an SRS:

```{r}
DESIGN.OBJECT <- svydesign(id = ~1, weights = ~WEIGHTVAR, fpc = ~N, data=DF)
# WEIGHTVAR = variable containing sampling weights
# N = variable that contains finite population size
```

## (Some) R Commands for Survey Estimation

Must first use `svydesign()` to make survey design object. Variables are `y`, `x1`, `x2`.

Estimate a mean:

```{r}
svymean(~y, design=DESIGN.OBJECT) # DESIGN.OBJECT = design object from svydesign()
```

Estimate a proportion:

```{r}
# proportions at each level of the factor
svymean(~factor(x1), design=DESIGN.OBJECT)
# if variable is 1/0 or TRUE/FALSE can use svymean() to get proportion
svymean(~x1, design=DESIGN.OBJECT)
```

Estimate a total:

```{r}
svytotal(~x1, design=DESIGN.OBJECT)
```

Linear and logistic regression:

```{r}
# linear regression (y on x1 and x2)
svyglm(y ~ x1 + x2, design=DESIGN.OBJECT)
# logistic regression (x1 on x2)
svyglm(x1 ~ x2, family=binomial(link="logit"), design=DESIGN.OBJECT)
```

## Activity 5.1

Stata/R for SRS
