---
title:  |
  Two-stage
  Cluster Sampling \
subtitle: "PUBHBIO 7225 Lecture 13"
# author: "Generative AI acknowledgment: MS Copilot was used to generate alt text for some images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
editor: 
  markdown: 
    wrap: 72
---

## Outline

::: hidden
\newcommand{\that}{\hat{t}}
\newcommand{\tbar}{\bar{t}}
\newcommand{\thatclus}{\hat{t}_{clus}}
\newcommand{\ybar}{\bar{y}}
\newcommand{\ybarclus}{\hat{\bar{y}}_{clus}}
\newcommand{\ybarr}{\hat{\bar{y}}_r}
\newcommand{\yij}{y_{ij}}
\newcommand{\Mbarhat}{\hat{\bar{M}}}
\newcommand{\fpcpar}{\left(1-\frac{n}{N}\right)}
\newcommand{\Cov}{\text{Cov}}
:::

::::: columns
::: {.column width="60%"}
**Topics**

-   Two-stage Cluster Sampling

    -   Unbiased Estimators

    -   Ratio Estimators

    -   Sampling Weights

    -   Variance Estimation
:::

::: {.column width="40%"}
**Activities**

-   13.1 Two-Stage Cluster Sampling
:::
:::::

**Readings**

-   Handout on Carmen with mathematical details on variance estimation
    (OPTIONAL)

**Assignments**

-   Quiz 3 due Thursday 10/9/2025 11:59pm via Carmen
-   Group Progress Report due Thursday 10/9/2025 11:59pm via Carmen
    (only one group member needs to upload this)

## Two-Stage Cluster Sampling

-   In **one-stage** cluster sampling, we sample **all SSUs** within the
    sampled PSUs

-   In **two-stage** cluster sampling, we only sample **some SSUs**
    within the sampled PSUs

-   Why might we do this?

. . .

-   SSUs within a cluster may be so similar that measuring all is a
    waste of resources

-   Might be expensive to sample SSUs, so only a subsample is taken for
    cost reasons

-   Might not be practical to sample all SSUs (ex: all people in a
    household)

## Stages of Two-Stage Cluster Sampling

1.  Select an SRS of $n$ PSUs from population of $N$ PSUs

2.  Select an SRS of $m_i$ SSUs from the $M_i$ SSUs in PSU $i$

-   The extra stage of sampling complicates the notation and estimators

    -   Two samples:
        -   $\mathcal{S}$ = the sample of PSUs
        -   $\mathcal{S}_i=$ the sample of SSUs from PSU $i$

-   Note that quite often in large-scale surveys something other than
    SRS is used at the first stage -- often it is **probability
    proportional to size (PPS)** sampling

-   We will return to this concept next lecture (but it is exactly what
    it sounds like!)

## Two Stages: More Variability, More Choices {.pause}

> **More Variability:** Now there is sample-to-sample variability from
> **both** stages of data collection

-   A different two-stage sample could contain the same or different
    PSUs

-   Even if (some/all) same PSUs are selected, a different set of SSUs
    could be sampled from those PSUs

-   As a result, we should expect two contributions to ("pieces of") the
    variance of estimators:

    -   \(1\) variability from sampling PSUs, and (2) variability from
        sampling SSUs within PSU

. . .

> **More Design Choices:** Have to decide [how many SSUs per PSU to
> sample]{.alert2}

-   We [do not have to sample the same number of SSUs from each
    PSU]{.alert}

-   If an PSU with only 1 SSU is sampled, the only choice is to sample 1
    SSU ("all")

-   Unlike stratified sampling, where you need at least 2 units per
    stratum (to estimate variance), [**you can sample just 1 SSU in a
    sampled PSU**]{.alert} (and often we do!)

    -   Ex: PSU = housing unit, SSU = person → can sample just 1 person
        in a household

## Example: Sampling School Districts {.pause}

```{r}
library(tidyverse)
library(haven)
library(survey)
theme_set(theme_bw())
pop <- read_dta("./datasets/apipop.dta")
M0 <- nrow(pop)
N <- length(unique(pop$dnum))
samp <- read_dta("./datasets/apipop_clustersample_2stage.dta")
n_psu <- length(table(samp$dnum))
n_ssu <- nrow(samp)
pop %>%
  group_by(dnum) %>%
  tally() -> num_psus
samp_psus <- num_psus %>% 
  filter(dnum %in% samp$dnum)
```

Population: $M_0=$ `r M0` schools in $N=$ `r N` school districts

::::: columns
::: {.column width="50%"}
-   PSU = school district

-   SSU = school

-   $y$ = school's Academic Performance Index\
    (based on standardized test results)

-   True population mean API: $\bar{y}_U =$ `r round(mean(pop$api00),1)`

-   Highly skewed number of SSUs per PSU
:::

::: {.column width="50%"}
```{r, out.width='100%'}
#| fig-cap: ""
#| fig-alt: "histogram of number of SSUs (schools) per PSU (district), showing an extremely right-skewed distribution"
num_psus %>%
  ggplot(aes(x = n)) + 
  geom_histogram(fill = "lightgreen", color = "black", binwidth = 5) +
  labs(y = "Frequency", x = "Number of Schools per District (SSUs per PSU)") +
  theme(axis.title = element_text(size = 18),
        axis.text = element_text(size = 16))
```
:::
:::::

. . .

**Two-stage cluster sample taken:**

-   **Stage 1:** $n =$ `r n_psu` PSUs (districts) out of $N=$ `r N` --
    roughly 5% of total

    -   Selected PSUs have a range of $M_i=$ `r min(samp_psus$n)` to
        `r max(samp_psus$n)` SSUs

. . .

-   **Stage 2:** $m_i=$ 2 SSUs (schools) per PSU

    -   For PSUs with only 1 SSU, $m_i=M_i=1$ (can only select the 1
        school!)

## Example (con't) {#slide-school-sample-figure}

-   Result is a total of `r nrow(samp)` SSUs (schools)
    -   Some sampled PSUs (districts) had only 1 SSU (school)
-   API $(y)$ by PSU (sampled SSUs only):

```{r}
#| fig-cap: ""
#| fig-alt: "Scatterplot of PSU (district) number on the x-axis and API score (y) on the y-axis. There are 1 or 2 points per PSU, reflecting the sampling of 1 or 2 SSUs per PSU."
#| fig-align: center
samp %>%
  ggplot(aes(x = factor(dnum), y = api00)) +
  geom_point(size = 3, shape = 1) + 
  labs(y = "API (y)", x = "PSU (district)") + 
  theme(axis.title = element_text(size = 18))
```

-   Number of SSUs per PSU is either 1 or 2 (by design)

-   We want to use this 2-stage cluster sample to estimate the overall
    mean API

## Activity 13.1 (Part 1)

Two-Stage Cluster Sampling (Part 1)

## Similar Estimation as One-Stage Cluster Sample {.pause}

-   **One-Stage Cluster Sample:** (SRS of $n$ PSUs out of $N$ total;
    take all $M_i$ SSUs in the selected PSUs):

    -   To estimate the mean of $y$ we...

        -   observe the true $t_i$ (total of $y$ for cluster $i$) in
            each sampled PSU,

        -   calculate the average cluster total across the sampled PSUs
            $(\bar{t})$,

        -   multiply this average by the number of PSUs ($N$) to get a
            grand total $(\hat{t}_{clus})$

        -   and divide by the total number of SSUs $(M_0)$ to get the
            unbiased estimator $(\hat{\bar{y}}_{clus})$

        -   OR, divide by the estimated total number of SSUs
            $(\widehat{M}_0)$ to get the ratio estimate
            $(\hat{\bar{y}}_r)$

. . .

-   In a **Two-Stage Cluster Sample**, **we don't observe the true**
    $\mathbf t_i$ for each sampled PSU

    -   However, we can *estimate* the PSU-specific totals because we
        took an SRS from each PSU!

    -   Essentially this is the only change to the estimation process --
        add in a step to estimate the PSU totals $(\hat{t}_i)$ and then
        use them like we used the true totals in one-stage cluster
        sampling

    -   In other words, replace $t_i$ with $\hat{t}_i$ above (and
        $\bar{t}$ with $\bar{\hat{t}}$ )

    -   Of course, variance estimation is much messier...

## Step 1: Cluster-Specific Estimates

Start with **cluster-specific estimates**:

::: smaller-text
| **Quantity** | **Estimate** | **In Words** |
|:---|:---|:---|
| PSU Mean (of $y$) | $\displaystyle \bar{y}_i= \frac{1}{m_i} \sum_{j\in \mathcal{S}_i} y_{ij}$ | average of $y$ for sampled SSUs in PSU $i$ |
| PSU Total (of $y$) | $\displaystyle \hat{t}_i = M_i \bar{y}_i$ | average of $y$ for sampled SSUs times the total number of SSUs in PSU $i$ |
| Variance of $y$ for SSUs within PSU | $\displaystyle s_i^2 = \frac{1}{m_i-1} \sum_{j \in \mathcal{S}_i} \left(y_{ij}-  \bar{y}_i\right)^2$ | variance of $y$ for sampled SSUs in PSU $i$ |
:::

-   These look messy -- but are just the SRS formulae applied within
    each PSU!

-   Note that the total, $\hat{t}_i$, has a "hat" on it (its an
    estimate) -- different from 1-stage sampling

-   If we have taken an SRS within PSU, then by properties of SRS,
    estimates are unbiased:

    -   $E[\bar{y}_i] = \bar{y}_{Ui}$

    -   $E[\hat{t}_i] = t_i$

    -   $E[s_i^2] = S_i^2$

## Step 2: Overall Estimates

Then combine cluster-specific estimates to get **overall estimates**:

::: smaller-text
| **Quantity** | **Estimate** | **In Words** |
|:---|:---|:---|
| Mean PSU Total (of $y$) | $\displaystyle \bar{\hat{t}} = \frac{1}{n} \sum_{i \in \mathcal{S}} \hat{t}_i$ | average of estimated PSU totals from sampled PSUs |
| Overall Total (of $y$) | $\displaystyle \hat{t}_{clus}= N\bar{\hat{t}}$ | estimated average PSU total times the number of PSUs (in pop.) |
| Overall Mean (of $y$) | $\displaystyle \hat{\bar{y}}_{clus}= \frac{\hat{t}_{clus}}{M_0}$ (unbiased estimator) | estimated total divided by **known** total number of SSUs |
|  | $\displaystyle \hat{\bar{y}}_r= \frac{\hat{t}_{clus}}{\widehat{M}_0}$ (ratio estimator) | estimated total divided by **estimated** total number of SSUs |
| Variance of PSU Totals | $\displaystyle s^2_t = \frac{1}{n-1} \sum_{i \in \mathcal{S}} \left(\hat{t}_i - \bar{\hat{t}} \right)^2$ | variance of $\hat{t}_i$ for sampled PSUs |
:::

-   Note that these steps look just like the steps for one-stage cluster
    samples -- but with (additional) "hats" on the pieces involving $t$
    (e.g., $\hat{t}_i, \bar{\hat{t}}$ )

## Steps, Combined

-   So to estimate the mean of $y$ we...

    -   calculate the mean of $y$ for each sampled PSU $(\bar{y}_i)$,

    -   multiply this average by the number of SSUs per PSU ($M_i$) to
        get the estimated totals in each PSU $(\hat{t}_i)$

    -   calculate the average of the estimated cluster totals
        $\hat{t}_i$ across the sampled PSUs $(\bar{\hat{t}})$,

    -   multiply this average by the number of PSUs ($N$) to get a grand
        total $(\hat{t}_{clus})$

    -   and divide by the total number of SSUs $(M_0)$ to get the
        unbiased estimate $(\hat{\bar{y}}_{clus})$

    -   OR, divide by the estimated total number of SSUs
        $(\widehat{M}_0)$ to get the ratio estimate $(\hat{\bar{y}}_r)$

## Expectation of the Estimators {.pause}

-   $\hat{t}_{clus}$ is an **unbiased** estimator of the true population
    total $t_U$ (derivation next slide)

    -   $E[\hat{t}_{clus}] = t_U$

-   Thus the [**unbiased estimator**]{.alert2} of the mean is an
    **unbiased** estimator of the true population mean, $\bar{y}_U$

    -   $E[\hat{\bar{y}}_{clus}] = E[\hat{t}_{clus}/M_0] = E[\hat{t}_{clus}]/M_0 = t_U/M_0 = \bar{y}_U$

-   But the [**ratio estimate**]{.alert} of the mean is **biased**, just
    like with one-stage cluster sampling

    -   $E[\hat{\bar{y}}_r] = E[\hat{t}_{clus}/\widehat{M}_0] \ne \bar{y}_U$

    -   However, as in that scenario, the drastic decrease in variance
        for $\hat{\bar{y}}_r$ outweighs its small bias, making it the
        preferred estimator

<p>

. . .

-   For multi-stage sampling we prove theoretical results by
    **conditioning on a realized sample** and then taking the
    expectation -- using the property of successive conditioning:
    $E(A) = E[E(A|B)]$

-   Inclusion indicator $Z_i$ (for the $i$th PSU) is a random variable

-   For equal probability sampling of PSUs:
    $P(Z_i=1) = \pi_i = n/N = E(Z_i) = E(Z_i^2)$

## Expectation of the Estimators (con't)

::: smaller-text
$$\begin{aligned}
    E(\hat{t}_{clus}) = E(N\bar{\hat{t}}~) = E \left(N \sum_{i \in \mathcal{S}} \frac{1}{n} \hat{t}_i \right) &= E \left(\sum_{i = 1}^N Z_i \frac{N}{n} \hat{t}_i \right) = E \left[ E \left(\sum_{i = 1}^N Z_i \frac{N}{n}\hat{t}_i  \bigg| Z_1, \dots, Z_N \right) \right] \\
    &= E \left[ \sum_{i = 1}^N Z_i \frac{N}{n} E\left(\hat{t}_i  \bigg| Z_1, \dots, Z_N \right) \right] \\
    &= E \left[ \sum_{i = 1}^N Z_i \frac{N}{n} E\left(\hat{t}_i \right) \right] \quad \text{because } \hat{t}_i \perp (Z_1,\dots,Z_N)^*\\
    &= E \left( \sum_{i = 1}^N Z_i \frac{N}{n} t_i \right) \quad \text{because } E(\hat{t}_i) = t_i \\
    &= \sum_{i = 1}^N E(Z_i) \frac{N}{n} t_i \\
    &= \sum_{i = 1}^N \frac{n}{N} \frac{N}{n} t_i = \sum_{i = 1}^N t_i = t_U \quad \text{unbiased!}
    \end{aligned}$$

$$E(\hat{\bar{y}}_{clus})  = E ( \hat{t}_{clus}/ M_0) = E ( \hat{t}_{clus})/M_0 = t_U/M_0 = \bar{y}_U \quad \text{unbiased!}$$

\*Expected value of the estimated total in PSU $i$ (based on the second
stage sampling) is independent of whether or not the PSU is selected
into the sample
:::

## Example (con't)

**Step 1: Cluster-Specific Estimates:** For each of the $n=$ `r n_psu`
sampled PSUs (districts):

::: smaller-text
| PSU | $M_i$ | $m_i$ | SSU values of $y$ | $\bar{y}_i$ | $\hat{t}_i = M_i \bar{y}_i$ |
|:---|:---|:---|:---|:---|:---|
| 1 | 10 | 2 | 753, 807 | 780 | 10 $\times$ 780 = 7800 |
| 2 | 10 | 2 | 762, 743 | 752.5 | 10 $\times$ 752.5 = 7525 |
| 3 | 19 | 2 | 536, 540 | 538 | 19 $\times$ 538 = 10222 |
| 4 | 1 | 1 | 524 | 524 | 1 $\times$ 524 = 524 |
| $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ |
:::

**Step 2: Combine to get Overall Estimates:**

```{r}
samp %>%
  group_by(dnum, Mi) %>%
  summarize(ybar = mean(api00)) %>%
  ungroup() %>%
  mutate(that = Mi*ybar) -> stats
# mean(stats$that)
# N*mean(stats$that)
# N*mean(stats$that)/M0
```

-   Mean PSU Total =
    $\bar{\hat{t}} = \frac{7800 + 7525 + 10222 + \dots}{38} = 4252.013$

-   Overall Estimated Total =
    $\hat{t}_{clus}= N\bar{\hat{t}} = 757 \times 4252.013 = {3218774}$

-   Overall Mean, [**Unbiased Estimator**]{.alert2} =
    $\frac{\hat{t}_{clus}}{M_0} = \frac{{3218773.8}}{6194} = {519.7}$

<!-- -   Overall Mean, **Ratio Estimator** = -->

<!--     $\frac{\hat{t}_{clus}}{\widehat{M}_0}$ $\leftarrow$ Need estimated -->

<!--     population size, $\widehat{M}_0$ -->

-   For **Ratio Estimator:** Estimated population size =
    $\widehat{M}_0 = \sum_{i \in \mathcal{S}} \sum_{j \in \mathcal{S}_i} w_{ij}$
    = sum of weights (sampled SSUs)

## Sampling Weights for Two-Stage Cluster Samples

-   For two-stage cluster sampling with **SRS** at first and second
    stage:

> [**Sample weight for sampled SSU** $j$ **in PSU** $i$]{.alert2}:  
> $\displaystyle w_{ij} = \frac{1}{\pi_{ij}} = \frac{N}{n} \frac{M_i}{m_i}$

-   Intuition:

    -   Each sampled PSU "represents" $\frac{N}{n}$ PSUs

    -   And within a PSU each sampled SSU "represents" $\frac{M_i}{m_i}$
        SSUs

    -   So in total a sampled SSU "represents"
        $\frac{N}{n} \times \frac{M_i}{m_i}$ SSUs

-   Weight is the inverse of the probability of selection:

    $$\begin{align}
     \pi_{ij} &= P(\text{SSU $j$ of PSU $i$ in sample}) = P(\text{SSU $j$ selected}|\text{PSU $i$ selected}) \times P(\text{PSU $i$ selected})\\
    &= \frac{m_i}{M_i} \times \frac{n}{N} = \frac{n}{N}\frac{m_i}{M_i}
    \end{align}$$

## Example (con't)

```{r}
# sum(samp$wt)
```

-   To get the ratio estimator, we need the estimated population size:

    -   $\widehat{M}_0 =  \sum_{i \in \mathcal{S}} \sum_{j \in \mathcal{S}_i} w_{ij}$
        = sum of weights, sampled SSUs = 4900.579 (from software)[^1]

-   Overall Mean, [**Ratio Estimator**]{.alert} =
    $\displaystyle \frac{\hat{t}_{clus}}{\widehat{M}_0} = \frac{3218773.8}{4900.6} = 656.8$

-   We can also write estimates in terms of **weights** (that's how
    software is calculating these!)

    -   $\displaystyle \hat{t}_{clus}= \sum_{i \in \mathcal{S}} \sum_{j \in \mathcal{S}_i} w_{ij} y_{ij}$
        = weight $\times$ y, added up for all sampled SSUs

    -   $\displaystyle \hat{\bar{y}}_r= \frac{\sum_{i \in \mathcal{S}} \sum_{j \in \mathcal{S}_i} w_{ij} y_{ij}}{\sum_{i \in \mathcal{S}} \sum_{j \in \mathcal{S}_i} w_{ij}} = \frac{\text{ weight $\times$ y, added up for all sampled SSUs}}{\text{sum of weights for all sampled SSUs}}$

[^1]: You can hand-calculate $\widehat{M}_0$ as
    $\widehat{M}_0 = N \hat{\bar{M}}= N \frac{1}{n} \sum_{i \in S} M_i$
    if you wanted to, but do you really want to?

## EPSEM Sampling {.pause}

-   In the schools example, the sampling weights **differed** across the
    65 sampled PSUs:

```{r}
summary(samp$wt)
```

<p>

-   Can a two-stage cluster sample be EPSEM (have all sampling weights
    **equal**)?

. . .

-   Q: When will $\frac{n}{N}\frac{m_i}{M_i}$ be the same for all PSUs?
    (for all $i$)

. . .

-   A: When $\frac{m_i}{M_i}$ is the same for all PSUs (for all $i$)!

    -   In other words, when the same **proportion** of SSUs are sampled
        from each PSU

<p>

. . .

-   A two-stage cluster sample (with SRS at the first stage) is only
    **self-weighting (EPSEM)** if [more SSUs are selected from the
    larger PSUs]{.alert} (kind of like proportional allocation to
    strata)

## Activity 13.1 (Part 2)

Two-Stage Cluster Sampling (Part 2)

## Why We Focus on Estimating the Total

-   We will spend a bit of time calculating the **variance of the
    estimated total** (and how to estimate it)

-   But -- our main interest is often **not** in the totals!

    -   Possible exception: estimating the number of people with a
        condition

-   So why do we focus on the total?

<p>

. . .

-   Nearly every quantity we want to estimate can be written as a
    function of **totals**

-   If we can write the quantity in terms of totals, we can use
    **linearization** to estimate its variance -- like we did with the
    ratio estimate of the mean in 1-stage cluster sampling

-   Linearization can be used for a wide range of estimates:

    -   Means (ratio estimator)

    -   Domain (subgroup) estimates

    -   Linear regression coefficients

    -   Logistic regression coefficients

    -   etc.

## Variance Estimation for the Estimated Total

-   As it turns out, $V(\hat{t}_{clus})$ under two-stage cluster
    sampling is equal to:

    -   the variance under one-stage cluster sampling, **plus**

    -   an extra term to account for the fact that we are **estimating**
        the $\hat{t}_i$\
        (as opposed to directly observing the true $t_i$ in one-stage
        cluster sampling)

-   True (theoretical) variance of the estimated total[^2]:
    $$V(\hat{t}_{clus}) = \underbrace{N^2 \left(1-\frac{n}{N}\right)\frac{S^2_t}{n}}_{V_{stage1}} + \underbrace{\frac{N}{n} \sum_{i=1}^N M_i^2 \left(1-\frac{m_i}{M_i}\right) \frac{S_i^2}{m_i}}_{V_{stage2}}$$

    $S^2_t$ = true (population) variance of the PSU totals, $t_i$\
    $S_i^2$ = true variance of the $y_{ij}$ within PSU $i$

-   Note that both these pieces look like SRS formulae -- we take an SRS
    at each stage.

[^2]: For derivation, see handout on Carmen

## Variance Estimation for the Estimated Total (con't)

::: smaller-text
-   Replacing the unknown variances $S_t^2$ and $S_i^2$ with their
    sample estimates gives us:

$$\widehat{V}(\hat{t}_{clus}) = \underbrace{N^2 \left(1-\frac{n}{N}\right)\frac{s^2_t}{n}}_{\widehat{V}_{stage1}} + \underbrace{\frac{N}{n} \sum_{i \in \mathcal{S}} \left(1-\frac{m_i}{M_i}\right)M_i^2 \frac{s_i^2}{m_i}}_{\widehat{V}_{stage2}}$$

-   $\widehat{V}_{stage1}$ = variability from sampling at the first
    stage (PSUs)

    -   Same as the variance formula for one-stage cluster sampling!

-   $\widehat{V}_{stage2}$ = added variability from sampling at the
    second stage (SSUs within PSUs)

    -   Note that if all SSUs are selected (as in one-stage cluster
        sampling), $m_i=M_i$ and $\widehat{V}_{stage2}=0$

-   In many (most) situations, when $N$ is large (and FPC small),
    $\widehat{V}_{stage1} >> \widehat{V}_{stage2}$

    -   Variance of the $\hat{t}_i$ can be very large -- affected by
        differences in cluster sizes $(M_i)$ and by differences in
        cluster means $(\bar{y}_i)$

    -   Even if clusters have similar means, different sizes will lead
        to stage one piece of variance dominating second stage piece

-   If $N$ is large then $\widehat{V}_{stage1}$ is **approximately
    unbiased** for the true variance, $V(\hat{t}_{clus})$ [^3]
:::

[^3]: See math details in handout on Carmen if interested.

## Variance Estimation in Software {.pause}

-   Many software packages simplify calculations by only using
    $\widehat{V}_{stage1}$

    -   $\widehat{V}_{stage1}$ is "close enough" to the true variance of
        the total, since it dominates $\widehat{V}_{stage2}$

. . .

-   Might not be able to calculate $\widehat{V}_{stage2}$ because:

    -   You don't have the PSU sizes $(M_i)$ for sampled PSUs

        -   The people who created the dataset had to have PSU sizes (to
            calculate weights!), but they may not be provided to the end
            user for publicly available datasets

    -   You might have sampled only 1 SSU per PSU, so cannot calculate
        $s_i^2$ (within-PSU variance of the SSUs)

        -   Very common design when sampling people within households

. . .

-   If the software is ignoring the second stage variance $(V_{stage2})$
    then you only have to tell it about the PSUs (i.e., don't need to
    tell it the $M_i$ or an identifier for the SSUs)

-   This is sometimes referred to as an **ultimate PSU analysis**

    <!-- -   This approach is usually used in complex designs with more than -->

    <!--     two stages -->

    -   *It's still a two-stage design, you're just ignoring the second
        stage of sampling when approximating the variance. Sampling
        weights still need to be constructed accounting for the multiple
        stages.*

## Example (con't)

[**Estimator of the Total**]{.alert3}

-   Analysis including information about both stages of sampling

```{r}
#| echo: true
#| classes: .cell-output
des <- svydesign(id = ~dnum + snum, weight = ~wt, fpc = ~N + Mi, data=samp)
svytotal(~api00, design = des)
```

-   Analysis ignoring second stage of sampling ("ultimate PSU analysis")

```{r}
#| echo: true
#| classes: .cell-output
des_ign <- svydesign(id = ~dnum, weight = ~wt, fpc = ~N, data=samp)
svytotal(~api00, design = des_ign)
```

-   Point estimates **identical**

-   Standard error estimates **very similar**

    -   Most variability is from sampling PSUs (not sampling SSUs within
        PSUs)

## Variance Estimation for the Ratio Estimator

-   For the [**unbiased estimator**]{.alert2}, since $M_0$ is a
    constant, we have
    $V(\hat{\bar{y}}_{clus}) = \frac{1}{M_0^2} V(\hat{t}_{clus})$

-   But we know from 1-stage cluster sampling that we can do better
    (smaller variance)

-   We usually use the [**ratio estimator of the mean**]{.alert} for
    2-stage cluster sampling:
    $\displaystyle \hat{\bar{y}}_r= \frac{\hat{t}_{clus}}{\widehat{M}_0}$

    -   Same as one-stage cluster sampling with unequal cluster sizes

    -   **Linearization** is again used to calculate the theoretical
        variance, $V(\hat{\bar{y}}_r)$

        -   We can use the same general result from last time (1-stage
            cluster sampling):
            $$V(\hat{\bar{y}}_r) \approx \frac{1}{M_0^2}\left[V(\hat{t}_{clus}) + \frac{t_U^2}{M_0^2} V(\widehat{M}_0) - 2\frac{t_U}{M_0} \text{Cov}(\hat{t}_{clus},\widehat{M}_0) \right]$$

        -   We just now have a different expression for
            $V(\hat{t}_{clus})$ because of the 2-stage sampling

    -   As we just discussed, often $V(\hat{t}_{clus})$ is approximated
        by only the stage 1 variance

## Variance Estimation for the Ratio Estimator (con't)

-   Variance Estimator:
    $$\widehat{V}(\hat{\bar{y}}_r) = \underbrace{\frac{1}{n \hat{\bar{M}}^2} \left(1-\frac{n}{N}\right)\frac{\sum_{i \in \mathcal{S}} (\hat{t}_i - M_i\hat{\bar{y}}_r)^2}{n-1}}_{\widehat{V}_{stage1}} + \underbrace{\frac{1}{nN\hat{\bar{M}}^2} \sum_{i \in \mathcal{S}}M_i^2 \left(1-\frac{m_i}{M_i}\right) \frac{s_i^2}{m_i} }_{\widehat{V}_{stage2}}$$

-   The first piece, $\widehat{V}_{stage1}$, is almost the same as
    1-stage cluster sampling

    -   Only difference is the estimated PSU total $\hat{t}_i$ in place
        of true PSU total $t_i$

    -   Uses the variability of the estimated PSU totals $(\hat{t}_i)$
        around a PSU-specific estimated total $(M_i\hat{\bar{y}}_r)$,
        instead of the (estimated) overall average $(\bar{\hat{t}})$ --
        thus has lower variance than unbiased estimator

-   As with the variance of $\hat{t}_{clus}$, the second term
    $\widehat{V}_{stage2}$ is usually very small compared to the first
    term

    -   Thus, we often only calculate the first term,
        $\widehat{V}_{stage1}$ ("ultimate PSU analysis")

## Example (con't)

[**Ratio Estimator of the Mean**]{.alert2}

-   Analysis including information about both stages of sampling

```{r}
#| echo: true
#| classes: .cell-output
des <- svydesign(id = ~dnum + snum, weight = ~wt, fpc = ~N + Mi, data=samp)
svymean(~api00, design = des)
```

-   Analysis ignoring second stage of sampling ("ultimate PSU analysis")

```{r}
#| echo: true
#| classes: .cell-output
des_ign <- svydesign(id = ~dnum, weight = ~wt, fpc = ~N, data=samp)
svymean(~api00, design = des_ign)
```

-   Point estimates **identical**

-   Standard error estimates **very similar**

    -   Most variability is from sampling PSUs (not sampling SSUs within
        PSUs) [(Look back at plot of API)](#slide-school-sample-figure)

## Example (con't)

[**Unbiased Estimator of the Mean**]{.alert}

-   **With** Second Stage Variance:

    -   $\widehat{V}(\hat{\bar{y}}_{clus}) = \frac{1}{M_0} \widehat{V}(\hat{t}_{clus}) = \frac{1}{6194} (612674^2) = 60{,}602{,}104$

    -   $\widehat{SE}(\hat{\bar{y}}_{clus}) = \sqrt{\widehat{V}(\hat{\bar{y}}_{clus})} = \sqrt{60{,}602{,}104} = {7{,}785}$

    -   Compare to $\widehat{SE}(\hat{\bar{y}}_r) = {32.65}$

-   **Without** Second Stage Variance:

    -   $\widehat{V}(\hat{\bar{y}}_{clus}) = \frac{1}{M_0} \widehat{V}(\hat{t}_{clus}) = \frac{1}{6194} (612593^2) = 60{,}586{,}081$

    -   $\widehat{SE}(\hat{\bar{y}}_{clus}) = \sqrt{\widehat{V}(\hat{\bar{y}}_{clus})} = \sqrt{60{,}586{,}081} = {7{,}784}$

    -   Compare to $\widehat{SE}(\hat{\bar{y}}_r) = {32.59}$

-   Ratio estimator has **much** smaller SEs!

-   SE is over 200 times larger for unbiased estimator!!

## Activity 13.1 (Part 3)

Two-Stage Cluster Sampling (Part 3)

## A Note on Notation for Variance of Ratio Estimator

-   We have the variance estimator for the ratio estimator for 2-stage
    cluster sampling:
    $$\widehat{V}(\hat{\bar{y}}_r) = \frac{1}{n \hat{\bar{M}}^2} \left(1-\frac{n}{N}\right)\frac{\sum_{i \in \mathcal{S}} (\hat{t}_i - M_i\hat{\bar{y}}_r)^2}{n-1} + \frac{1}{nN\hat{\bar{M}}^2} \sum_{i \in \mathcal{S}}M_i^2 \left(1-\frac{m_i}{M_i}\right) \frac{s_i^2}{m_i}$$

-   We can also write that first stage piece as:

$$\widehat{V}_{stage1}(\hat{\bar{y}}_r) = \frac{1}{\hat{\bar{M}}^2} \left(1-\frac{n}{N}\right)\frac{{\sum_{i \in \mathcal{S}} (\hat{t}_i - M_i\hat{\bar{y}}_r)^2}}{n{(n-1)}} = \frac{1}{\hat{\bar{M}}^2} \left(1-\frac{n}{N}\right)\frac{{s_r^2}}{n}$$

-   ${s_r^2} = \frac{1}{n-1} \sum_{i \in \mathcal{S}} (\hat{t}_i - M_i\hat{\bar{y}}_r)^2$
    = estimated variance of the **residuals**

    -   Dfference between estimated PSU totals, $\hat{t}_i$, and
        predicted total based on the ratio estimator,
        $M_i\hat{\bar{y}}_r$

    -   No "model" involved, but a residual nonetheless
