---
title: |
  Stratified Sampling: Software \
  (Stata, R)
subtitle: "PUBHBIO 7225 Lecture 8"
# author: "Generative AI acknowledgment: MS Copilot was used to generate alt text for images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
    eval: false
    echo: true
    # code-line-numbers: false
    highlight-style: arrow #default
---

## Outline

::::: columns
::: {.column width="50%"}
**Topics**

-   Stata for SRS
-   R for SRS
:::

::: {.column width="50%"}
**Activities**

-   8.1 Stata/R for SRS
:::
:::::

<br>

**Assignments**

-   Peer Evaluation of Problem Set 2 due Tuesday 9/23/25 11:59pm via Carmen
-   Quiz 2 due Thursday 9/25/2025 11:59pm via Carmen
-   Group Progress Report due Thursday 9/25/2025 11:59pm via Carmen (only one group member needs to upload this)

## Stata Commands for Stratified Samples

Select a stratified random sample of $n_h=50$ in each stratum:

```{r}
sort STRATVAR IDVAR     # STRATVAR = variable defining strata, IDVAR = record ID
set seed 217 
by STRATVAR: sample 50, count
```

Select a stratified random sample by taking 50% of each stratum:\

```{r}
sort STRATVAR IDVAR 
set seed 531 
by STRATVAR: sample 50 
```

Tell Stata about the sample design for a stratified sample:\

```{r}
svyset [pweight=WEIGHTVAR], fpc(N_h) strata(STRATVAR) 
# WEIGHTVAR = variable containing sampling weights
# N_h = variable that contains finite population sizes in each stratum
```

With the survey design set, for estimation use same commands as for SRS (e.g., `svy:`)

## R Commands for Stratified Samples

Select a stratified random sample of $n_h=50$ in each stratum:

```{r}
set.seed(217)
DF <- POPDATA %>% group_by(STRATVAR) %>% slice_sample(n = 50)
# STRATVAR = variable defining strata
```

Select a stratified random sample by taking 50% of each stratum:

```{r}
set.seed(217)
DF <- POPDATA %>% group_by(STRATVAR) %>% slice_sample(prop = 0.5)
```

Tell R about the sample design for a stratified sample:

```{r}
DESIGN.OBJECT <- svydesign(id = ~1, weights = ~WEIGHTVAR, fpc = ~N_h, 
                           strata = ~STRATVAR, data=DF)
# WEIGHTVAR = variable containing sampling weights
# N_h = variable that contains finite population sizes in each stratum
```

With the survey design object created, for estimation use same commands as for SRS (e.g., `svymean()`)

## Activity 8.1

Stata/R for Stratified Samples
