---
title:  |
  Cluster Sampling: \
  One-stage with Unequal Size Clusters
subtitle: "PUBHBIO 7225 Lecture 12"
author: "Generative AI acknowledgment: MS Copilot was used to generate alt text for some images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
---

## Outline

<!-- ::: hidden -->

<!-- \newcommand{\fpchpar}{\left(1-\frac{n_h}{N_h}\right)} -->

<!-- \newcommand{\fpcpar}{\left(1-\frac{n}{N}\right)} -->

<!-- ::: -->

::::: columns
::: {.column width="60%"}
**Topics**

-   One-stage Cluster Sampling with Unequal Cluster Sizes
-   Ratio Estimator
:::

::: {.column width="40%"}
**Activities**

-   12.1 Clusters of Unequal Sizes
:::
:::::

<br>

**Assignments**

-   Peer Evaluation of Problem Set 3 due Tuesday 10/7/25 11:59pm via Carmen
-   Quiz 3 due Thursday 10/9/2025 11:59pm via Carmen
-   Group Progress Report due Thursday 10/9/2025 11:59pm via Carmen (only one group member needs to upload this)

## Review of One-Stage Cluster Sampling

-   $N$ = number of PSUs (clusters) in the **population**

-   $n$ = number of PSUs (clusters) in the [**sample**]{.alert}

-   $M_i$ = number of SSUs (e.g., people) in the $i$th PSU in the **population**

-   $m_i$ = number of SSUs (e.g., people) in the $i$th PSU in the [**sample**]{.alert}

-   $M_0$ = total number of SSUs (e.g., people) in the **population**

For **one-stage cluster sampling**:

1.  Take an SRS of $n$ PSUs from a population of $N$ PSUs

2.  Measure $y_{ij}$ for **all** $M_i$ SSUs within the $i$th sampled PSU

## Review of One-Stage Cluster Sampling (con't)

To estimate the **mean of** $y$ we...

1.  Calculate the average cluster total across the sampled PSUs: $\displaystyle \bar{t}= \frac{1}{n} \sum_{i\in \mathcal{S}} t_i$

2.  Multiply average by number of PSUs to get a grand total: $\displaystyle \hat{t}_{clus}= N \bar{t}= \frac{N}{n} \sum_{i\in \mathcal{S}} t_i$

3.  Divide by the total number of SSUs to get an overall mean: $\displaystyle \hat{\bar{y}}_{clus}= \frac{\hat{t}_{clus}}{M_0} = \frac{N\bar{t}}{M_0}$

4.  Estimate variance with: $\displaystyle \widehat{V}(\hat{\bar{y}}_{clus}) = \frac{N^2}{M_0^2} \left(1-\frac{n}{N}\right)\frac{s_t^2}{n}$ where $s_t^2=\widehat{V}(t_i)$

## In a Picture

::::: columns
::: {.column width="80%"}
![](./figures/onestage_illustration){width="100%" fig-alt="Diagram illustrating the process of selecting a Simple Random Sample (SRS) from a population of Primary Sampling Units (PSUs) and aggregating data to totals."}
:::

::: {.column width="20%"}
-   $\bar{t}=$

<br>

<br>

<br>

-   $\hat{t}_{clus}$=

<br>

<br>

<br>

-   $\hat{\bar{y}}_{clus} =$
:::
:::::

## The Problem: Clusters (PSUs) Are Not Usually Equal Sizes

-   [Reality:]{.alert} often PSUs are unequal sizes *(all the examples from last lecture had* $M_i$ the same in all PSUs)

-   Really easy to imagine clusters that are **unequal** in size:

    -   Students within classrooms

    -   Households within Census tracts

    -   Patients within hospitals

    -   Schools within school districts

    -   Etc.

. . .

-   [**Good news:**]{.alert2} the same estimation method **can** be used when **cluster sizes are unequal**

. . .

-   [**Bad news:**]{.alert} when PSUs are unequal in size, often the PSU totals $(t_i)$ are [highly variable]{.alert}, leading to [large $s_t^2$]{.alert} and thus [large $\widehat{V}(\hat{\bar{y}}_{clus})$]{.alert}

## In a Picture

Example from last class: sampling rooms, measuring GPA of all students in a room

::::: columns
::: {.column width="50%"}
**We had all rooms the same size:**

![](./figures/onestage_cluster_totals){width="60%" fig-alt="A comparative data visualization consisting of two tables. The left table presents individual student data with three columns: id, room, and gpa. It includes 20 entries grouped by room numbers, showing student IDs and their corresponding GPAs. The right table aggregates this data, displaying the total GPA per room (all rooms have 4 students) with two columns: room and total of GPA. Red arrows visually link each student's GPA in the left table to the corresponding room's total GPA in the right table, illustrating the summation process to show an SRS of totals."}
:::

::: {.column width="50%"}
**What if rooms were different sizes:**

![](./figures/onestage_cluster_totals_unequal){width="60%" fig-alt="A comparative data visualization consisting of two tables. The left table presents individual student data with three columns: id, room, and gpa. It includes 20 entries grouped by room numbers, showing student IDs and their corresponding GPAs. The right table aggregates this data, displaying the total GPA per room (with rooms of different sizes) with two columns: room and total of GPA. Red arrows visually link each student's GPA in the left table to the corresponding room's total GPA in the right table, illustrating the summation process to show an SRS of totals."}
:::
:::::

-   What happens to the **variability** of the **totals** (when rooms are different sizes)?

## Example: Sampling School Districts

```{r}
library(tidyverse)
library(haven)
library(survey)
theme_set(theme_bw())
pop <- read_dta("./datasets/apipop.dta")
M0 <- nrow(pop)
N <- length(unique(pop$dnum))
```

Population: $M_0=$ `r M0` schools in $N=$ `r N` school districts

::::: columns
::: {.column width="50%"}
-   PSU = school district

-   SSU = school

-   $y$ = Academic Performance Index (based on standardized test results)

-   True population mean API: $\bar{y}_U =$ `r round(mean(pop$api00),1)`
:::

::: {.column width="50%"}
```{r, out.width='70%'}
#| fig-cap: ""
#| fig-alt: "histogram of API scores, approximately normally distributed with slightly heavy tails"
pop %>%
  ggplot(aes(x = api00)) + 
  geom_histogram(fill = "pink", color = "black", binwidth = 25) +
  labs(y = "Frequency", x = "API")
```
:::
:::::

-   Number of SSUs per PSU (schools per district) ranges wildly:

::::: columns
::: {.column width="60%"}
```{r}
pop %>%
  group_by(dnum) %>%
  tally() -> num_psus
round(summary(num_psus$n))
```
:::

::: {.column width="40%"}
```{r}
#| fig-cap: ""
#| fig-alt: "histogram of number of SSUs (schools) per PSU (district), showing an extremely right-skewed distribution"
num_psus %>%
  ggplot(aes(x = n)) + 
  geom_histogram(fill = "lightgreen", color = "black", binwidth = 5) +
  labs(y = "Frequency", x = "Number of Schools per District (SSUs per PSU)")
```
:::
:::::

## Example (con't)

```{r}
samp <- read_dta("./datasets/apipop_clustersample.dta")
n_psu <- length(table(samp$dnum))
n_ssu <- nrow(samp)
```

-   One-stage cluster sample taken of $n=$ `r n_psu` PSUs (districts) -- roughly 5% of total

-   Result is a total of **`r n_ssu` SSUs**

::::: columns
::: {.column width="40%"}
Boxplot of API $(y)$ by PSU (for the **sample**):

```{r}
#| fig-cap: ""
#| fig-alt: "histogram of API scores for the sample, approximately normally distributed with slightly heavy tails"
samp %>%
  ggplot(aes(y = api00, x = factor(dnum))) + 
  geom_boxplot() +
  labs(y = "API", x = "School Number")
```
:::

::: {.column width="60%"}
Number of SSUs per PSU **in the sample** has pretty big range:

```{r}
samp %>%
  group_by(dnum) %>%
  tally() -> num_psus
round(summary(num_psus$n))
```
:::
:::::

-   Notice that the one really big PSU didn't happen to get sampled

## Example: Estimating the Mean API {.pause}

```{r}
#| echo: true
#| classes: .cell-output
des <- svydesign(id = ~dnum, weight = ~wt, fpc = ~popsize, data=samp)
svytotal(~api00, design = des)
```

```{r}
options(scipen = 999)
tot <- svytotal(~api00, design = des)
```

<p>

1.  Estimate of total:

    -   $\hat{t}_{clus} = \text{`r round(coef(tot))`}$

. . .

2.  Estimate of mean:

    -   $\hat{\bar{y}}_{clus}= \hat{t}_{clus}/M_0 = \text{`r round(coef(tot))`}/\text{`r M0`} = \text{`r round(coef(tot)/M0,1)`}$    ($M_0=$ `r M0` districts in pop.)

. . .

3.  Estimate of variance of the mean:

    -   $\widehat{V}(\hat{\bar{y}}_{clus}) = \frac{1}{M_0^2}\widehat{V}(\hat{t}_{clus}) = \frac{1}{\text{`r M0`}^2} (\text{`r round(SE(tot))`}^2) = \text{`r round(SE(tot)^2/M0^2,1)`}$

. . .

4.  Estimated SE of the mean: $\widehat{SE}(\hat{\bar{y}}_{clus}) = \sqrt{\text{`r round(SE(tot)^2/M0^2,1)`}} = \text{`r round(sqrt(SE(tot)^2/M0^2),1)`}$

    -   That is a pretty big standard error! 95% CI = `r round(coef(tot)/M0,1)` $\pm$ `r round(sqrt(SE(tot)^2/M0^2),1)` = (`r round(coef(tot)/M0 - 2*sqrt(SE(tot)^2/M0^2),1)`, `r round(coef(tot)/M0 + 2*sqrt(SE(tot)^2/M0^2),1)`)

## Activity 12.1 (Part 1)

Clusters of Unequal Sizes (Part 1)

## Our Mean Estimate: The Unbiased Estimator {.pause}

-   Our mean estimator is: $\displaystyle \hat{\bar{y}}_{clus}=  \frac{N \bar{t}}{M_0}  = \frac{\hat{t}_{clus}}{M_0}$

-   Note the denominator: **known** total number of SSUs

-   This is called the [**unbiased estimator of the mean**]{.alert2}, since it is unbiased for the true population mean ($\bar{y}_U$)

    -   We have taken an SRS of PSU totals $(t_i)$

    -   Thus by property of SRS, the estimated total is unbiased: $E[\hat{t}_{clus}] = t_U$

    -   Thus $E[\hat{\bar{y}}_{clus}] = E[\hat{t}_{clus}/M_0]=E[\hat{t}_{clus}]/M_0 = t_U/M_0 = \bar{y}_U$

-   The **unbiased estimator** can be written in terms of sampling weights: $\displaystyle \hat{\bar{y}}_{clus}= \frac{\sum_{i\in \mathcal{S}} \sum_{j\in \mathcal{S}_i}  w_{ij} y_{ij}}{M_0}$

    -   Remember that when an SRS of PSUs is taken, $w_{ij} = N/n$ for all sampled units

. . .

-   [But, what if we don't know $M_0$?]{.alert}

    -   For example, if we only have a list of housing units, not people in the units

-   We can definitely still estimate the total -- but can we still estimate the mean?

## Another Way: Ratio Estimation {.pause}

> [**Ratio estimator of the mean:**]{.alert} $$\hat{\bar{y}}_r= \frac{\hat{t}_{clus}}{\widehat{M}_0} = \frac{N \bar{t}}{N \hat{\bar{M}}} = \frac{N \frac{1}{n} \sum_{i\in \mathcal{S}} t_i}{N\frac{1}{n} \sum_{i\in \mathcal{S}} M_i}  = \frac{\sum_{i\in \mathcal{S}} t_i}{\sum_{i\in \mathcal{S}} M_i} = \frac{\sum_{i\in \mathcal{S}} M_i \bar{y}_i}{\sum_{i\in \mathcal{S}} M_i}$$ where $\hat{\bar{M}}= \frac{1}{n} \sum_{i \in \mathcal{S}} M_i =$ average number of SSUs per PSU (based on sample)

-   Note the denominator: [estimated]{.alert} total number of SSUs

. . .

-   The **ratio estimator** can also be written in terms of sampling weights: $\displaystyle \hat{\bar{y}}_r= \frac{\hat{t}_{clus}}{\widehat{M}_0} = \frac{\sum_{i\in \mathcal{S}} \sum_{j\in \mathcal{S}_i}  w_{ij} y_{ij}}{\sum_{i\in \mathcal{S}} \sum_{j\in \mathcal{S}_i}  w_{ij} }$

-   This estimator can be used **even if you don't know the true total number of SSUs,** $M_0$

. . .

-   It is, however, **not unbiased** ($E[\hat{\bar{y}}_r] \ne \bar{y}_U$) and is a **ratio of two random variables**

    -   Both the numerator, $\hat{t}_{clus}$, and denominator, $\widehat{M}_0$, will be different in another sample

## Another Way (con't)

-   The bias will be small, however, if the correlation between $M_i$ and $t_i$ is close to 1\
    *(If cluster sizes and cluster totals are highly correlated)*

-   And, this is usually the case!

-   The [total for each PSU is usually roughly proportional to the size of the PSU]{.alert}

```{r}
#| fig-cap: ""
#| fig-alt: "scatterplot with number of schools on the x-axis and total API on the y-axis, showing essentially a 45 degree line"
#| fig-align: center
samp %>%
  group_by(dnum) %>%
  summarize(num_schools = n(),
            total_api = sum(api00)) -> samp_collapse
samp_collapse %>%
  ggplot(aes(y = total_api, x = num_schools)) +
  geom_point(size = 4, shape = 1) +
  labs(x = "Number of Schools (SSUs)", y = "Total API") + 
  theme(axis.title = element_text(size = 18),
        axis.text = element_text(size = 14))

cor_t_M <- cor(samp_collapse$total_api, samp_collapse$num_schools)
```

<p>

-   $\text{Corr}(t_i,M_i) = \text{`r round(cor_t_M, 3)`}$

-   Also think about the dorm room example -- the more students in a room, the larger the total GPA will be

## Example: Estimating the Mean API with Ratio Estimator {.pause}

```{r}
that <- coef(tot)
w <- N/n_psu
M0hat <- n_ssu*N/n_psu
```

-   Hand-calculating the ratio estimator:

    -   From before, $\hat{t}_{clus} = \text{`r round(that)`}$

    -   Weights for each sampled unit = $\frac{N}{n} = \frac{\text{`r N`}}{\text{`r n_psu`}} = \text{`r round(w, 2)`}$

    -   Sample has `r n_ssu` schools (SSUs), so sum of weights = $\widehat{M}_0 = \text{`r n_ssu`} \times \text{`r round(N/n_psu, 2)`} = \text{`r round(M0hat, 2)`}$

<p>

. . .

-   Ratio estimate of mean API: $\displaystyle \hat{\bar{y}}_r= \frac{\hat{t}_{clus}}{\widehat{M}_0} = \frac{\text{`r round(that)`}}{\text{`r round(M0hat, 2)`}} = \text{`r round(that/M0hat,1)`}$

<p>

. . .

-   Compare to the unbiased estimator, $\displaystyle \hat{\bar{y}}_{clus} = \frac{\hat{t}_{clus}}{M_0} = \frac{\text{`r round(that)`}}{\text{`r M0`}} = \text{`r round(that/M0,1)`}$

. . .

-   For this sample, the ratio estimate of the mean is [larger]{.alert} than the unbiased estimate [for this sample]{.alert} -- why?

    -   *Note: this is just true for this sample -- ratio estimator could be bigger, could be smaller, depends on the specific sample*

## Variance Estimation for the Ratio Estimator {.pause}

-   Why would we want to use a **biased** estimator, even if the bias is small?

. . .

-   Because of potentially HUGE gains in efficiency -- much smaller variance estimate

. . .

<p>

-   But, it is also a **ratio of two random variables** $(\hat{t}_{clus},\widehat{M}_0)$, thus tricky to deal with

-   To calculate its variance we use Taylor series **linearization**[^1]

[^1]: It is important to know this name, **linearization**, but not to understand the technical details

## What is Linearization? {.pause}

-   We want the variance of a nonlinear function of two random variables: $V(\hat{t}_{clus}/\widehat{M}_0)$

-   We cannot "pull apart" this **nonlinear** function with rules that work for sums of random variables, like $V(X+Y) = V(X) + V(Y) + 2\text{Cov}(X,Y)$

-   But, if we could [**approximate the nonlinear function with a linear function**]{.alert} then we **could** use the rules about sums of random variables

<p>

. . .

-   **Linearization** reduces the nonlinear quantity (the ratio, here) to an approximate linear quantity (an expression involving adding and subtracting random variables -- not multiplying or dividing them) by using the linear terms of a Taylor series expansion[^2]

-   Then a variance formula can be constructed using usual rules (and an estimate obtained by plugging in estimates of unknown quantities)

[^2]: Again, math details not important -- just the concept

## Linearization for the Variance of the Ratio Estimator {.pause}

-   For our estimate of the mean, $\hat{\bar{y}}_r$, the first order Taylor series expansion is: $$\hat{\bar{y}}_r= \frac{\hat{t}_{clus}}{\widehat{M}_0} \approx \frac{t_U}{M_0} + \frac{1}{M_0}\left(\hat{t}_{clus}- \frac{t_U}{M_0} \widehat{M}_0 \right)$$

-   The **random variables** are the pieces with **hats**

    -   $\hat{t}_{clus}, \widehat{M}_0$ -- estimates, which are [**random variables**]{.alert2}

    -   $t_U, M_0$ -- fixed (unknown) quantities ([**constants**]{.alert})

. . .

-   So now the rules of variance of a linear combination of random variables can be used: $$V(\hat{\bar{y}}_r) \approx V\left[\frac{t_U}{M_0} + \frac{1}{M_0}\left(\hat{t}_{clus}- \frac{t_U}{M_0} \widehat{M}_0 \right)\right] = \frac{1}{M_0^2}\left[V(\hat{t}_{clus}) + \frac{t_U^2}{M_0^2} V(\widehat{M}_0) - 2\frac{t_U}{M_0} \text{Cov}(\hat{t}_{clus},\widehat{M}_0) \right]$$

. . .

-   To get the estimated variance, plug in $\hat{t}_{clus}$ and $\widehat{M}_0$ (in place of $t_U$ and $M_0$) in variance expression above

## Linearization (con't) {.pause}

-   Looking closer at this variance expression:

$$V(\hat{\bar{y}}_r) \approx \frac{1}{M_0^2}\left[V(\hat{t}_{clus}) + \frac{t_U^2}{M_0^2} V(\widehat{M}_0) - 2\frac{t_U}{M_0} \text{Cov}(\hat{t}_{clus},\widehat{M}_0) \right]$$

-   If the cluster sizes are all the same size:

    -   $\widehat{M}_0 = M_0$ because every sample of $n$ PSUs contains the same number of SSUs, so $V(\widehat{M}_0) = 0$

    -   $\text{Cov}(\hat{t}_{clus},\widehat{M}_0) = 0$ because $\widehat{M}_0$ is a constant, and the covariance of a random variable with a constant is 0

. . .

-   Thus, the linearization variance estimator reduces to: $\displaystyle V(\hat{\bar{y}}_r) = \frac{1}{M_0^2} V(\hat{t}_{clus})$

-   This is the variance of the unbiased estimator!

-   Software (e.g., Stata, R, SAS) uses the linearization estimator even when the clusters are all the same size

-   Linearization is also used for other, complex statistics -- such as regression coefficients

## Variance Estimation for One-Stage Cluster Sampling (Ratio Estimator)  {.pause}

-   By plugging in estimates for the quantities in $V(\hat{\bar{y}}_r)$, for a **one-stage cluster sample** where the **sampling of PSUs is via SRS**, we can [estimate]{.alert} the variance of the (ratio estimator of the) mean with: $$\widehat{V}(\hat{\bar{y}}_r) = \frac{1}{n \hat{\bar{M}}^2} \left(1-\frac{n}{N}\right)\frac{\sum_{i \in \mathcal{S}} (t_i - M_i\hat{\bar{y}}_r)^2}{n-1} = \frac{1}{n \hat{\bar{M}}^2} \left(1-\frac{n}{N}\right)\frac{\sum_{i \in \mathcal{S}} M_i^2 (\bar{y}_i - \hat{\bar{y}}_r)^2}{n-1}$$

-   Notable pieces:

    -   $\hat{\bar{M}}= \frac{1}{n} \sum_{i \in \mathcal{S}} M_i =$ average number of SSUs per PSU (based on sample)

    -   FPC = $\left(1-\frac{n}{N}\right)$ is incorporated **based on number of PSUs sampled**

    -   Piece with the summation looks kind of like a variance of $t_i$

. . .

-   **Key part:** $\sum_{i \in \mathcal{S}} (t_i - M_i\hat{\bar{y}}_r)^2$\
    = how far the PSU totals $(t_i)$ are from the [**estimated total for each specific PSU** $(M_i \hat{\bar{y}}_r)$]{.alert}

    -   $M_i \hat{\bar{y}}_r$ = (size of the PSU) $\times$ (estimated mean) = estimate of the total for that PSU

## Compare Ratio Estimator to Unbiased Estimator {.pause}

-   Recall the estimated variance of the **unbiased estimator**: $$\widehat{V}(\hat{\bar{y}}_{clus}) = \frac{N^2}{M_0^2} \left(1-\frac{n}{N}\right)\frac{s_t^2}{n} = \frac{1}{n \bar{M}^2} \left(1-\frac{n}{N}\right)\frac{\sum_{i \in\mathcal{S}_i} (t_i - \bar{t})^2}{n-1}$$ (plugging in: $s_t^2 = \frac{1}{n-1}\sum_{i \in\mathcal{S}_i} (t_i - \bar{t})^2$; $\bar{M}=M_0/N=$ true mean number of SSUs per PSU)

-   **Key part:** $\sum_{i \in\mathcal{S}_i} (t_i - \bar{t})^2$\
    = how far the PSU totals $(t_i)$ are from the [**estimated mean total across all PSUs** $(\bar{t})$]{.alert}

. . .

-   Variance of the **ratio estimate** of the mean: $\displaystyle \widehat{V}(\hat{\bar{y}}_r) = \frac{1}{n \hat{\bar{M}}^2} \left(1-\frac{n}{N}\right)\frac{\sum_{i \in \mathcal{S}} (t_i - M_i\hat{\bar{y}_r})^2}{n-1}$

. . .

-   Key differences:

    -   Ratio estimator uses an **estimate** of the average number of SSUs per PSU $(\hat{\bar{M}})$ where the unbiased estimator uses the actual, **true** average $(M_0/N)$

    -   Ratio estimator uses $(t_i - M_i\hat{\bar{y}}_r)$ while unbiased estimator uses $(t_i-\bar{t})$

        <!-- -   These differences will be much smaller for ratio estimator since the size of the PSU $(M_i$) is taken into account! -->

## Visualization

```{r}
samp %>%
  mutate(M0hat = sum(wt)) %>%
  group_by(M0hat, dnum) %>%
  summarize(Mi = n(),
            ti = sum(api00)) %>%
  ungroup() %>%
  mutate(tbar = mean(ti),
         that = tbar*757,
         ybarr = that/M0hat,
         Miybarr = Mi*ybarr) -> dat
dat$id <- 1:nrow(dat)
```

::::::: columns
:::: {.column width="50%"}
```{r, fig.asp=0.7}
#| fig-cap: ""
#| fig-alt: "plot with PSU number on the x-axis and total on the y-axis, with a line from each dot (total) to the overall average total, which is shown as a horizontal line of triangles. Some of the distances are very large, some are small."
dat %>%
  ggplot(aes(x=id, y=ti)) + 
  geom_errorbar(aes(ymin=ti, ymax=tbar), width=0, col="darkgrey") + 
  theme(panel.grid=element_blank()) + 
  geom_point(aes(x=id, y=ti), size = 3) + 
  geom_point(aes(x=id, y=tbar), col="red", shape=2, size = 3) + 
  ylim(0,25000) + 
  labs(x="PSU", y="Total") + 
  theme(axis.title = element_text(size=18),
        title = element_text(size=20)) +
  ggtitle("Unbiased Estimator")
```

::: smaller-text
Distance from PSU total $t_i$ to **overall average total**, $\bar{t}$
:::
::::

:::: {.column width="50%"}
```{r, fig.asp=0.7}
#| fig-cap: ""
#| fig-alt: "plot with PSU number on the x-axis and total on the y-axis, with a line from each dot (total) to the estimated total for each PSU, which is shown as a triangle for each PSU. Most of the distances are small."
dat %>%
  ggplot(aes(x=id, y=ti)) + 
  geom_errorbar(aes(ymin=ti, ymax=Miybarr), width=0, col="darkgrey") + 
  theme(panel.grid=element_blank()) + 
  geom_point(aes(x=id, y=ti), size = 3) + 
  geom_point(aes(x=id, y=Miybarr), col="red", shape=2, size = 3) + 
  ylim(0,25000) + 
  labs(x="PSU", y="Total") +
  theme(axis.title = element_text(size=18),
        title = element_text(size=20)) +
  ggtitle("Ratio Estimator")
```

::: smaller-text
Distance from PSU total $t_i$ to **estimated PSU-specific total**, $M_i \hat{\bar{y}}_r$
:::
::::
:::::::

-   Think about squaring those distances and summing them... HUGE difference!

-   Since the PSU totals are usually correlated with the PSU sizes (no matter what $y$ is!), in general the ratio estimator will have **smaller variance**!

## Example: Ratio Estimator Variance for Mean API

-   Ratio estimator for the sample of schools:

```{r}
#| echo: true
#| classes: .cell-output
svymean(~api00, design = des)
```

<p>

```{r}
#| echo: true
#| classes: .cell-output
confint(svymean(~api00, design = des))
```

-   Compare to the unbiased estimator (previously calculated):

    -   $\hat{\bar{y}}_{clus}$ = `r round(coef(tot)/M0,1)`

    -   $\widehat{SE}(\hat{\bar{y}}_{clus})$ = `r round(sqrt(SE(tot)^2/M0^2),1)`

    -   95% CI = (`r round(coef(tot)/M0 - 2*sqrt(SE(tot)^2/M0^2),1)`, `r round(coef(tot)/M0 + 2*sqrt(SE(tot)^2/M0^2),1)`)

-   **HUGE** gain in efficiency with ratio estimator! SE almost 4 times larger for $\hat{\bar{y}}_{clus}$!

## Activity 12.1 (Part 2)

Clusters of Unequal Sizes (Part 2)
