---
title: "Complex Designs"
subtitle: "PUBHBIO 7225 Lecture 16"
# author: "Generative AI acknowledgment: MS Copilot was used to generate alt text for some images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
editor: 
  markdown: 
    wrap: 72
---

## Outline

::::: columns
::: {.column width="60%"}
**Topics**

-   Domain Estimation

-   Chi-Square Tests

-   Discussion of Individual Project
:::

::: {.column width="40%"}
**Activities**

-   16.1 Domain Estimation/Comparison
:::
:::::

**Readings**

-   West BT, Berglund P, Heeringa SG (2008). A closer examination of
    subpopulation analysis of complex-sample survey data. *The Stata
    Journal*, 8(4); 520-531. (PDF on Carmen)

**Assignments**

-   Problem Set 4 due Thursday 10/23/2025 11:59pm via Carmen

## Estimation in Domains

> [**Domain**]{.alert3} = a subgroup in the population that is not part
> of the sampling design.\
> Also called [subpopulations]{.alert3} or [subgroups]{.alert3} or
> [subclasses]{.alert3}.

-   Not a stratum, not a cluster/group of clusters

    -   Domains can cross strata (and often/usually do)

    -   Domains can include units from many clusters (but maybe not all
        clusters)

-   *Like* strata and clusters, each unit $i$ either belongs to a
    certain domain or does not belong to the domain

-   *Unlike* strata and clusters, we don't know what domain unit $i$
    belongs to until it is sampled (i.e., record domain membership on
    the survey)

-   Examples:

    -   SRS of OSU students on Columbus campus, domain = students who
        live off-campus

    -   Stratified (by county) random sample of women who recently gave
        birth in Ohio, domain = women who did not have a prenatal visit

    -   Two-stage cluster sample of students within schools, domain =
        students with a disclosed disability

## Domain Sizes are Random Variables {.pause}

-   **Key Point:** We don't know which domain a unit belongs to until it
    is sampled

    -   Thus, the number of units in the domain is a [random
        variable]{.alert2}

    -   Take another sample, get a different number of units in the
        domain

        -   Contrast with stratum -- take another sample, get same
            number of units in the stratum!

    -   This extra variability **must be accounted for in analysis
        stage** to get correct variance estimates

. . .

-   Variance estimates must properly account for the survey design

-   You should **never subset your data for analysis**, even if there
    are no units in a particular domain

. . .

-   Subsetting will definitely cause problems when:

    -   A stratum does not contain any members of the domain

        -   Usually unlikely to occur -- but possible in highly
            stratified designs

    -   A PSU does not contain any members of a domain

        -   Quite possible -- imagine sampling households, and the
            domain is males aged 50+

. . .

-   **One exception:** It's okay to subset an SRS -- an SRS of a
    population contains SRS of a subpopulation

## Domain Indicators

-   Indicator variable, $x_i$, for domain membership:
    $$x_i = \begin{cases} 1 & \text{unit $i$ is in domain $d$}\\ 
    0 & \text{unit $i$ not in domain $d$} 
    \end{cases}$$

-   Estimates:

    -   Domain total:  
        $\displaystyle \hat{t}^{(d)} = \sum_{i \in \mathcal{S}} w_i x_i y_i$

    -   Domain mean:  
        $\displaystyle \bar{y}^{(d)} = \frac{\sum_{i \in \mathcal{S}} w_i x_i y_i}{\sum_{i \in \mathcal{S}} w_i x_i}$

<p>

-   Since $x_i=0$ for units not in the domain, you'd get the **same
    point estimate** if you simply dropped the non-domain units

-   But -- variance estimation would be incorrect

-   **You must use the whole sample, even to obtain estimates for a
    subpopulation**

## Example: Domain Estimation in NHAMCS {.pause}

-   Analysis of data from the 2004 National Hospital Ambulatory Medical
    Care Survey (paper on Carmen)

-   NHAMCS collects data on the use and provision of ambulatory care
    services in hospital emergency departments (ED) and outpatient
    departments and ambulatory surgery locations

-   4-stage design with stratification:

    1.  PSUs = geographic areas (selection via PPS) (stratified by SES
        variables)

    2.  hospitals within PSUs (selection via PPS)

    3.  clinics within hospitals (complex selection procedure involving
        PPS)

    4.  patient visits within clinics (systematic sampling)

-   2004 sample had [8 strata]{.alert} and [294 PSUs]{.alert} (between 6
    and 86 PSUs per stratum)

. . .

-   **Domain of interest:** visits to EDs by elderly (age $\geq$ 60)
    African-American males

    -   Domain members could theoretically appear across the strata and
        PSUs -- but there are likely PSUs with no domain members

-   **Goal:** estimate the percentage of visits in this domain with
    dizziness or vertigo as a reason for the visit

## Example (con't)

-   Two analysis approaches taken:

    1.  **Conditional Approach**: Subsetting the data to domain members
        only and performing estimation ([**bad!**]{.alert})

    2.  **Unconditional Approach:** Appropriately including all sampled
        units in calculations ([**good!**]{.alert2})

::: smaller-text
|   | Unconditional Approach (Correct) | Conditional Approach (Incorrect) | When subsetting, we have... |
|------------------|------------------|------------------|------------------|
| Sample Size | 68,372 | 397 |  |
| Subpopulation Sample Size | 397 | 397 |  |
| Design Strata | 8 | 8 | Correct \# strata |
| Design Clusters | 294 | 114 | Too few PSUs |
| Design DF (# PSUs − \# Strata) | 286 | 106 | DF too small |
| Estimated Percentage | 4.8201 | 4.8201 | Correct point estimate |
| Standard Error | 1.5904 | 1.5761 | SE too small |
| 95% CI | (1.6897, 7.9504) | (1.6954, 7.9448) | CI too narrow |

Adapted from Table 1 in West et al. (2008)
:::

## Domain Estimation in Software

::: smaller-text
**Stata:**

```{r}
#| echo: true
#| eval: false

# estimates just for the domain of interest -- units with DOMAIN not equal to 0
svy, subpop(DOMAIN): mean y
svy, subpop(DOMAIN): proportion y

# estimates in all domains defined by the unique values of DOMAIN
svy: mean y, over(DOMAIN)
svy: proportion y, over(DOMAIN)
```

-   **Stata user warning:** do not use **if** with Stata for domain
    estimates!

-   This will subset the data (even if you used the **svyset** command
    on the full dataset)

**R:**

```{r}
#| echo: true
#| eval: false

# estimates just for the domain of interest -- units with DOMAIN = 1/TRUE
svymean(~y, subset(DESIGN_OBJECT, DOMAIN==1))

# estimates in all domains defined by the unique values of DOMAIN
svyby(~y, ~DOMAIN, DESIGN_OBJECT, svymean)
```
:::

## Activity 16.1 (Part 1)

Domain Estimation and Comparison (Part 1)

## Comparing Domains

-   So far in this course we have done a lot of estimating of means and
    totals and proportions, either for the whole population or (today)
    for domains

-   We have yet to do any formal comparisons -- i.e., any hypothesis
    testing

-   If we want to compare two (or more) domains, often we will be using
    **chi-square tests** -- since so much of survey data is
    **categorical**

<br>

-   I am going to assume you are [familiar with (and comfortable with)
    chi-square tests in the context of infinite populations]{.alert2}
    (i.e., what you've seen before this class), but stop me if you have
    questions!

-   We will start with an example that illustrates why taking into
    account the survey design is important when doing tests like a
    chi-square test

## Example: Households With and Without Netflix

-   Suppose we have an SRS of $n$ = 500 co-habiting couples from a very
    large population (i.e., $N$ large enough to be able to ignore the
    FPC)

-   We want to compare two domains: [**People who have
    Netflix**]{.alert} and [**People who do not have Netflix**]{.alert2}

-   [Question: Are people who have Netflix less likely to subscribe to
    cable TV?]{.alert3} (We recognize that no causal link can be made
    with a cross-sectional survey!)

::::: columns
::: {.column width="50%"}
-   Data from our SRS of 500 co-habiting couples:
:::

::: {.column width="50%"}
|         Has | Cable TV | No Cable TV | Total |
|------------:|:--------:|:-----------:|:-----:|
| Has Netflix |   119    |     88      |  207  |
|  No Netflix |   188    |     105     |  293  |
|       Total |          |             |  500  |
:::
:::::

-   Resulting estimates:

    -   119/207 = 57.5% of couples with Netflix have cable TV

    -   188/293 = 64.2% of couples without Netflix have cable TV

## Example (con't)

-   A chi-square test gives us:

```{r}
library(tidyverse)
dat <- tibble(id = 1:500) %>%
  mutate(netflix = case_when(id > 207 ~ "No Netflix", .default = "Has Netflix"),
         cable = case_when(id > 119 & id <= 207 ~ "No Cable TV",
                           id > 395 ~ "No Cable TV",
                           .default = "Cable TV"))
dat_larger <- bind_rows(dat, dat)
```

```{r}
#| echo: true
chisq.test(dat$netflix, dat$cable, correct = FALSE)
```

-   Conclusion: There is **no evidence of an association** between
    having Netflix and having cable (i.e., not true that the proportion
    of people with cable differs for people with/without Netflix)

-   Summarize observed association with an **odds ratio (OR)** or
    **prevalence ratio (PR)**:

<br>

<br>

*Surveys are (usually) cross-sectional → prefer PR to RR (risk ratio)*

## Example (con't)

-   Now, suppose we asked *both partners* about Netflix and cable

    -   We'll have *twice as many people* (1000 responses, instead of
        500), with the same proportions (assuming both partners answered
        the same), but a different test result:

```{r}
dat_larger %>%
  janitor::tabyl(netflix, cable) %>%
  janitor::adorn_percentages("row") %>%
  janitor::adorn_pct_formatting(digits = 2) %>%
  janitor::adorn_ns()
```

```{r}
#| echo: true
chisq.test(dat_larger$netflix, dat_larger$cable, correct = FALSE)
```

-   Ignoring the clustering has **inflated the test statistic** value
    and **lowered the p-value**

-   ICC for this design is very high, in fact it is:

-   What is the DEFF for this design?

## Chi-Square Tests for Complex Survey Data

<!-- -   This is an extreme example (and in fact you could make it more -->

<!--     extreme by interviewing any children in the house too!), but it -->

<!--     illustrates the problem -->

-   [**Ignoring clustering can inflate test statistics and deflate
    p-values**]{.alert}

    -   Remember, DEFF \> 1 for many outcomes in a cluster sample

    -   A cluster sample of size $n$ provides the same precision as an
        SRS of size $n$/DEFF

        -   Effective sample size is *smaller* than $n$

    -   Thus you are **overstating your signficance** (p-value too
        small)

    -   Note that the $X^2$ test statistic does not have a
        $\chi^2_{(r-1)(c-1)}$ distribution under the null hypothesis
        when a cluster sample is taken and ICC \> 0

-   [**Ignoring stratification can deflate test statistics and inflate
    p-values**]{.alert2}

    -   Remember, DEFF \< 1 for many outcomes in a stratified sample

    -   A stratified sample of size $n$ provides the same precision as
        an SRS of size $n$/DEFF

        -   Effective sample size is *larger* than $n$

    -   Thus you are **being too conservative** (p-value too high)

-   The worse crime (usually) is ignoring clustering and "overstating"
    results (inflate Type 1 error)

## Corrections to Chi-Square Tests for Complex Survey Data {.pause}

-   For an [SRS in an infinite population]{.alert}:
    -   $X^2$ from a chi-square test has a $\chi^2_{(r-1)(c-1)}$
        distribution under $H_0$
    -   Thus, $E[X^2] = (r-1)(c-1)$
        -   (the expected value of a chi-square random variable is equal
            to its DF)
-   For [complex survey designs]{.alert2}:
    -   $X^2$ from a chi-square test **does not** follow a
        $\chi^2_{(r-1)(c-1)}$ distribution under $H_0$

. . .

<p>

-   However, we can adjust the test statistic $X^2$ in order to make a
    *rescaled* test statistic have an approximately $\chi^2$
    distribution

-   Two methods used most often (not very creatively named):

    -   **First Order Correction** (also called Rao-Scott first order
        correction)

    -   **Second Order Correction** (also called Rao-Scott second order
        correction)

## Option 1: (Rao-Scott) First Order Correction {.pause}

-   **Idea:** Match the [**mean**]{.alert} of the test statistic $X^2$
    (under the complex design) to the [**mean**]{.alert} of a
    $\chi^2_{(r-1)(c-1)}$ distribution, thus "rescaling" the test
    statistic

-   **"First Order"** because we're rescaling based on the mean

    1.  Calculate $E[X^2]$ = the expected value of $X^2$ for the complex
        design under $H_0$

    2.  Calculate first order corrected test statistic as  
        $X^2_F = \frac{(r-1)(c-1) X^2}{E[X^2]}$

    3.  Compare $X^2_F$ to a $\chi^2_{(r-1)(c-1)}$ distribution

        -   In practice, works better to compare $X^2_F/(r-1)(c-1)$ to
            an $F$ distribution

. . .

-   Calculation of $E[X^2]$ depends on the cell and marginal
    probabilities and the DEFF for estimating each of these
    probabilities (there's the survey design part!)

. . .

-   [Slight problem:]{.alert} we are matching the mean of the
    distribution -- but p-values often come from the tail of the
    distribution (at least the p-values we care about most!)

    -   Variance of $X^2_F$ is actually larger than variance of
        $\chi^2_{(r-1)(c-1)} \rightarrow$ p-values from $X^2_F$ are
        slightly smaller than they should be

## Option 2: (Rao-Scott) Second Order Correction

-   **Idea:** Match the [**mean and variance**]{.alert2} of the test
    statistic $X^2$ (under the complex design) to the [**mean and
    variance**]{.alert2} of a $\chi^2_{(r-1)(c-1)}$ distribution, thus
    "rescaling" the test statistic

-   **Second Order** because we're rescaling based on the mean and
    variance

    1.  Calculate $E[X^2]$ = the expected value of $X^2$ for the complex
        design under $H_0$

    2.  Calculate $V[X^2]$ = the variance of $X^2$ under $H_0$

    3.  Calculate second order corrected test statistic as  
        $X^2_S = \frac{\nu X^2_F}{(r-1)(c-1)}$ where
        $\nu = 2\frac{(E[X^2])^2}{V[X^2]}$

    4.  Compare $X^2_F$ to a $\chi^2_{\nu}$ distribution

        -   In practice, works better to compare $X^2_S/\nu$ to an $F$
            distribution

    -   Similar to the Satterthwaite approximation for ANOVA

. . .

-   Calculation of $V[X^2]$ even more involved than $E[X^2]$, requires
    covariance matrix of estimated cell proportions

-   For many software packages (Stata, R), the second order correction
    is the default

## Activity 16.1 (Part 2)

Domain Estimation and Comparison (Part 2)
