---
title: "Sampling Distributions"
subtitle: "PUBHBIO 7225 Lecture 3"
author: "Generative AI acknowledgment: MS Copilot was used to generate alt text for images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
---

## Outline

::::: columns
::: {.column width="50%"}
**Topics**

-   Sampling Distributions

    -   Creating sampling distributions

    -   Calculating expectation and variance
:::

::: {.column width="50%"}
**Activities**

-   3.1 Sampling Cats
:::
:::::

**Readings**

-   Probability Review Handout (on Carmen)

**Assignments**

-   Problem Set 1 due Thursday 9/4/25 11:59pm via Carmen
-   Peer Evaluation of Problem Set 1 due Tuesday 9/9/25 11:59pm via Carmen

## Sampling Distributions

> **Sampling distribution of a statistic** = the distribution of different values of a statistic obtained from all possible samples from the population

-   A sampling distribution is a **discrete probability distribution**

-   For a given sampling design, each statistic (mean, total, proportion, etc.) has a sampling distribution

    -   Sampling distribution = described by a **probability mass function (PMF)**

-   In your previous statistics classes, you usually assumed a sample of size $n$ from an *infinite* population -- hence sampling distributions such as $\bar{x}\sim N(\mu, \sigma^2/n)$

-   With a **finite** population, the sampling distribution takes a different form

-   **Notation**

    -   $N$ = size of the finite population $(N < \infty)$

    -   $n$ = sample size

    -   $\hat{\theta}$ = statistic for which you want the sampling distribution (e.g., $\hat{\theta}=\bar{y}$)

## Example of a Sampling Distribution

-   Suppose there are $N=5$ clinics in a town and you want to estimate the [average number of patients seen in a day (in the town)]{.alert}

-   The **population** of clinics is:

::: smaller-text
|                                        |     |     |     |     |     |
|----------------------------------------|:---:|:---:|:---:|:---:|:---:|
| Clinic ID:                             |  1  |  2  |  3  |  4  |  5  |
| Number of Doctors:                     |  2  |  3  |  4  |  5  | 10  |
| Number of Patients Seen in a Day ($y$) | 10  | 20  | 50  | 60  | 150 |
:::

<p>

-   True mean \# of patients = $(10+20+50+60+150)/5 = 58$

-   [Number of doctors]{.alert2}: we know in advance of the survey (i.e., its on the sampling frame)

-   [Number of patients seen per day]{.alert}: we **don't know** until we take a sample (and then we will only know it for the sampled units)

-   Suppose we only have enough money to sample $n=2$ of the $N=5$ clinics

-   Statistic for which we want the sampling distribution: $\hat{\theta}= \bar{y}$ = mean \# patients

## Example (con't)

-   There are lots of ways we could select $n=2$ clinics from the population of $N=5$. Consider two options:

:::::: smaller-text
::::: columns
::: {.column width="30%"}
| Clinics | Probability |
|:-------:|:-----------:|
|  1, 2   |     0.1     |
|  1, 3   |     0.1     |
|  1, 4   |     0.1     |
|  1, 5   |     0.1     |
|  2, 3   |     0.1     |
|  2, 4   |     0.1     |
|  2, 5   |     0.1     |
|  3, 4   |     0.1     |
|  3, 5   |     0.1     |
|  4, 5   |     0.1     |
:::

::: {.column width="70%"}
| Clinics | Probability | Clinics | Probability | Clinics | Probability |
|:-------:|:-----------:|:-------:|:-----------:|:-------:|:-----------:|
|  1, 1   |    0.04     |  3, 1   |    0.04     |  5, 1   |    0.04     |
|  1, 2   |    0.04     |  3, 2   |    0.04     |  5, 2   |    0.04     |
|  1, 3   |    0.04     |  3, 3   |    0.04     |  5, 3   |    0.04     |
|  1, 4   |    0.04     |  3, 4   |    0.04     |  5, 4   |    0.04     |
|  1, 5   |    0.04     |  3, 5   |    0.04     |  5, 5   |    0.04     |
|  2, 1   |    0.04     |  4, 1   |    0.04     |         |             |
|  2, 2   |    0.04     |  4, 2   |    0.04     |         |             |
|  2, 3   |    0.04     |  4, 3   |    0.04     |         |             |
|  2, 4   |    0.04     |  4, 4   |    0.04     |         |             |
|  2, 5   |    0.04     |  4, 5   |    0.04     |         |             |
:::
:::::
::::::

-   For both schemes (designs), the probability of selection is equal for every pair of units

-   Can you name these two sampling schemes?

## Number of Possible Samples in an SRS or SRSWR

-   Without actually writing out all the possible samples, we can use math tools to calculate the number of possible samples for each of these designs (SRSWR, SRS)

> [**With replacement:**]{.alert} (Exponentiation)
>
> <p>
>
> Total number of outcomes when repeating an experiment with $N$ outcomes $n$ times is $N^n$

> [**Without replacement:**]{.alert2} (Combination)
>
> <p>
>
> Total number of different ways to pick $n$ subjects out of $N$ subjects without regard to order is $\displaystyle{_NC_n} = \binom{N}{n} = \text{``$N$ choose $n$"} = \frac{N!}{n!(N-n)!}$

<p>

-   Thus for sampling $n=2$ clinics from population of $N=5$ clinics:

    -   SRSWR: \# possible samples = $N^n = 5^2 = 25$

    -   SRS: \# possible samples = $\binom{N}{n} = \binom{5}{2} = \frac{5!}{2! (5-2)!} = \frac{120}{2 \times 6} = \frac{120}{12} = 10$

## Selection Probabilities

> [**Selection Probability**]{.alert3} = the probability of a unit being sampled (ending up in our sample) under a specific sampling design

-   **Selection probabilities are a crucial/critical part of probability sampling!**

-   We need them to draw a random sample under a given design, and we also need them to obtain unbiased estimates for population quantities using a single sample

-   Let's calculate the selection probabilities for each of our designs, SRSWR and SRS

<br>

-   Note: for both SRS and SRSWR, selection probabilities are the same for every unit in the population. But this is not necessarily true for all sampling designs (and I'll show you an example where it's not true later in this lecture).

## Selection Probabilities: SRSWR

-   Sampling design: [**SRSWR**]{.alert} of size $n=2$ from a population of size $N=5$.

-   Define:

    -   A = Unit (clinic) 1 is randomly selected on the first draw

    -   B = Unit (clinic) 1 is randomly selected on the second draw

-   Since this is [SRSWR]{.alert}, each time we pick a clinic we "put it back"

-   Thus $P(A) = P(B) = 1/N = 1/5 = 0.2$

-   Selection probability = [Probability that unit (clinic) 1 is selected into the sample]{.alert2} $$\begin{aligned}
    P(\text{Unit 1 in sample}) &{}= P(\text{selected on 1st draw OR selected on 2nd draw})\\
    &{}= P(A \cup B)  \quad\qquad\qquad\qquad\qquad\qquad \text{(union)}\\
    &{}= P(A) + P(B) - P(A \cap B) \qquad\qquad \text{(addition rule)} \\
    &{}= P(A) + P(B) -  P(A) \times P(B) \qquad \text{(A and B are independent)} \\
    &{}= 0.2 + 0.2 - 0.2 \times 0.2 = \mathbf{0.36}
    \end{aligned}$$

-   For a review of these probability rules, see review sheet on Carmen.

## Selection Probabilities: SRS {.pause}

-   Now suppose we are taking a [simple random sample WITHOUT replacement (SRS)]{.alert2} of size $n=2$ from a population of size $N=5$.

    -   A = Unit (clinic) 1 is randomly selected on the first draw

    -   B = Unit (clinic) 1 is randomly selected on the second draw

-   As before, $P(A) = 1/5 = 0.2$ but $P(B) \ne P(A)$ !

    -   [Why?]{.alert2}

. . .

<p>

-   [The probability of selection on 2nd draw depends on whether or not the unit was selected on 1st draw]{.alert2}

. . .

-   We can, however, use the **Law of Total Probability** to calculate $P(B)$: $$\begin{aligned}
    P(B) &{}= P(B|A) \times P(A) + P(B|A^C) \times P(A^C) \\
    &{} = P(\text{chosen on 2nd }|\text{ chosen on 1st}) \times P(\text{chosen on 1st}) \\
    &{} \qquad + P(\text{chosen on 2nd }|\text{ NOT chosen on 1st}) \times P(\text{NOT chosen on 1st})
    \end{aligned}$$

-   These are all quantities we know or can calculate!

## Selection Probabilities: SRS (con't) {.pause}

-   Each of the pieces from the Law of Total Probability:

    -   $P(A) = P(\text{chosen on 1st}) = 0.2$

    -   $P(A^C) = P(\text{NOT chosen on 1st}) = 1 - P(A) = 0.8$     [*(complement rule)*]{.alert}

    -   $P(B|A) = P(\text{chosen on 2nd }|\text{ chosen on 1st}) = 0$     [*(sampling without replacement!)*]{.alert}

    -   $P(B|A^C) = P(\text{chosen on 2nd }|\text{ NOT chosen on 1st}) = 1/4$

. . .

-   Putting these pieces all together gives us $P(B) = P(\text{Unit 1 selected on 2nd draw})$
    -   $P(B) = P(B|A) \times P(A) + P(B|A^C) \times P(A^C) = 0 \times 0.2 + 1/4 \times 0.8 = 0.2$

. . .

-   We want to know the [probability that unit (clinic) 1 is selected into the sample]{.alert2}: $$\begin{aligned}
    P(\text{Unit 1 in sample}) &{}= P(\text{selected on 1st draw OR selected on 2nd draw}) \\
    &{}= P(A \cup B)  \quad\qquad\qquad\qquad\qquad\qquad \text{(union)}\\
    &{}= P(A) + P(B) - \underbrace{P(A \cap B)}_{=0 \text{(why??)}} \qquad\qquad \text{(addition rule)} \\
    &{}= 0.2 + 0.2 - 0 = \mathbf{0.4}
    \end{aligned}$$

## Activity 1.1 (Part 1)

Sampling Cats (Part 1)

## Sampling Distribution for the Clinics SRS

-   [SRS]{.alert2} of size $n=2$ from a population of size $N=5$

<p>

::::::: columns
:::: {.column width="50%"}
Population values:

::: smaller-text
|                            |     |     |     |     |     |
|---------------------------:|:---:|:---:|:---:|:---:|:---:|
|                    Clinic: |  1  |  2  |  3  |  4  |  5  |
| \# Patients per day ($y$): | 10  | 20  | 50  | 60  | 150 |
:::
::::

:::: {.column width="50%"}
Estimate for each sample:

::: smaller-text
| Clinics Sampled | Probability | $y$ values | $\hat{\theta}=\bar{y}$ |
|:---------------:|:-----------:|:----------:|:----------------------:|
|      1, 2       |    1/10     |   10, 20   |           15           |
|      1, 3       |    1/10     |   10, 50   |           30           |
|      1, 4       |    1/10     |   10, 60   |           35           |
|      1, 5       |    1/10     |  10, 150   |           80           |
|      2, 3       |    1/10     |   20, 50   |           35           |
|      2, 4       |    1/10     |   20, 60   |           40           |
|      2, 5       |    1/10     |  20, 150   |           85           |
|      3, 4       |    1/10     |   50, 60   |           55           |
|      3, 5       |    1/10     |  50, 150   |          100           |
|      4, 5       |    1/10     |  60, 150   |          105           |
:::

This is the [**sampling distribution**]{.alert} of $\hat{\theta}$ (i.e., $\bar{y}$)!
::::
:::::::

## Sampling Distribution for the Clinics SRS

-   Notice that one value of $\hat{\theta}$ (35) appears twice -- so the sampling distribution is:

:::::: columns
:::: {.column width="40%"}
-   PMF:

::: smaller-text
| $\hat{\theta}=\bar{y}$ | Probability |
|:----------------------:|:-----------:|
|           15           |     0.1     |
|           30           |     0.1     |
|           35           |     0.2     |
|           40           |     0.1     |
|           55           |     0.1     |
|           80           |     0.1     |
|           85           |     0.1     |
|          100           |     0.1     |
|          105           |     0.1     |
:::
::::

:::: {.column width="60%"}
-   Graph:

```{r}
#| echo: false
#| fig-alt: "Bar graph showing the probability distribution of y-bar corresponding to the table, with the x-axis labeled y-bar and the y-axis labeled 'Probability'."

library(tidyverse)
theme_set(theme_bw())
tibble(ybar = c(15,30,35,40,55,80,85,100,105),
       prob = c(0.1,0.1,0.2,0.1,0.1,0.1,0.1,0.1,0.1)) %>%
  ggplot(aes(y = prob, x = ybar)) + geom_bar(stat = "identity") + theme_bw() +
  labs(y = "Probability", x = expression(bar(y))) +
  theme(axis.title = element_text(size = 24),
        axis.text = element_text(size = 16)) +
  scale_x_continuous(breaks = seq(15,105,5))
```
::::
::::::

## Expectation

> **Expectation** = $\displaystyle E(\bar{y}) = \sum_k k P(\bar{y}=k)$\
> *multiply each possible value (*$k$) of the variable by its probability and sum

:::::: {columns}
:::: {.column width = "70%"}

-   For the clinics SRS:

$$\begin{align}
E(\bar{y}) &= \sum_k k P(\bar{y}=k) \\
&= 15 \times 0.1 + 30 \times 0.1 + 35 \times 0.2 + \cdots + 105 \times 0.1\\
&= \mathbf{58}= \text{true population mean}
\end{align}$$

-   Under the SRS sampling scheme, $\bar{y}$ is an [unbiased estimate]{.alert} of mean of $y$\

-   Though, notice that [no sample actually produces the true mean ($\theta=58$)!]{.alert}
::::

:::: {.column width = "30%"}
::: {.smaller-text}

| $\hat{\theta}=\bar{y}$ | Probability |
|:----------------------:|:-----------:|
|           15           |     0.1     |
|           30           |     0.1     |
|           35           |     0.2     |
|           40           |     0.1     |
|           55           |     0.1     |
|           80           |     0.1     |
|           85           |     0.1     |
|          100           |     0.1     |
|          105           |     0.1     |

::: 
:::: 
::::::

## Variance

> **Variance** = $\displaystyle V(\bar{y}) = \sum_k (k-E(\bar{y}))^2 P(\bar{y}=k)$\
> *take each possible value (*$k$*), subtract the expected value and square the difference, multiply each by its probability, and sum*

-   For the clinics SRS:

$$\begin{align}
V(\bar{y}) &= \sum_k (k-E(\bar{y}))^2 P(\bar{y}=k) \\
& = (15-58)^2 \times 0.1 + (30-58)^2 \times 0.1 + \cdots + (105-58)^2 \times 0.1 = \mathbf{921}
\end{align}$$

## Design-Based Inference

-   Note that we **did not make any assumption about the distribution of the** $y$ **values**

-   The "randomness" is from the sampling

    -   A different sample will produce a different value of $\bar{y}$.

-   This is called [**design-based inference**]{.alert2}.

-   Simulated 10,000 draws of $n=2$ clinics from the population (without replacement)

```{r}
#| fig-alt: "Bar graph showing the simulated probability distribution of y-bar corresponding to 10,000 replicate draws, with the x-axis labeled y-bar and the y-axis labeled 'Probability'."
#| fig-align: center
pop <- tibble(clinic = 1:5,
              patients = c(10, 20, 50, 60, 150))
set.seed(7489243)
pop %>%
  infer::rep_slice_sample(n=2, reps=10000) %>%
  group_by(replicate) %>%
  summarize(ybar = mean(patients)) %>%
  group_by(ybar) %>%
  tally() %>%
  mutate(prob = n/10000) -> temp
temp %>%
  ggplot(aes(y = prob, x = ybar)) + geom_bar(stat = "identity") + 
  geom_bar(stat = "identity") + 
  labs(y = "Fraction of Replicates", x = expression(bar(y))) +
  theme(axis.title = element_text(size = 24),
        axis.text = element_text(size = 16)) +
  scale_x_continuous(breaks = seq(15,105,5))
```

## An Alternative Sampling Distribution

::::: columns
::: {.column width="60%"}
-   Consider the following alternative sampling scheme:
:::

::: {.column width="40%"}
| Clinics | Probability |
|:-------:|:-----------:|
|  1, 5   |    0.25     |
|  2, 5   |    0.25     |
|  3, 5   |    0.25     |
|  4, 5   |    0.25     |
:::
:::::

-   In this scheme, clinic 5 is **always** sampled.

    -   Clinic 5 is called a **certainty unit** → $P(\text{clinic 5 is sampled}) = 1$

-   Because of this, [not all units are selected with equal probability]{.alert}

-   Thus the simple unweighted mean $(\bar{y})$ in each sample is **not** a good estimate of the population mean! (Why?)

<p>

-   Can you come up with a way to estimate the mean?\
    (hint: each clinic 1-4 "represents" 4 clinics, and the 5th is itself...)

## An Alternative Sampling Distribution (con't)

-   A reasonable estimator for the population mean under this sampling scheme is:

    -   $\hat{\theta}_i = (4 y_i + y_5)/5$ for $i = 1, 2, 3, 4$

-   Whichever clinic is sampled (in addition to clinic 5) "represents" all 4 other units

:::: smaller-text
::: center
| Clinics Sampled | Probability | $y$ values |    $\hat{\theta}$    |
|:---------------:|:-----------:|:----------:|:--------------------:|
|      1, 5       |    0.25     |  10, 150   | (4\*10 + 150)/5 = 38 |
|      2, 5       |    0.25     |  20, 150   | (4\*20 + 150)/5 = 46 |
|      3, 5       |    0.25     |  50, 150   | (4\*50 + 150)/5 = 70 |
|      4, 5       |    0.25     |  60, 150   | (4\*60 + 150)/5 = 78 |
:::
::::

-   $E(\hat{\theta}) = \sum_k k P(\hat{\theta}=k) = 38 \times 0.25 + 46 \times 0.25 + 70 \times 0.25 + 78 \times 0.25 = \mathbf{58}$

    -   True population mean → unbiased estimator

-   $V(\hat{\theta}) = \sum_k (k-E(\hat{\theta}))^2 P(\hat{\theta}=k) = (38-58)^2 \times 0.25 + \cdots + (78-58)^2 \times 0.25 = \mathbf{272}$

    -   A whole lot smaller than under SRS (variance for SRS was 921)!

## Why is Variance Reduced?

-   The number of patients seen is positively correlated with the number of doctors. (Makes sense!)

```{r}
#| echo: false
#| fig-cap: ""
#| fig-alt: "Scatterplot showing number of doctors on x-axis and number of clinics on y-axis. The relationship is positive."
#| fig-align: center

tibble(doctors = c(2,3,4,5,10),
       patients = c(10,20,50,60,150)) %>%
  ggplot(aes(y = patients, x = doctors)) + geom_point(size=3) + theme_bw() +
  labs(y = "Number of Patients per Day", x = "Number of Doctors") +
  theme(axis.title = element_text(size = 24),
        axis.text = element_text(size = 20))
```

-   Using [**auxiliary data**]{.alert2} (the number of doctors at each clinic) can create a more efficient sampling scheme

    -   This idea will come up repeatedly in this course

## MSE: Combining Bias and Variance

-   Sometimes, we might prefer to allow a little bias in trade-off for much reduced variance, and use another measure of the "accuracy" of an estimator:

> [**Mean Squared Error (MSE):**]{.alert2} $MSE(\hat{\theta}) = E[(\hat{\theta}-\theta)^2] = V(\hat{\theta}) + [\text{Bias}(\hat{\theta})]^2$ \
> *Variance plus bias-squared*

-   We describe estimators as:

    -   [**Unbiased**]{.alert} if $E[\hat{\theta}] = \theta$

    -   [**Precise**]{.alert} if $V(\hat{\theta})$ is small

    -   [**Accurate**]{.alert} if $MSE$ is small

-   An estimator can be biased and precise -- but then its not accurate

-   **MSE** vs **Variance**
      -   MSE measures how close an estimate is to the truth
      -   Variance measures how close estimates are across different samples

## Activity 1.1 (Part 2)

Sampling Cats (Part 2)
