---
title:  |
  Logistic/Log-Binomial Regression \ 
  with Survey Data
subtitle: "PUBHBIO 7225 Lecture 21"
# author: "Generative AI acknowledgment: Google Gemini was used to assist in generating alt text for most images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
editor: 
  markdown: 
    wrap: 72
---

## Outline

::: hidden
\newcommand{\logit}{\mathrm{logit}}
:::

::::: columns
::: {.column width="60%"}
**Topics**

-   Logistic regression with survey data\
    (from a probability sample)
-   Log-Binomial regression with survey data\
    (from a probability sample)
:::

::: {.column width="40%"}
**Activities**

-   21.1 Logistic Regression and Log-Binomial Regression with NHANES
:::
:::::

**Assignments**

-   Problem Set 5 (last one!) due Thursday 11/13/2025 11:59pm via
    Carmen\
    (this will be graded by Dr Andridge, not peer reviewed)
-   Quiz 5 (last one!) due Thursday 11/20/2025 11:59pm via Carmen

## Regression with Categorical Outcomes

-   In linear regression, outcome (response variable) at least
    approximately [continuous]{.alert}

-   On surveys, lots (many, most) outcomes of interest are
    [categorical]{.alert}

    -   Multiple choice questions abound!

-   Natural approaches are **generalized linear models** that extend
    linear regression to allow the outcome to be **binary** or
    **ordinal**

-   We will consider two such regression models, in the survey context:

    -   Logistic regression

    -   Log-binomial regression

-   I assume you have seen logistic regression before

-   You may not have seen log-binomial regression; that's okay

## Logistic Regression: Review

-   Outcome = binary $Y$

-   Predictor = continuous or discrete or binary $X$ (or multiple $X$s)

> **Logistic regression model**:
> $$\log \underbrace{\left(\frac{P(Y=1|X)}{1-P(Y=1|X)}\right)}_{\text{odds}} = \mathrm{logit}\left(P(Y=1|X)\right) = \beta_0 + \beta_1 X$$

-   Logit link bounds the predicted probability $0 \le P(Y=1|X) \le 1$

    -   $P(Y=1|X) = \frac{e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}} = \frac{\exp(\beta_0 + \beta_1 X)}{1+\exp(\beta_0 + \beta_1 X)}$

-   Interpretation of unknown parameters $(\beta_0, \beta_1)$ comes via
    exponentiation (to get rid of the log)

## Logistic Regression: Review (con't)

-   Model: $\mathrm{logit}\left(P(Y=1|X)\right) = \beta_0 + \beta_1 X$

-   Interpreting $\beta_1$:

    -   $\beta_1$ = log-odds ratio for a 1 unit increase in $X$

    -   $e^\beta_1$ = odds ratio for a 1 unit increase in $X$

    -   $X$ binary:

        -   The *odds* of $Y=1$ for the group with $X=1$ are $e^\beta_1$
            times the odds for the group with $X=0$

        -   Similar interpretation if $X$ is an indicator variable for
            one category of a multi-category categorical variable --
            compares odds for that category to the reference category

    -   $X$ continuous/discrete:

        -   The *odds* of $Y=1$ multiply by $e^\beta_1$ per one-unit
            increase in $X$

-   Can sound awkward to talk about "odds"

-   But ...**do not** substitute "risk" or "more likely"!

    -   Odds and risk are different things!

## Logistic Regression: Review (con't)

-   Model: $\mathrm{logit}\left(P(Y=1|X)\right) = \beta_0 + \beta_1 X$

-   Interpreting $\beta_0$:

    -   $\beta_0$ = intercept = log-odds that $Y=1$ when $X=0$

    -   $e^\beta_0$ = odds that $Y=1$ when $X=0$

    -   Not super-useful interpretations!

    -   Easier to re-write as:
        $\displaystyle P(Y=1|X=0) = \frac{e^{\beta_0}}{1+e^{\beta_0}}$

    -   The *probability* of $Y=1$ when $X=0$ is
        $\frac{e^{\beta_0}}{1+e^{\beta_0}}$

<p>

-   These interpretations extend to models with more than one
    predictor... interpret as usual, "controlling for" or "adjusting
    for" or "holding constant" the other predictors

## Logistic Regression with Survey Data

-   With **survey data (from probability sample)**, similar scenario to
    linear regression

    -   Basically everything we discussed with linear regression also
        applies to logistic regression

-   Complex sampling design will affect point estimates and standard
    errors

    -   Especially if the design is non-ignorable (a.k.a., informative)
        with respect to the outcome $y$

-   Remember: logistic regression with one binary predictor is
    essentially equivalent to chi-square test

    -   Same ideas apply about needing an adjusted chi-square test (same
        arguments for needing survey-weighted logistic regression)

## One Important Point About Logistic Regression {.pause}

-   If logistic regression model is valid and contains an intercept,
    then the [intercept is **the only** parameter estimate affected by a
    sampling design that depends on $y$]{.alert2}

. . .

-   If you've taken epidemiology, this may sound familiar...
    case-control studies!

    -   Example: Divide population into 2 strata, people with lung
        cancer (cases) and people without lung cancer (controls), and
        select sample from each stratum

        -   Lung cancer is rare → probability of selection for cases is
            much larger than for controls → sampling weights are smaller
            for cases than controls

        -   If primary interest is in estimating effects of age, smoking
            history, etc., on presence of cancer in a logistic
            regression, the [unequal sampling does not matter]{.alert2}

            -   If the model is correctly specified, only difference
                between an unweighted and weighted analysis will be in
                the intercept

            -   Would you ever use $\hat{\beta}_0$ to estimate the
                prevalence of lung cancer in such a study? No!\
                (But if you had the sampling weights you could!)

. . .

-   Of course, if there is clustering it would impact standard errors so
    can't be ignored

## Details: Logistic Regression with Survey Data

-   How do we actually get the survey-weighted ("design-based") estimates?

-   Unlike linear regression, no "closed form" equation for $\hat{\beta}$s

-   In the [**infinite population**]{.alert} (model-based) world, use maximum likelihood (ML) estimation

    -   Find the estimates ($\hat{\beta}$s) that maximize the
        probability ("likelihood") that the observed data $(y_i,x_i)$
        came from the model

-   In the [**finite population**]{.alert2} world, the true $\beta$s -- denoted with Roman letters, $B$ -- are what you'd get
    *if you calculated ML estimates using the full population*

-   Specifically, solving this system of equations for $B$:
    $$\sum_{i=1}^N x_{ij} \left[y_i - \frac{\exp(\mathbf x_i^T \mathbf B)}{1+\exp(\mathbf x_i^T \mathbf B)} \right] = 0 \quad \text{for } j=1,\dots,p$$ produces the true population quantities ($\mathbf B$)

## Details (con't)

-   To obtain estimates we solve the system of equations using the sample and the **sampling weights**:
    $$\sum_{i \in \mathcal{S}} w_i x_{ij} \left[y_i - \frac{\exp(\mathbf{x_i}^T \mathbf{\hat{B}})}{1+\exp(\mathbf{x_i}^T \mathbf{\hat{B}})} \right] = 0 \quad \text{for } j=1,\dots,p$$

-   Variance estimation can be done via **linearization**

<br>

-   Logistic regression gives us **odds ratios**

-   Most surveys are cross-sectional, and we can alternatively calculate
**prevalence ratios**


## Example: Sampling Schools

```{r}
library(tidyverse)
library(haven)
library(survey)
samp <- read_dta("./datasets/apipop_clustersample_2stage.dta") %>%
  mutate(highapi = 1*(api00 > 700),
         highell = 1*(ell > 25))
# set survey design ignoring second stage ("ultimate PSU" analysis)
# des <- svydesign(id = ~dnum + snum, weight = ~wt, fpc = ~N + Mi, data=samp)
des <- svydesign(id = ~dnum, weight = ~wt, fpc = ~N, data=samp)
```

-   Example: 2-stage cluster sample of schools (SSUs) within school districts (PSUs)
from Lecture 13

      -   Stage 1: $n=38$ PSUs out of $N=757$ (districts) -- roughly 5% of total

          -   Total of 246 SSUs (schools) across these 38 sampled PSUs (range: 1 to 36 per PSU)

      -   Stage 2: $m_i=2$ SSUs (schools) per PSU (district)
      
           -   For districts with only 1 school, select the 1 school (i.e., $m_i=M_i=1$)

      -   Result is a total of 65 SSUs

-   Interested in two binary measures:

      -   High percent of English language learners (>25% ELL)
      -   High API (performance) score (>700)
      
## Example (con't)

```{css, echo=FALSE}
.cell .cell-output-stdout pre code {
  font-size: 0.7em !important;
  max-height: 650px; // Adjust this value as needed
  overflow-y: auto; // Add this to handle overflow
}

.reveal code {
  font-size: 0.7em !important;
}
```

```{r}
#| echo: TRUE
# Setting design ignoring second stage sampling ("ultimate PSU analysis")
des <- svydesign(id = ~dnum, weight = ~wt, fpc = ~N, data=samp)
```

```{r}
#| echo: TRUE
# Estimating the population percentage of schools with high API
svymean(~highapi, design=des)
# Estimating the population percentage of schools with high ELL
svymean(~highell, design=des)
```

-   Based on the sample, we estimate that:

      -   `r scales::percent(as.numeric(svymean(~ highapi, des)), accuracy=0.1)` of schools have a high API score (API score >700)

      -   `r scales::percent(as.numeric(svymean(~ highell, des)), accuracy=0.1)` of schools have a high percent of English language learners (>25% ELL)
      
-   Let's see if having high API is associated with the percentage of ELL

## Example (con't)

**Crosstabulation**

```{r}
crosstab <- svyby(~highapi, ~highell, des, svymean)
crosstab_p <- svychisq(~ highapi + highell, des)$p.value
```

```{r}
#| echo: TRUE
svyby(~highapi, ~highell, design=des, svymean)
```

-   `r scales::percent(crosstab[1,2], accuracy=0.1)` of schools **without** a high percentage of ELL have high API
-   `r scales::percent(crosstab[2,2], accuracy=0.1)` of schools **with** a high percentage of ELL have high API

**Chi-square test**

```{r}
#| echo: TRUE
svychisq(~ highapi + highell, design=des)
```

-   p-value = `r round(crosstab_p, 4)`

## Example (con't)

```{r}
p1 <- crosstab[1,2]
p2 <- crosstab[2,2]
or <- round(p1/(1-p1) / (p2/(1-p2)),1)
pr <- round(p1/p2,1)
```

-   Comparing schools **without** high percent ELL to schools **with** high
percent ELL: 

      -   Odds ratio = OR = $\displaystyle \frac{.6615/(1-.6615)}{.1660/(1-.1660)} = \frac{1.954}{.1990} = 9.82$
      -   Prevalence ratio = PR = $\displaystyle \frac{.6615}{.1660} = 3.98$
-   Interpretations:

    -   The odds of having high API for schools with a low percent of ELL
        are `r or` times the odds for schools with a high percent of ELL.
    
    -   Schools with a low percent of ELL are `r pr` times as likely to have high
        API as schools with a high percent of ELL.

-   Odds ratio feels "exaggerated" -- hard to interpret "odds"

-   Prevalence ratio "easier" (more intuitive)

## Example (con't)

-   A survey-weighted logistic regression with high ELL
as the only predictor produces same odds ratio:

```{r}
#| echo: TRUE
svyglm(highapi ~ I(1-highell), family = binomial, design=des)
```

```{r}
beta <- round(as.numeric(coef(svyglm(highapi ~ I(1-highell), family = binomial, design=des))[2]),3)
```

-   OR = $e^\text{`r beta`}$ = `r round(exp(beta),2)`

-   We can also perform a regression analysis that gives us **prevalence ratios (PRs)** instead of odds ratios

## Log-Binomial Regression

-   In the **logistic** regression model, we have:
$$\log \underbrace{\left(\frac{P(Y=1|X)}{1-P(Y=1|X)}\right)}_{\text{odds}} = \mathrm{logit}\left(P(Y=1|X)\right) = \beta_0 + \beta_1 X$$

-   By using a different transformation of $P(Y=1|X)$, we have
**log-binomial** regression:
$$\log \underbrace{\left(P(Y=1|X)\right)}_{\text{"risk"}} = \beta_0 + \beta_1 X$$

-   Note that this different transformation (different link function) does
not have as nice theoretical properties, in the sense that $E(Y|X)$ is
not bounded between 0 and 1 (what are the bounds?)

-   However, unless $P(Y=1|X)$ is close to 0 or 1 it is generally well-behaved

-   Interpretation:

      -   $e^\beta_1$ = prevalence ratio (of $Y=1$) for a 1 unit increase in X

## Log-Binomial Regression with Survey Data

-   Same ideas apply for log-binomial regression as for logistic when you
have survey data

-   Trick for estimation is that you can use **Poisson** regression to obtain estimates (even though it is a log-**binomial** model)

      -   Note that in the [infinite population]{.alert} world, you must use [robust
          standard errors]{.alert} (also called "sandwich estimators") for proper
          variance estimation if you use Poisson regression to fit a
          log-binomial model (otherwise too small)

      -   But in the [finite population]{.alert2} world (survey data), we are not relying on a distributional assumption, and [variances derived from linearization are fine]{.alert2} (using Poisson)

<p>

-   Note: You may have used log-binomial regression before to estimate risk ratios -- this is the same method
    -   Since survey data most often cross-sectional, we prefer to not use "risk" and instead use "prevalence"

## Example (con't)

-   Survey-weighted log-binomial regression (fit using Poisson "trick")

```{r}
#| echo: TRUE
svyglm(highapi ~ I(1-highell), family = poisson, design=des)
```

```{r}
beta <- round(as.numeric(coef(svyglm(highapi ~ I(1-highell), family = poisson, design=des))[2]),3)
```

-   PR = $e^\text{`r beta`}$ = `r round(exp(beta),2)`


## Activity 21.1

Logistic Regression and Log-Binomial Regression with NHANES
