---
title: "Poststratification"
subtitle: "PUBHBIO 7225 Lecture 10"
author: "Generative AI acknowledgment: MS Copilot was used to generate alt text for some images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
editor: 
  markdown: 
    wrap: 72
---

## Outline

::: hidden
\newcommand{\fpchpar}{\left(1-\frac{n_h}{N_h}\right)}
\newcommand{\fpcpar}{\left(1-\frac{n}{N}\right)}
:::

::::: columns
::: {.column width="50%"}
**Topics**

-   Adjusting Survey Weights via:
    -   Poststratification
    -   Raking
:::

::: {.column width="50%"}
**Activities**

-   10.1 Poststratification Weight Adjustments
:::
:::::

<br>

**Assignments**

-   Problem Set 3 due Thursday 10/2/25 11:59pm via Carmen

## Limitation of Stratification at the Design Stage

-   To use a variable for stratification at the design stage, you must
    know the value of that variable for all units in the population
    **before** you conduct the survey

    -   In other words, stratification variables must be available on
        the sampling frame

-   For some surveys, frame data may be quite limited (especially phone
    surveys)

-   Example: Ohio Medicaid Assessment Survey (OMAS) 2017

    -   Phone survey – only geography was available for stratification
    -   Can estimate population-level characteristics (e.g., total
        number of white people who live in Ohio)
    -   But, we might know the true population totals for these
        characteristics (e.g., true number of white Ohioans) from
        another source
    -   Suppose we get an estimate from the data of
        $\hat{t}_{str} = 8,320,817$ white Ohioans
    -   We know the “true” number of white Ohioans is $t= 7,164,904$\
        (from American Community Survey 5-year estimates)

-   How can we reconcile this? Seems like we **oversampled** white
    people. . .

## Example: SRS of OSU Students

```{r}
options(scipen = 999)
N <- 61443
n <- 100
n1 <- 67
n2 <- 23
n3 <- 10
p1 <- n1/n
p2 <- n2/n
p3 <- n3/n
N1 <- 46815 
N2 <- 11404
N3 <- 3224
w <- N/n
w_round <- round(w,1)
Nhat1 <- n1*w
Nhat2 <- n2*w
Nhat3 <- n3*w
o1 <- 49
o2 <- 14
o3 <- 4
po1 <- round(o1/n1,3)
po2 <- round(o2/n2,3)
po3 <- round(o3/n3,3)
ps_o <- N1/N*o1/n1 + N2/N*o2/n2 + N3/N*o3/n3
psw1 <- w*N1/Nhat1
psw2 <- w*N2/Nhat2
psw3 <- w*N3/Nhat3
M1 <- 42191
M2 <- N - M1
```

-   You are interested in student perception of campus culture at OSU
    (Columbus Campus)

-   There are $N=$ `r N` students on the Columbus campus

-   You have access to a sampling frame that contains email addresses,
    but no additional information on the students (i.e., rank, gender,
    etc.)

-   Thus, you take a **simple random sample** of $n=$ `r n` students and
    send them a survey (cannot stratify on anything b/c your sampling
    frame doesn't contain anything useful)

-   Probability of selection and sample weight for each of the $n=$
    `r n` students sampled:

    <br>

    -   Probability of selection: $\pi_i =$

    <br>

    -   Sampling weight: $w_i = 1/\pi_i =$

    <br>

-   Miraculously, everyone replies (we will get to nonresponse later!)

## Example: SRS of OSU Students (con't)

-   On the survey you ask questions about:

    -   Level: Undergraduate, Graduate, Professional

    -   Residency: From Ohio, Not from Ohio

-   Your sample has `r n1` undergrads, `r n2` grad students, `r n3`
    professional students

-   Since an SRS, unweighted proportions are the same as weighted

-   Resulting estimates:

    -   Undergrads: $\hat{p}_{ug}$ = `r n1`/`r n` = `r n1/n`
    -   Grad students: $\hat{p}_{grad}$ = `r n2`/`r n` = `r n2/n`
    -   Professional students: $\hat{p}_{prof}$ = `r n3`/`r n` =
        `r sprintf("%.2f", n3/n)`

## Example: SRS of OSU Students (con't)

-   But, you know from the OSU Statistical Summary that there are `r N1`
    undergraduates, `r N2` graduate students, and `r N3` professional
    students ([Statistical Summary
    2024-2025](https://irp.osu.edu/sites/default/files/documents/2025/03/2024-25-statistical-summary.pdf))

-   Comparing the proportions:

|               | Sample |            | Population |                                  |
|---------------|--------------:|---------------|--------------:|---------------|
|               |  Count | Proportion |      Count | Proportion                       |
| Undergraduate | `r n1` | `r p1`     |     `r N1` | `r N1`/`r N` = `r round(N1/N,3)` |
| Graduate      | `r n2` | `r p2`     |     `r N3` | `r N2`/`r N` = `r round(N2/N,3)` |
| Professional  | `r n3` | `r p3`     |     `r N3` | `r N3`/`r N` = `r round(N3/N,3)` |
| Total         |  `r n` |            |      `r N` |                                  |

-   It seems that grad students and professional students are
    **over-represented** in your sample, and undergrads are
    **under-represented**

-   What can you do about this?

## Post-Survey Weight Adjustments

-   Each student in your sample had a weight of $w_i = N/n =$
    `r N`/`r n` = `r round(w,1)`

-   Thus, our estimates of the population size in each group :

| **Group** | **\# in sample** | **Weight** $\mathbf{(w_i)}$ | **Estimated Total** $(\hat{t})$ | **True Total** $(t)$ |
|:--------------|:-------------:|:-------------:|:--------------|:-------------:|
| Undergrads | `r n1` | `r w_round` | `r n1` $\times$ `r w_round` = `r round(n1*w,1)` | `r N1` |
| Grads | `r n2` | `r w_round` | `r n2` $\times$ `r w_round` = `r round(n2*w,1)` | `r N2` |
| Profs | `r n3` | `r w_round` | `r n3` $\times$ `r w_round` = `r round(n3*w,1)` | `r N3` |

<br>

-   Our estimates are not equal to the known truth

-   But, we can do something about this!

-   We can adjust the weights within each group so that estimated totals
    equal true totals

-   We want to **down-weight** grad students and professionals and
    **up-weight** undergrads

-   This is called **Poststratification**

## Poststratification

> [**Poststratification =**]{.alert} dividing the sample into subgroups
> (post-strata) based on known population characteristics and adjusting
> the survey weights within those subgroups to match the population
> totals

-   Steps for poststratification:

    -   Divide sampled units into subgroups (post-strata) based on
        characteristics **measured in the sample**
    -   Population size of each subgroup is assumed to be known
    -   Adjust sample weights within each post-stratum so that estimated
        counts equal the true counts
        -   All units within a subgroup have the same adjustment made\
            (e.g., all up-weighted or down-weighted similarly)

<br>

-   Post-stratification **aligns the representation of the
    poststratification variable(s) to match that of the population**

## Adjusting the Weights for the SRS of OSU Students

| **Group** | **\# in sample** | **Weight** $\mathbf{(w_i)}$ | **Estimated Total** $(\hat{t})$ | **True Total** $(t)$ |
|:--------------|:-------------:|:-------------:|:--------------|:-------------:|
| Undergrads | `r n1` | `r w_round` | `r n1` $\times$ `r w_round` = `r round(n1*w,1)` | `r N1` |
| Grads | `r n2` | `r w_round` | `r n2` $\times$ `r w_round` = `r round(n2*w,1)` | `r N2` |
| Profs | `r n3` | `r w_round` | `r n3` $\times$ `r w_round` = `r round(n3*w,1)` | `r N3` |

-   Start with the undergrads -- what should the weight be for the 67
    undergrads if you want to end up with the true population count as
    your estimate? (How many units should each of the 67 UGs
    "represent"?)

<br> <br>

-   What about the grad students and the professional students?

## Poststratified Weights

-   Should remind you of stratified sampling -- different weight in each
    (post)stratum -- but now the $n_h$ are not fixed in advance so we
    have added variability

-   For unit $j$ in post-stratum $h$: $$\begin{aligned}
    \text{\textcolor{blue}{Poststratified weight}} &= \text{\textcolor{red}{original weight}} \times \text{\textcolor{purple}{adjustment factor}} & \\
    \textcolor{blue}{w_{hj}^{PS}} &= \textcolor{red}{w_{hj}} \times \textcolor{purple}{\frac{\text{Population total in post-stratum}~h}{\text{Estimated total in post-stratum}~h}} \\
    & = \textcolor{red}{w_{hj}} \times \textcolor{purple}{\frac{N_h}{\widehat{N}_h}}
    \end{aligned}$$

-   You can poststratify on multiple variables if you know the
    population totals **in their crosstabulation**

    -   For example, you need to know level-by-residency totals:
        -   Undergraduates from Ohio
        -   Undergraduates not from Ohio
        -   Grad students from Ohio... etc.

## Poststratification of the SRS of OSU Students

|  |  |  |  |  |  |
|------------|:----------:|:----------:|-----------:|-----------:|-----------:|
|  |  | **Orig. Weight** | **Est. Total** | **True Total** | **Poststratified Weight** |
| **Post-stratum** | $n_h$ | $w_{hj}$ | $\widehat{N}_h$ | $N_h$ | $w_{hj}^{PS} = w_{hj} \times \frac{N_h}{\widehat{N}_h}$ |
| Undergrad | `r n1` | `r w_round` | `r Nhat1` | `r N1` | `r w_round`$\times$ `r N1` / `r Nhat1` = `r round(w*N1/Nhat1, 1)` |
| Grad | `r n2` | `r w_round` | `r Nhat2` | `r N2` | `r w_round`$\times$ `r N2` / `r Nhat2` = `r round(w*N2/Nhat2, 1)` |
| Professional | `r n3` | `r w_round` | `r Nhat3` | `r N3` | `r w_round`$\times$ `r N3` / `r Nhat3` = `r round(w*N3/Nhat3, 1)` |

-   Notice that the weights went **up** for undergrads, **down** for
    grad and professional

-   These new weights, $w_{hj}^{PS}$, replace the original weights,
    $w_{hj}$ for all calculations

-   This eﬀectively creates a stratified sample → use stratified
    sampling formulae for estimates

    -   E.g., Overall proportion:
        $\hat{p}_{PS} = \sum_{h=1}^H \frac{N_h}{N} \hat{p}_h$

## Estimates after Poststratification

-   For example, estimating the **proportion of students from Ohio**:

| **Post-stratum** | **Sample Proportion from Ohio** $\mathbf{(\hat{p}_h)}$ | **Stratum Size** $\mathbf{(N_h)}$ |
|:------------------|:--------------------------:|:-----------------------:|
| Undergrad $(h=1)$ | `r o1`/`r n1` = `r o1/n1` | `r N1` |
| Grad $(h=2)$ | `r o2`/`r n2` = `r o2/n2` | `r N2` |
| Professional $(h=3)$ | `r o3`/`r n3` = `r o3/n3` | `r N3` |
| Total | `r (o1+o2+o3)`/`r n` = `r (o1+o2+o3)/n` | `r N` |

-   **Poststratified** estimate of overall proportion from Ohio:
    $$\begin{flalign}
    \hat{p}_{PS} &= \sum_{h=1}^H \frac{N_h}{N} \hat{p}_h = \frac{N_1}{N} \hat{p}_1 + \frac{N_1}{N} \hat{p}_1 + \frac{N_3}{N} \hat{p}_3 & \\
    &= \frac{\text{`r N1`}}{\text{`r N`}} \times \text{`r po1`} + \frac{\text{`r N2`}}{\text{`r N`}} \times \text{`r po2`} + \frac{\text{`r N3`}}{\text{`r N`}} \times \text{`r po3`} = \textbf{`r round(ps_o,3)`}
    \end{flalign}$$
-   Slightly higher than the original estimate of `r 100*(o1+o2+o3)/n`%
    -- undergrads were **underrepresented** in the sample, and they are
    **more likely** to be from Ohio

## Estimates after Poststratification Using Adjusted Weights

-   Let
    $y_i = \begin{cases} 1 & \text{person $i$ is from Ohio} \\ 0 & \text{person $i$ is NOT from Ohio} \end{cases}$

-   Poststratified weights:
    $w_i^{PS} = \begin{cases} \text{`r round(psw1,1)`} & \text{Undergrad (stratum 1)} \\ \text{`r round(psw2,1)`} & \text{Grad (stratum 2)} \\ \text{`r round(psw3,1)`} & \text{Professional (stratum 3)} \end{cases}$
    $$\begin{aligned}
    \hat{t}_{PS} = \sum_{i \in S} w_i^{PS} y_i &= \sum_{h = 1}^{3} \sum_{j \in S_h} w_{hj}^{PS} y_{hj} = \underbrace{\sum_{j \in S_1} \text{`r round(psw1,1)`} \times y_{1j}}_{\text{stratum 1}} + \underbrace{\sum_{j \in S_2} \text{`r round(psw2,1)`} \times y_{2j}}_{\text{stratum 2}} + \underbrace{\sum_{j \in S_3} \text{`r round(psw3,1)`} \times y_{3j}}_{\text{stratum 3}} \\
    &= \text{`r round(psw1,1)`} \times \underbrace{\sum_{j \in S_1} y_{1j}}_{\text{\# from OH}} + \text{`r round(psw2,1)`} \times \underbrace{\sum_{j \in S_2} y_{2j}}_{\text{\# from OH}} + \text{`r round(psw3,1)`} \times \underbrace{\sum_{j \in S_3} y_{3j}}_{\text{\# from OH}}\\
    & = \text{`r round(psw1,1)`} \times \text{`r o1`} + \text{`r round(psw2,1)`} \times \text{`r o2`} + \text{`r round(psw3,1)`} \times \text{`r o3`} = \text{`r round(psw1*o1 + psw2*o2 + psw3*o3,1)`}
    \end{aligned}$$

-   Thus
    $\hat{p}_{PS} = \frac{\hat{t}_{PS}}{N} = \frac{\text{`r round(psw1*o1 + psw2*o2 + psw3*o3,1)`}}{\text{`r N`}} = \text{`r round((psw1*o1 + psw2*o2 + psw3*o3)/N,3)`}$

## Why Does Poststratification Work?: The Theory

-   Unlike stratification at the design stage, the $n_h$ in poststrata
    **are random variables**

-   Take a different SRS and the number in each poststrata will be
    **different** (unlike fixed sizes of design strata)

-   But -- the poststratified mean estimate is still an estimator for
    the true mean $$\begin{aligned}
    E[\bar{y}_{PS}] &= E\left[\sum_{h=1}^H \frac{N_h}{N} \bar{y}_h \right] = E\left[ E\left(\sum_{h=1}^H \frac{N_h}{N} \bar{y}_h \bigg\vert n_1, n_2, \dots, n_h \right) \right]   \qquad \text{\small (condition on fixed poststratum sizes)} \\
    &= E\left[ \sum_{h=1}^H \frac{N_h}{N} E\left( \bar{y}_h \bigg\vert n_1, n_2, \dots, n_h \right) \right] = E\left[ \sum_{h=1}^H \frac{N_h}{N} \bar{y}_{hU} \right] = \bar{y}_U
    \end{aligned}$$\

-   So why would we ever stratify at the design stage?

    -   Technically, the variance of $\bar{y}_{PS}$ is larger than
        $\bar{y}_{str}$ because the $n_h$ are random

    -   Also, guard against a fluke "bad sample"

## Variance Estimation after Poststratification (of an SRS)

-   If the strata are fixed at the design stage ($n_h$ are fixed), we
    have the usual formula:
    $V(\bar{y}_{str}) = \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2  \left(1-\frac{n_h}{N_h}\right)\frac{S_h^2}{n_h}$

-   With poststratification, $n_h$ are random variables and we can do
    some trickery to get: $$\begin{aligned}
    V(\bar{y}_{PS}) &= E[ V(\bar{y}_{str} | n_1, n_2, \dots, n_h)] + V[ E(\bar{y}_{str} | n_1, n_2, \dots, n_h)] & \\
    &= E\left[ \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2  \left(1-\frac{n_h}{N_h}\right)\frac{S_h^2}{n_h}\right] + \underbrace{V(\bar{y}_U)}_{\text{\small (= 0)}} = E\left[\sum_{h=1}^H \left(\frac{N_h}{N}\right)^2  \frac{S_h^2}{n_h} - \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2  \frac{S_h^2}{N_h}\right]\\
    &= \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2 S_h^2 E\left(\frac{1}{n_h}\right) - \sum_{h=1}^H \left(\frac{N_h}{N}\right)^2  \frac{S_h^2}{N_h}  \quad (\text{\small only random variable is }n_h)\\
    &= \dots \text{using an approximation for } E(1/n_h) \text{ and some algebra}\dots \\
    &\approx \underbrace{\sum_{h=1}^H \frac{N_h}{N}\left(1-\frac{n}{N}\right)\frac{S_h^2}{n}}_{\text{variance under proportional allocation!}} + \underbrace{\sum_{h=1}^H \left(1-\frac{N_h}{N} \right) \frac{S_h^2}{n^2}}_{\text{extra variability due to $n_h$ being random}}
    \end{aligned}$$

## Variance Estimation (con't)

$\displaystyle V(\bar{y}_{PS}) = \textcolor{red}{\sum_{h=1}^H \frac{N_h}{N}\left(1-\frac{n}{N}\right)\frac{S_h^2}{n}} + \textcolor{blue}{\sum_{h=1}^H \left(1-\frac{N_h}{N} \right) \frac{S_h^2}{n^2}}$

-   Notice that the is of order $n^{-2}$ ($n^2$ in denominator)

-   Usually the 2nd term is really small compared to the first term

-   Thus, variance of the poststratified mean (of an SRS) is usually
    effectively **the same as variance of the stratified mean under
    proportional allocation**!

-   This holds as long as the total sample size $n$ is "large" (and it
    is in most surveys)

-   Poststratification (or related weight adjustment methods) are **very
    commonly used** in household surveys

    -   Often don't have demographic info on the frame (example: OMAS),
        or it is limited

    -   Do have access to population counts of people based on
        demographics such as age/race groups from the Census or other
        source

    -   Thus, use age and race (cross-classified) to post-stratify

-   Importantly -- **poststratification can be used with complex survey
    designs(not just SRS)**

## Activity 9.1

Poststratification

## Post-Survey Weight Adjustments: Raking

-   Often we can get the population totals for a set of variables --
    age, race, etc. -- but only the **marginal counts**

    -   We know the number in each age group, and the number in each
        race group, but not in groups formed by crossing age and race
        (e.g., 18-25 year old white individuals, 18-25 year old Black
        individuals, etc.)

-   In this case, a method called **raking** can be used to adjust the
    weights to achieve the correct (true) marginal totals

    -   You may see raking also referred to as "iterative proportional
        fitting"

    -   In **raking** you iteratively adjust the sample weights to match
        each set of marginal totals, going back and forth between
        margins (variables) until convergence (until all margins "match"
        the truth)

    -   This is easiest seen in an example...

## Example: Raking

```{r}
n11 <- 49
n12 <- 18
n21 <- 14
n22 <- 9
n31 <- 4
n32 <- 6
n_oh <- n11 + n21 + n31
n_notoh <- n12 + n22 + n32
```

:::::: smaller-text
::::: columns
::: {.column width="50%"}
|              |               |        |        |
|--------------|:-------------:|:------:|:------:|
| **MARGINS**  | **Residency** |        |        |
| **Level**    |      OH       | Non-OH | Total  |
| Undergrad    |       ?       |   ?    | `r N1` |
| Grad         |       ?       |   ?    | `r N2` |
| Professional |       ?       |   ?    | `r N3` |
| Total        |    `r M1`     | `r M2` | `r N`  |
:::

::: {.column width="50%"}
|                 |               |             |        |
|-----------------|:-------------:|:-----------:|:------:|
| **SAMPLE DATA** | **Residency** |             |        |
| **Level**       |      OH       |   Non-OH    | Total  |
| Undergrad       |    `r n11`    |   `r n12`   | `r n1` |
| Grad            |    `r n21`    |   `r n22`   | `r n2` |
| Professional    |    `r n31`    |   `r n32`   | `r n1` |
| Total           |   `r n_oh`    | `r n_notoh` | `r n`  |
:::
:::::
::::::

-   Start with the row variable (**level**), and adjust the sample
    weights via poststratification

    -   Same as when we poststratified by level -- new weights for each
        level\
        (undergrad: $w_i$ = `r round(psw1,1)`; grad: $w_i$ =
        `r round(psw2,1)`; professional: $w_i$ = `r round(psw3,1)`)

-   Now totals for **level** are right -- but totals for the other
    variable (**residency**) are not

    -   Ex: Sum of weights for OH = `r n11` $\times$ `r round(psw1,1)` +
        `r n21` $\times$ `r round(psw2,1)` + `r n31` $\times$
        `r round(psw3,1)` = `r round(n11*psw1+n21*psw2+n31*psw3)` -- but
        should be `r M1`

-   So, adjust the weights via poststratification using the other
    variable (**residency**)

    <!-- -   Example: For OH resident, multiply the (previously adjusted) -->

    <!--     weight times `r M1`/`r round(n11*psw1+n21*psw2+n31*psw3)` ($= N_h$/$\widehat{N}_h$) -->

-   But now the totals for the first variable (**level**) are wrong...so
    go back and adjust with those margins

-   Repeat until weights don't change (and marginal totals are correct!)

## Example: Raking (con't)

```{r}
library(survey)
library(tidyverse)
dat <- tibble(id = 1:n, w = N/n, 
              residency = c(rep("OH", n11+n21+n31), rep("NOT", n12+n22+n32)),
              level = c(rep(1, n11), rep(2, n21), rep(3, n31),
                        rep(1, n12), rep(2, n22), rep(3, n32)))
des <- svydesign(~1, weights = ~w, data = dat)
pop_residency <- data.frame(residency = c("OH","NOT"), Freq = c(M1,M2))
pop_level <- data.frame(level = c(1,2,3), Freq = c(N1, N2, N3))
des_raked <- rake(des, 
                  list(~residency, ~level), 
                  list(pop_residency, pop_level))
dat$w_rake <- weights(des_raked)
wr11 <- round(dat$w_rake[dat$level==1 & dat$residency=="OH"][1],3)
wr21 <- round(dat$w_rake[dat$level==2 & dat$residency=="OH"][1],3)
wr31 <- round(dat$w_rake[dat$level==3 & dat$residency=="OH"][1],3)
wr12 <- round(dat$w_rake[dat$level==1 & dat$residency=="NOT"][1],3)
wr22 <- round(dat$w_rake[dat$level==2 & dat$residency=="NOT"][1],3)
wr32 <- round(dat$w_rake[dat$level==3 & dat$residency=="NOT"][1],3)
```

Resulting survey weights for the SRS of OSU students:

| Group                   | Sample Count | Original Weight | Raked Weight |
|:------------------------|:-------------|:----------------|:-------------|
| Undergrad / Ohio        | `r n11`      | `r w_round`     | `r wr11`     |
| Grad / Ohio             | `r n21`      | `r w_round`     | `r wr21`     |
| Professional / Ohio     | `r n31`      | `r w_round`     | `r wr31`     |
| Undergrad / Not Ohio    | `r n12`      | `r w_round`     | `r wr12`     |
| Grad / Not Ohio         | `r n22`      | `r w_round`     | `r wr22`     |
| Professional / Not Ohio | `r n32`      | `r w_round`     | `r wr32`     |

-   Notice that now we will get the **correct totals for level and for
    residency**!

    -   Ex: Number of students from Ohio = `r n11` $\times$ `r wr11` +
        `r n21` $\times$ `r wr21` + `r n31` $\times$ `r wr31` =
        `r round(n11*wr11 + n21*wr21 + n31*wr31)` ✓

    -   Ex: Number of undergrads = `r n11` $\times$ `r wr11` + `r n12`
        $\times$ `r wr12` = `r round(n11*wr11 + n12*wr12)` ✓

-   But, we are not guaranteed to get the correct level-by-residency
    totals

    -   Ex: Estimated number of undergrads from Ohio = `r n11` $\times$
        `r wr11` = `r round(n11*wr11)` ← **not guaranteed to be right**

## Raking: Some Properties

**Pros**

-   Can be done with more than two variables

-   Does not always converge... diagnostics are often used to make sure
    adjustments are not too severe

-   Used extensively in U.S. household surveys where Census or other
    source can provide high quality totals

-   Used even when totals are not technically "known" but rather
    estimated with high precision -- and with higher precision than
    fully crossed cells (e.g., level-by-residency) that would be needed
    for poststratification

**Con**

-   Can cause inflation of variances if variables used for
    poststratification are not related to the survey outcomes of
    interest

    -   Just like you want to stratify on variable(s) related to your
        key survey outcomes, you also want to poststratify or rake to
        variable(s) related to your key survey outcomes

Importantly -- **raking can be used with complex survey designs (not
just SRS)**

## Caveats for Post-Survey Weight Adjustments

-   Theoretically, could use a lot of variables for poststratification
    and raking

    -   Problem: Resulting poststrata can end up with small estimated
        total counts $(\widehat{N}_h)$, which can make the weights
        highly variable and unstable

-   Options:

    -   Collapse categories in the variable(s) used for
        poststratification/raking (ad hoc solution)

    -   Use a model-based method that can handle more poststrata/raking
        variables and ensure groupings are not too small or weight
        adjustments are not too extreme

        -   Example: OMAS uses something called a Generalized
            Exponential Model (GEM) to simultaneously adjust the weights
            to a large set of auxiliary variables -- basically,
            super-charged raking that sets bounds on how much weights
            can be changed

-   Poststratification and raking are also commonly used for
    **nonresponse adjustment** and to **adjust for undercoverage**, so
    we will come back to them later in the course

-   Both poststratification and raking fall under an umbrella method
    called **weight calibration** -- you may see this terminology used
    to describe your chosen group project surveys
