---
title:  |
  Nonresponse: Intro to Missing Data
subtitle: "PUBHBIO 7225 Lecture 17"
author: "Generative AI acknowledgment: Google Gemini was used to generate alt text for some images"
format:
  revealjs: 
    slide-number: true
    smaller: true
    footer: "PUBHBIO 7225"
    width: 1280
    height: 720
    theme: custom.scss
    html-math-method: mathjax
editor: 
  markdown: 
    wrap: 72
self-contained: true
---

## Outline

<!-- ::: hidden -->

<!-- \newcommand{\ybar}{\bar{y}} -->

<!-- ::: -->

::::: columns
::: {.column width="60%"}
**Topics**

-   Missing Data

    -   Nonresponse Bias

    -   Preventing Nonresponse

    -   Unit and Item Nonresponse

    -   Taxonomy of Missing Data
:::

::: {.column width="40%"}
**Activities**

-   17.1 The Impact of Missing Data
:::
:::::

**Readings**

-   Andridge RR and Lemeshow S (2017). *Nonresponse in Sample Surveys*,
    In Wiley StatsRef: Statistics Reference Online (PDF on Carmen)

**Assignments**

-   Peer Evaluation of Problem Set 4 due Tuesday 10/28/25 11:59pm via
    Carmen
-   Quiz 4 due Thursday 10/30/2025 11:59pm via Carmen

## The Problem of Nonresponse {.pause}

-   To this point we have assumed every sampled unit responded to the
    survey

-   When we sample "things" that aren't people, this might be reality

    -   E.g., sampling to determine percentage of beds using netting in
        rural Gambia

    -   Presumably the surveyors were able to (correctly) count beds and
        beds with netting for all sampled compounds (within villages,
        within districts)

-   But, when sampling people there is usually (always?) some fraction
    of the sample that fails to respond

. . .

-   Two main problems caused by nonresponse:

    -   Reduction in power / loss of efficiency

    -   Potential for bias

-   For surveys, in general **bias** is the major concern

## Total Survey Error Diagram

Nonresponse is only one part of the TSE framework...

![Adapted from Groves et al. (2009), Survey Methodology, 2nd
Edition](figures/TSEdiagram.png){fig-alt="A flowchart diagram of the Total Survey Error (TSE) framework. The diagram is split into two main vertical columns, \"Measurement\" on the left and \"Representation\" on the right. In the Measurement column, the flow goes from \"Construct\" to \"Measurement\" to \"Response\" to \"Edited Data.\" Red arrows pointing to the right of each step label the sources of error: \"Specification Error (Validity)\" for the first step, \"Measurement Error\" for the second, and \"Processing Error\" for the third. The Representation column flows from \"Target Population\" to \"Sampling Frame\" to \"Sample\" to \"Respondents\" to \"Postsurvey Adjustments.\" Similarly, red arrows label the errors at each step: \"Coverage Error,\" \"Sampling Error,\" \"Nonresponse Error,\" and \"Adjustment Error.\" Both columns converge at the bottom, with arrows from \"Edited Data\" and \"Postsurvey Adjustments\" pointing to a final box labeled \"Survey Statistic.\""}

...but it is arguably the biggest source of error, or at least one of
the biggest

## How Nonresponse Causes Bias {.pause}

-   Imagine that the target population can be partitioned into two
    groups: **respondents** and **nonrespondents**

| **Group**     | **Population Size** | **Population Mean**        | **Population Variance** |
|:----------------|:---------|:----------------|:-------------|
| Respondents     | $N_R$    | $\bar{y}_{RU}$  | $S^2_R$      |
| Non-Respondents | $N_{NR}$ | $\bar{y}_{NRU}$ | $S^2_{NR}$   |
| Total           | $N$      | $\bar{y}_U$     | $S^2$        |

-   The overall population mean is a weighted average of the two group
    means:
    $$\bar{y}_U = \frac{N_R}{N} \bar{y}_{RU} + \frac{N_{NR}}{N} \bar{y}_{NRU}$$

. . .

-   Now suppose we take an SRS of size $n$ (from the full population of $N$ units), and $n_R$ units respond

      -   Respondents are (by definition) all from the "Respondents" group
      
      -   Nonrespondents are from the "Nonrespondents" group
    
## How Nonresponse Causes Bias (con't) {.pause}

| **Group**     | **Population Size** | **Population Mean**        | **Population Variance** |
|:----------------|:---------|:----------------|:-------------|
| Respondents     | $N_R$    | $\bar{y}_{RU}$  | $S^2_R$      |
| Non-Respondents | $N_{NR}$ | $\bar{y}_{NRU}$ | $S^2_{NR}$   |
| Total           | $N$      | $\bar{y}_U$     | $S^2$        |

-   Sample size: $n$

-   Number of respondents: $n_R$

-   Sample mean based on respondents only: $\bar{y}_R$

. . . 

-   $E[\bar{y}_R] = \bar{y}_{RU}$

    -   Respondent sample mean is an unbiased estimate of the **respondent** mean
   
. . . 
 
-   $E[\bar{y}_R] \ne \bar{y}_U$

    -   Respondent sample mean is **not** an unbiased estimate of the **overall** mean

. . . 

-   ...unless the mean is the same for respondents and nonrespondents (i.e., unless $\bar{y}_{RU} = \bar{y}_{NRU} = \bar{y}_U$)

## How Nonresponse Causes Bias (con't) {.pause}

-   The bias of the respondent mean is: $$\begin{aligned}
    \text{Bias}(\bar{y}_R) &= E(\bar{y}_R) - \bar{y}_U\\
    &= \bar{y}_{RU} - \left[ \frac{N_R}{N} \bar{y}_{RU} + \frac{N_{NR}}{N} \bar{y}_{NRU} \right] = \textcolor{red}{\frac{N_{NR}}{N}} \textcolor{blue}{(\bar{y}_{RU} - \bar{y}_{NRU})}
    \end{aligned}$$

-   Bias in the respondent mean depends on [nonresponse rate]{.alert}
    and diﬀerence between [respondents and nonrespondents]{.alert2}

-   We cannot ever actually know this bias -- we don't observe the
    nonrespondents (can't estimate $\bar{y}_{NRU}$)

. . . 

-   But this expression tells us that **nonresponse bias is small if**:

    1.  [Nonresponse rate is small]{.alert} ($\frac{N_{NR}}{N}$ close to 0)\
        or

    2.  [Mean for nonrespondents is close to the mean for the
        respondents]{.alert2} $(\bar{y}_{NRU} \approx \bar{y}_{RU})$

-   We can't ever know if [(2)]{.alert2} is true, so our best bet is to
    try to achieve [(1)]{.alert}

## How Nonresponse Causes Bias (con't) {.pause}

-   Some additional comments:

    -   Increasing sample size does not help -- doesn't guarantee
        $\bar{y}_{NRU} \approx \bar{y}_{RU}$
        -   Might just get "more of the same"
    -   Bias can (and will) differ for different survey items (i.e., for
        different $y$)

-   Best strategy is to try to avoid nonresponse altogether!

<br>

. . .

> **Key idea:** If an outcome, $y$, is "related" to nonresponse --
    i.e., it differs between respondents and nonrespondents -- then an
    estimate based on respondents only will be biased (without
    additional adjustment)

-   Essentially, if there is nonresponse, the best-case scenario is that
    the respondents are a **random sample** of the full sample.

## Activity 17.1 (Part 1)

The Impact of Missing Data (Part 1)

## Nonresponse a Growing Problem

-   Response rates steadily declining for large Federally-sponsored
    surveys[^1]

-   For seven major surveys, 1995 to 2015 (Y-axis is 65% to 95%):

[^1]: Czajka JL and Beyler A (2016) Declining Response Rates in Federal
    Surveys: Trends and Implications. *Washington, DC: Mathematica
    Policy Research*

![](./figures/Czajka_RRs){fig-alt="A line graph depicting the response rates for seven major surveys from 1995 to 2015. The y-axis shows response rates from 65% to 95%. While the CPS survey started with the highest response rate around 94% and remained high, all surveys show a general downward trend over time."
fig-align="center" width="60%"}

## Nonresponse a Growing Problem (con't)

-   Recent response rates (2015-2025) for **Bureau of Labor Statistics
    household surveys** [(chart
    source)](https://www.bls.gov/osmr/response-rates/)

![](./figures/household-survey-respons){fig-alt="A line graph showing the monthly household survey response rates from April 2015 to April 2025. The CPS consistently has the highest response rates, generally staying above 70%, except for a sharp decline in 2020. The other surveys, including ATUS, CE-Diary, CE-Interview, CPI-Housing, and TPOPS, show similar trends with lower overall rates, all experiencing a sharp decline in 2020 and a gradual downward trend over the decade."
fig-align="center"}

-   Note the COVID dip!

## Nonresponse a Growing Problem (con't)

-   **ACS (American Community Survey)** = primary source of information
    about U.S. demographics and housing

-   Recent response rates for ACS (2000-2023) [(chart data
    source)](https://www.census.gov/acs/www/methodology/sample-size-and-data-quality/response-rates/)

```{r}
#| fig-alt: "A line graph depicting the American Community Survey Housing Unit response rates from 2000 to 2023. The y-axis ranges from 70% to 100%. The graph shows a general decline after around 2014, with a small dip in 2004, a larger dip in 2013, and a very large dip in 2020 to a low of just over 70%. The response rate recovers to around 85% in 2021 and stays there through 2023."
#| fig-align: center
library(tidyverse)
library(readxl)
theme_set(theme_bw())
acs <- read_xlsx("./datasets/Response_Rate_UnitedStates.xlsx", range="A4:J28") %>%
  janitor::clean_names()
acs %>%
  ggplot(aes(x = year, y = response_rate)) +
  geom_point(size = 2) +
  geom_line() +
  labs(x = "", y = "Response Rate (%)") + 
  scale_x_continuous(breaks = seq(2000, 2023, 1)) +
  lims(y = c(70,100)) + 
  # scale_y_continuous(breaks = seq(70, 100, 10)) + 
  theme(axis.text.x = element_text(size = 14, angle = 90, vjust = 0.5),
        axis.text.y = element_text(size = 16),
        axis.title = element_text(size = 20))
```

-   RR declined
    `r acs$response_rate[acs$year==2010] - acs$response_rate[acs$year==2023]`
    percentage points between 2010 and 2023

-   Again, note the COVID dip!

-   Note: ACS estimates are often used as the “truth” for post-survey
    adjustments (e.g., post-stratification) for other surveys

## What Are Some Factors That Could Be Causing This?

<br>

<br>

<br>

<br>

<br>

<br>

**And What Can Be Done To Mitigate/Prevent It?**

## A Particular Growing Problem

-   Looking at the ACS reasons for nonresponse shows that **refusal** is a growing problem

```{r}
#| fig-alt: "A line graph depicting the American Community Survey Housing Unit percentage of refusals and percentage of no one at home from 2000 to 2023. The y-axis ranges from 0% to 10%. The graph shows a striking increase in percent of refusals around 2017, jumping up from around 2.5% to around 9% in 2023. The percentage of no one home is steady around 1% until around 2020 when there is a modest increase to around 2% in 2023."
#| fig-align: center

# make long to plot 
long <- acs %>%
  pivot_longer(c(refusal, no_one_home), names_to = "reason", values_to = "pct") %>%
  mutate(reason_text = case_when(reason == "refusal" ~ "Refusal",
                                 reason == "no_one_home" ~ "No One Home"))
last <- long %>%
  filter(year == 2023)
long %>%
  ggplot(aes(x = year, y = pct, color = reason, linetype = reason)) +
  geom_point(size = 2) +
  geom_line() +
  labs(x = "", y = "Percentage (%)") + 
  scale_x_continuous(breaks = seq(2000, 2023, 1)) +
  lims(y = c(0,10)) + 
  theme(axis.text.x = element_text(size = 14, angle = 90, vjust = 0.5),
        axis.text.y = element_text(size = 16),
        axis.title = element_text(size = 20),
        legend.position = "none") + 
  coord_cartesian(x = c(2000,2030)) +
  geom_text(data = last, aes(x = year+0.5, y = pct, label = reason_text), hjust = 0, size = 6)
```

## Designing Surveys to Minimize Nonresponse

Many "design" considerations to maximize response rates:

:::: incremental
::: smaller-text
-   **Survey mode selection** -- some modes generally provide higher
    response rates; sensitive topics may provide better response with
    self-administered modes

-   **Time survey is administered** -- some calling periods or seasons
    of the year may yield higher response rates (e.g., avoid holidays)

-   **Good interviewer training** -- persistence, gentle conversion of
    reluctant persons to respondents, etc.

-   **Careful "physical" design of mail and web surveys** -- "official"
    logos, ensure mail doesn't look like junk mail/SPAM

-   **Carefully designed questionnaire** -- poorly worded questions can
    lead to skipped questions, easy-to-navigate paper survey if mailed

-   **Low respondent burden** -- try to make the survey short and "easy"
    to take, possibly using split-questionnaire design (not all sampled
    units get all questions)

-   **Use of incentives** -- given either before completion
    ("pre-incentive") or upon completion ("post-incentive" or "reward")

-   **Extensive follow-up** -- Might not get response on the first try,
    might consider alternate modes for follow-up (e.g., telephone
    follow-up to mailed survey)

-   **Responsive Design** -- Adjust data collection process while it is
    ongoing, e.g., by reallocating effort to subgroups with higher
    non-contact or non-response
:::
::::

## Two Types of Nonresponse in Surveys

> [**Unit nonresponse**]{.alert} = a sampled unit completely fails to
> respond (to entire survey)

> [**Item nonresponse**]{.alert2} = a sampled unit responds but fails to
> answer to a specific item

-   In general, we:

    -   Deal with [unit nonresponse]{.alert} by making adjustments to
        the base survey weights

    -   Deal with [item nonresponse]{.alert2} with imputation (or, by
        ignoring it...)

-   For large-scale surveys, **unit nonresponse** is arguably the more
    severe problem\
    (declining response rates...)

-   Before we talk about how to handle the missing data, we need to talk
    about *different types of missing data*

## Taxonomy of Missing Data

-   We classify missing data based on the [**underlying reason the data
    are missing**]{.alert2}

-   In general, not testable -- just an assumption

-   Three types of missing data:

    -   **MCAR:** Missing Completely At Random 😀

    -   **MAR:** Missing At Random 😐

    -   **MNAR:** Missing Not At Random 😱

-   Defined by what factors are associated with the probability of missingness
    (or conversely, with the probability of response)

## Notation

-   $y_i$ = survey outcome of interest for unit $i$, [only observed for
    **respondents**]{.alert}

-   $\mathbf{x}_i$ = vector of information known about unit $i$,
    [observed for **ALL sampled units**]{.alert2}

    -   E.g., design information (stratum, cluster membership), or data
        available on the sampling frame

-   $R_i$ = response indictor for unit $i$ (=1 for respondents, =0 for
    nonrespondents)

-   $\phi_i= P(R_i=1)$ = probability unit $i$ responds = response
    propensity (unknown)

And we also have:

-   $Z_i$ = selection indicator (=1 for sampled units, =0 for
    non-sampled units)

-   $\pi_i = P(Z_i=1)$ = selection probability (known!)

## MCAR: Missing Completely At Random {.pause}

> [**MCAR**]{.alert2} = Best-case scenario:  
> $P(R_i=1 | y_i, \mathbf{x}_i) = P(R_i=1)$\
> Probability a variable has missing values does not depend on any
> factors/variables (including itself)

-   Examples:

    -   Coin flip to determine if respond
    -   Computer glitch that wipes out 10% of records

-   [No systematic difference between respondents and
    nonrespondents]{.alert}

    -   Respondents are indistinguishable from nonrespondents\]

    -   $\bar{y}_{RU} = \bar{y}_{NRU}$   →  
        $\text{Bias}(\bar{y}_R) =  \frac{N_{NR}}{N} (\bar{y}_{RU} - \bar{y}_{NRU}) = \frac{N_{NR}}{N} \times 0 = 0$

-   A complete case analysis implicitly makes this assumption!

. . . 

-   Estimates of **survey totals** will be **biased** (literally not
    counting up enough people!)

-   Estimates of quantities like **means** or **proportions** will be
    **unbiased**

-   **Standard errors** will be **larger than if no nonresponse**

## MAR: Missing At Random {.pause}

> [**MAR**]{.alert2} = Next-best-case scenario:  
> $P(R_i=1 | y_i, \mathbf{x}_i) = P(R_i=1|\mathbf{x}_i)$\
> Probability a variable has missing values depends on observed data

-   Examples:

    -   Older adults more likely to respond than younger adults -- and
        we observe all sampled units' ages (i.e., it's on the sampling
        frame)

    -   Unmarried moms less likely to respond than married moms -- and
        we have marital status on the sampling frame

    -   People in one stratum are less likely to respond than in another
        stratum

-   [Nonrespondents are diﬀerent from respondents]{.alert}, but we [can
    see how they are diﬀerent]{.alert}

    -   Since we observe how the respondents and nonrespondents diﬀer,
        we can adjust for it

. . . 

-   Estimates for a $y$ that is related to the variable(s) that diﬀer
    between respondents and nonrespondents will be **biased**

-   But, we can adjust survey weights to "make-up" the diﬀerence and
    remove bias (coming next lecture!)

## Example: MAR

-   SRS of women who recently gave birth in Ohio

    -   **Outcome of interest**: Postpartum feelings of social isolation

    -   **Nonresponse Problem**: All married moms respond, but only 50%
        of unmarried moms respond

-   If married women are **less likely** to experience social isolation
    than unmarried women, our estimate of the prevalence of social
    isolation will be \_\_\_\_\_\_\_\_\_\_ (too low? too high?)

-   Suppose you sampled 100 married moms and 100 non-married moms.

::::::: columns
:::: {.column width="50%"}
::: smaller-text
In truth, 25% of the married moms feel isolated, and 50% of the
non-married moms feel isolated:

|       TRUTH | Isolated | Not Isolated | Total |
|------------:|:--------:|:------------:|:-----:|
|     Married |    25    |      75      |  100  |
| Not Married |    50    |      50      |  100  |
:::
::::

:::: {.column width="50%"}
::: smaller-text
But only half the non-married moms respond (assume nonresponse unrelated
to isolation status):

| WITH NONRESPONSE | Isolated | Not Isolated | Total |
|-----------------:|:--------:|:------------:|:-----:|
|          Married |          |              |       |
|      Not Married |          |              |       |
:::
::::
:::::::

-   What is the prevalence of isolation if the full sample had
    responded?
-   With nonresponse?

## Example: MAR (con't)

:::::: {columns}
:::: {.column width="50%"}
::: smaller-text
| WITH NONRESPONSE | Isolated | Not Isolated | Total |
|-----------------:|:--------:|:------------:|:-----:|
|          Married |    25    |      75      |  100  |
|      Not Married |    25    |      25      |  50   |
:::
::::

::: {.column width="50%"}
-   Observed prevalence of isolation = 0.333 = 33.3%
-   True prevalence = 0.375 = 37.5%
:::
::::::

<p>

-   To "adjust" for this MAR mechanism, we can **upweight** the
    non-married moms

-   Essentially, we want to make the responding sample look like the
    full sample with respect to marital status

    -   Only half the unmarried moms responded (50 out of 100)
    -   Make each responding unmarried mom "represent" 2 unmarried moms
    -   Thus we double the weights of the unmarried moms to "adjust" for
        the nonresponse
    -   Estimated prevalence is:

<br>

<br>

*Note: this assumes that marital status is on the sampling frame – so
that we know it for all sampled women – so we know whose weights to
double!*

## MNAR: Missing Not At Random (sometimes NMAR) {.pause}

> [**MNAR**]{.alert2} = Worst-case scenario:  
> $P(R_i=1 | y_i, \mathbf{x}_i) = P(R_i=1|y_i, \mathbf{x}_i)$\
> Probability a variable has missing values depends on values of
> unobserved data

-   Examples:

    -   Depressed moms less likely to respond than non-depressed moms,
        but we don't know depression status unless they respond

    -   Higher income earners less likely to report their income than
        low wage earners, but this information isn't on the sampling
        frame

-   [Nonrespondents are diﬀerent from respondents]{.alert}, and we [do
    not observe the characteristic(s) that make them diﬀerent]{.alert}

. . . 

-   Can never know if we have MNAR (not possible to test for it)

-   Inference under MNAR generally requires us to posit a model for the
    response mechanism where response depends on unobserved variables
    (like a survey outcome, $y$), which we can never confirm is correct

-   Usually we assume MAR, and (sometimes) assess how wrong our
    conclusions would be if the truth were MNAR -- called a
    **sensitivity analysis**

## Activity 17.1 (Part 2)

The Impact of Missing Data (Part 2)
